### 2022-06-21 docs/rmd/paper_word.Rmd
first commit

---
title: |
  Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression:
    A comparison of $\epsilon$-Lexicase Selection and Tournament Selection
date: "Date of Submission: `r Sys.Date()`"
output:
  word_document:
    reference_docx: "templates/template_ls.docx"
    fig_caption: yes
    number_sections: true
always_allow_html: true 
fontsize: 12pt
toc: yes
toc_depth: 4
bibliography: bib/bib/ref.bib
csl: bib/csl/harvard1.csl
include-before: '`\newpage{}`{=latex}'
---

```{r setup, include=FALSE}

library(kableExtra)
library(magrittr)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = "!H")

```

\newpage

# Abstract

...

# Introduction

```{=html}
<!-- 
Structure:
  - Genetic Programming
  - Selection
  - Symbolic Regression
  - Generalization
  - Research Question
-->
```

<!-- GP --->
Genetic programming (GP), a subfield of evolutionary computation (EC), is a metaheuristic that is used to search for computer programs that solve a given problem by simulating the process of darwinian evolution. The basic principle of GP is to gradually evolve solutions by repeatedly selecting parent solutions from a randomized population of computer programs based on a fitness metric. Then, genetic operators are applied on the selected solutions to genereate new offspring candidate solutions. By repeating this process over many generations, GP acts as a guided search for high fitness solutions throughout the decision space.
A unique feature of GP among other evolutionary optimization procedures is the possibility to evolve solutions of variable length.

<!-- Selection --->
The overall performance of GP can depend strongly on the choice of its underlying operators, one crucial component in this is the operator for parent selection. 
Tournament Selection is a commonly used selection operator in EC and is the most used operator in GP systems [@10.1007/978-3-642-16493-4_19, p.181]. A parent solution is selected by randomly sampling $k$ individuals from the current population into a tournament pool and then the solution with the highest fitness score from the tournament pool is selected [@10.1007/978-3-642-16493-4_19, p.182]. 
Lexicase selection has been suggested as an alternative to tournament selection that is not based on aggregating fitness scores. It samples $n$ test cases in random order and then eliminates solutions from the selection pool on a per test case basis if they are not performing on an elite level [@6920034, p.1].

Since regular Lexicase selection has been shown to perform suboptimal on continous-valued optimization problems, a modified variation called $\epsilon$-Lexicase selection has been suggested for symbolic regression by La Cava et. al (2016). Here, $\epsilon$-Lexicase selection has shown itself to outperform both in overall performance while showing only negligible computational overhead [@epsilon_lexicase_main, p.747].

<!-- Symbolic Regression --->
The goal of symbolic regression is to find a mathematical model for an observed set of datapoints [@10.1007/978-3-540-24621-3_22, p.794]. Symbolic Regression has been one of the first GP applications and to this day is an actively studied and highly relevant area of research [@poli08:fieldguide, p.114]. In most symbolic regression problems, little to no a priori knowledge about the optimal form and structure of the target function is available. The ability of GP to optimize for model structure as well as for parameters has lead to it being one of the most prevalent methods used in the domain of symbolic regression [@10.1007/978-3-540-24621-3_22, p.795].

<!-- Generalization --->
An important quality of all supervised machine learning applications, including GP, is the ability to not only optimize performance for the test cases a model is trained on but to also perform well on previously unseen cases, this is refered to as generalization. In most real world applications of symbolic regression only a small subset of labeled data is available for training. The aim is to produce a model that not only accurately predicts the provided training data but can also predict previously unseen cases with high precision [@Gonalves2016AnEO, p.6]. A model that is extensivley optimized on the provided training data, may overfit to this data sample which may lead to a decrease in generalization.

<!-- Research Question --->
This research project tries to answer the question if the usage of $\epsilon$-Lexicase selection influences the generalization behaviour of programs that are evolved using GP for symbolic regression if compared to programs that are evolved using traditional tournament selection.

# An Overview of the current state of research

## Selection

In the first description of lexicase^[The alternative, more descriptive name given by @lexicase_first_desciption: global pool, uniform random sequence, elitist lexicase parent selection] selection for GP by @lexicase_first_desciption it was suggested as a novel parent selection method for target problems that are modal in nature. The author classified modal problems as "problems that qualitatively different modes of response are required for inputs from different regions of the problem's domain"[@lexicase_first_desciption, p.1].

In a following article Lexicase selection has been proposed specifically for the purpose of solving so called uncompromising problems with GP. @6920034 defined uncompromising problems as problems that require the final solution to perform optimal on each of the cases it is tested on, examples include symbolic regression, the design of digital multipliers or finding terms in finite algebras. The authors also provided evidence that lexicase selection can significantly improves GP's ability to solve some uncompromising problems if compared to selection methods that are based on aggregated fitness [@6920034, p.12].

Both articles argued that many of the problem domains that GP is commonly used for are prevalent for problems that are modal/ uncompromising in nature. In comparison to other selection operators, populations that are evolved using variations of the lexicase selection operator show a very high degree of genetic diversity which might be a key contributer to the improved performance [@6920034, p.1] [@epsilon_lexicase_main, p.745]. In theory, Lexicase based selection should be more prone to select solutions that are specialist (high fitness on a small subset of training cases) than solutions that are generalist (high average fitness among all training cases).

The basic algorithm for lexicase parent selection as described by @lexicase_first_desciption is given below:

```
Lexicase - Parent-Selection:

1. Initialize:
  (a) Set candidates to be the entire population.
  (b) Set cases to be a list of all of the fitness cases in random order.

2. Loop:
  (a) Set candidates to be the subset of the current candidates that have 
      exactly the best fitness of any individual currently in candidates for 
      the first case in cases.
  (b) If candidates or cases contains just a single element then return
      the first individual in candidates.
  (c) Otherwise remove the first case from cases and go to Loop.
```

More recent research by @epsilon_lexicase_main suggested $\epsilon$-Lexicase selection as a modified selection operator that can improve overall GP performance if applied to continous-valued symbolic regression tasks in comparison to tournament and standard lexicase selection.  @epsilon_lexicase_main argued that the original concept behind lexicase selection as formulated by @lexicase_first_desciption does not fit the requirements of real-world symbolic regression tasks. The authors identified the pass condition of regular lexicase selection as the main problem if applied to symbolic regression: Individual solutions can only be selected if they perform on an elite level but with noisey and continous-valued data it is very unlikely for two individuals to achieve an exactly equal error on a test case, this finally results in filtering out too many individuals during the selection process[@epsilon_lexicase_main, p.742]. $\epsilon$-Lexicase adresses this problem by introducing an $\epsilon$ parameter that specifies a range around the elite error, individuals that perform inside this range pass the current selection iteration. @https://doi.org/10.48550/arxiv.1709.05394 describe $\epsilon$-Lexicase selection as a more "relaxed version of lexicase selection".
Different methods to configure the $\epsilon$ Parameter have been explored, for the limited scope of this project I will focus on the most promising implementation of $\epsilon$-Lexicase selection that is automatically adjusted on the basis of the median absolute deviation of errors inside the selection pool [@epsilon_lexicase_main, p.742] [@https://doi.org/10.48550/arxiv.1709.05394, p.6].

The performance increase of $\epsilon$-Lexicase selection for symbolic regression problems has been demonstrated and reported for many benchmark problems which led to widespread adaption of it in symbolic regression applications [@epsilon_lexicase_main, p.744-745].

In comparison, traditional GP selection methods such as tournament selection most commonly compute the fitness of a program as the mean of its error for each individual fitness case. One downside to this approach is that the total amount of information is reduced from a wide range of individual errors to a single metric. @6920034 suspected that this loss in overall information provided to GP might reduce overall performance especially if applied to the class of uncompromising problems that require the solution to perform well on a wide range of diverse cases. Since tournament selection selects individuals based on their average fitness across all test cases, the resulting solutions should be expected to be more biased towards being generalists instead of specialists.

The basic algorithm for tournament is given below for comparison [@10.1007/978-3-642-16493-4_19, p.182-183]:

```
Tournament - Parent-Selection:

k = tournament size

1. Sample:
  (a) Randomly sample k individuals from the current population
      into a tournament pool.
2. Select:
  (a) Compute the mean fitness for each individual inside the tournament pool
      based on all fitness cases.
  (b) Select individual from the tournament pool with the highest mean fitness.
```

## Generalization

Generalization, the ability of a model to perform well on previously unseen cases, is one of the fundamental goals in most real world machine learning applications. @open_issues_gp raised awareness to the fact, that the topic of generalization in GP has not gotten the attention that other machine learning-based areas, e.g, deep learning, have attributed to it. A similiar statement was proposed by @generalisation_in_gp in his review of research in generalization in GP. Both authors adressed the need for additional research regarding the topic.

@generalisation_in_gp critized that almost all applications published in the initial GP literature by John Koza [@koza_main] are not using separate training and testing datasets which might result in overfitting and a poor overall generalization ability of the programs that are produced. The author calls for a more widespread adoption of methods like generational sampling of new training cases or the overall separation into training and testing datasets to improve generalization in GP [@generalisation_in_gp, p.10].

An aspect that is suspected to negatively correlate with GP's generalization is overall size and in particular bloating of the resulting programs. Bloating can be described as a growth in the total size of a program that does not also improve it's performance in any meaningful way [@bloat_overfitting_gp, p. 1].
It has been widely suspected that GP configurations that produce very large programs also have a higher tendency to specialize on difficult or unusual test cases which could result in lower generalization [@wang_wagner_rondinelli_2019, p. 268]. The minimum description length principle (MDLP) describes this phenomenom by claiming that less complex models are more likely to perform better at generalizing than more complex models that achieve a comparable fitness during the training phase [@open_issues_gp, p. 349]. However MDLP should be considered carefully, experiments by @bloat_overfitting_gp on the prediction of bioavailability for drugs demonstrated that GP based techniques can either bloat without overfitting or overfit without bloating [@bloat_overfitting_gp, p. 8].

## Symbolic Regression

The task of finding a mathematical function that fits a given set of datapoints has been one of the first applications of GP in [@koza_main, p. 238].
Practical examples given by @koza_main included tasks such as the discovery of scientific laws, finding solutions to differential and integral equations or the programmatic compression of images.

Since it's inception, GP based symbolic regression has been widely used in practical application ranging over a wide field of disciplines. A well known example of an industrial use-case has been published by @soft_sensor_gp. The authors used GP based symbolic regression to derive nonlinear functions that could be used to significantly increase the robustness of industrial sensors which resulted in wide-spread adoption and a significant reduction in costs.

Many variations of GP based symbolic regression have been proposed to improve overall performance, @wang_wagner_rondinelli_2019 summarized the main characteristics and differences for 6 common variations including procedures such as geometric semantic genetic programming, cartesian genetic programming or GP-based relevance vector machines. A large benchmarking study on symbolic regression published by @Orzechowski_2018 compared 4 GP-based procedures to several state-of-the-art machine leaning techniques. The authors find that GP, eventhough more time consuming, can achieve better results if compared to the state-of-the-art machine learning algorithm gradient-boosting.

# Experimental study

## Research Design

<!-- Dataset and Task --->
To study the influence of selection on generalization I selected a dataset about energy efficiency in buildings that is part of the UC Irvine Machine Learning Repository [@Dua:2019]. The dataset contains eight individual building attributes that map to two different outcomes, heating and cooling load, for N=768 cases [@fe8fa39e88a040bbacba5a465c48043f].

Using two otherwise identical GP systems, one deploying tournament selection and the other $\epsilon$-lexicase selection, the objective is to find a computer program that best predicts the outcome variable heating load $(Y1)$^[The symbolic regression is performed on one of the two provided outcome variables, the variable Cooling Load will be excluded.] of the buildings using a subset of the eight building attributes $(X1,..,X8)$ for input. The specific meaning of all attributes inside the dataset are described in table 1.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
fig.pos="H"

knitr::kable(
  read.csv("./tables/csv/variables_energy_heating.csv"),
  digits=3,
  caption="Overview - Energy Heating Dataset"
)    %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
```

<!-- Method:single run --->
To measure the generalization ability of each model the dataset will be randomly split in half, resulting in a training and testing dataset each containing 384 individual cases. Each model will be evolved by traditional GP using only the fitness cases present in the training dataset. For each generation the highest fitness model of the current population will be tested with the previously unseen fitness cases that are part of the testing dataset. For each run of the experiment statistics will be collected on fitness which will form the basis of my further statistical analysis. Additional statistics will be collected for the average length of the population and the length of the elite program for each generation to explore differences in their distribution and possible correlations to generalization.

<!-- Statistical Analysis --->
Since GP is a stochastic optimization algorithm, the basic experiment will be run for a total of 50 times to ensure a fair and meaningful comparison based on a large number of runs for both algorithms. For each run of the experiment the dataset will be randomly split in half as described, both models are then trained and tested using the exact same set of fitness cases.

The statistical analysis of the collected data will first focus on examining the question if, on average, the usage of $\epsilon$-Lexicase selection will result in models that perform significantly different than models that are evolved using tournament selection both.

> $H0_{1}$: The distribution underlying the samples of training/testing errors produced by torunament selection is the same as the distribution underlying samples of training/testing errors produced by $\epsilon$-lexicase selection

The next question of interest is, to examine if statistical significant differences between the mean errors of training and testing data exist for both selection operators:

> $H0_{2}$: The distribution underlying the samples of training errors produced by $\epsilon$-Lexicase/torunament selection is the same as the distribution underlying samples of testing errors produced by $\epsilon$-Lexicase/tournament selection

To gather further insight into the differences between both GP systems, an additional test will be performed on the question if differences in the total size of the resulting programs exist:

> $H0_{3}$: Size differences exist between the distribution underlying the samples produced by $\epsilon$-Lexicase and the distribution underlying samples of tournament selection

All hypothesis will be tested using a level of significance of $\alpha=0.05$.

To further examine the difference in generalization behaviour, the mean testing and training errors over each generation will be visualized for both algorithms. The specific aim of this visualization is to explore the mechanism of overfitting and to examine if differences between both methods can be detected.

All GP experiments will be implemented by using the python programming language in conjunction with DEAP, a framework for distributed evolutionary algorithms that implements various tools and algorithms for genetic programming [@DEAP_JMLR2012].

## Gentetic Programming Configuration

### Evolutionary parameters

The basic evolutionary parameters for both systems are presented in table 2. Tournament selection is used with a default tournament size of $3$ individuals while the $\epsilon$ parameter in lexicase selection is selected automatically as previously mentioned in subsection 3.1.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
fig.pos="H"

knitr::kable(
  read.csv("./tables/csv/gp_parameters.csv"),
  digits=3,
  caption="Evolutionary Parameters"
) %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
```

### Fitness Evaluation

The fitness $f$ for each model will be based on the mean squared error (MSE) over all fitness cases for prediction and the empirically measured values as described by eq.1.

\begin{equation}
\tag{eq. 1}
MSE = \frac{1}{n} * \sum_{i=1}^{n} (Y_i - \hat{Y_1})^2
\end{equation}

where:

-   $n$: total number of test cases
-   $Y_i$: empirical value for case $i$
-   $\hat{Y_1}$: predicted value for case $i$.

The resulting fitness function $f$ for an individual program $i$ is descibed by eq 2:

\begin{equation}
\tag{eq. 2}
 f(i, \tau ) = \frac{1}{N} * \sum_{t \epsilon \tau} (y_t - y\hat{}_t(i, x_t))^2 
\end{equation}

where^[Naming and notation was adopted from @epsilon_lexicase_main]:

-   $\tau$: the set of $N$ fitness cases
-   $y_t$: empirical value of the target for case $t$
-   $y\hat{}_t(i, x_t)$: predicted value for the target for case $t$ by running the program $i$ with the total set of input variables $x_t$

### Primitive Set

The primitve set consists of the terminals that are listed in table 3 and the functions that are listed in table 4. To avoid runtime errors I implemented protected version of the operators for division, natural logarithm and square root [@koza_main, p.82-83]. As suggested by @koza_main I also included ephemeral constants to the set of terminals to provide the evolutionay search the opportunity to explore and include randomly generated constants.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
fig.pos="H"

knitr::kable(
  read.csv("./tables/csv/terminals.csv"),
  digits=3,
  caption="Terminals"
) %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
fig.pos="H"

knitr::kable(
  read.csv("./tables/csv/primitives.csv"),
  digits=3,
  caption="Functions"
) %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
```

### Genetic Operators

GP operators are represented in table 5. Both genetic operators use a static limit to control for the height of the reulting trees [@koza_main, p.104]. Individual programs are initialized by using the ramped half-and-half method, 50% of the population are created by using the Growth algorithm and the remaining 50% are created by using the Full algorithm [@koza_main, p.93].

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
fig.pos="H"

knitr::kable(
  read.csv("./tables/csv/genetic_operators.csv"),
  digits=3,
  caption="GP Operators"
) %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
```

The crossover operator implemented by DEAP randomly selects a crossover point in each individual and exchanges each subtree with the point as root between each individual [@DEAP_JMLR2012]. Mutation also randomly selects a point in the tree individual, it then replaces the subtree below that point as a root by the expression generated using the full grow initialization method [@DEAP_JMLR2012].

# Results

## Descriptive Statistics

Tables 6 and 7 summarize the results for all fitness scores collected over 50 total runs of the experiment^[All floating point numbers in text have been rounded to 3 decimal places].

```{r, echo=FALSE, warning=FALSE}

fig.pos="H"

df <- read.csv("./tables/csv/master_tournament_descriptive.csv")

knitr::kable(
  df,
  format = "latex",
  booktabs = TRUE,
  digits=3,
  caption="Summary - Tournament Selection"
) %>% kable_styling(latex_options = "scale_down") %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")

```

```{r, echo=FALSE, warning=FALSE}

fig.pos="H"

df <- read.csv("./tables/csv/master_elexicase_descriptive.csv")

knitr::kable(
  df,
  format = "latex",
  booktabs = TRUE,
  digits=3,
  caption="Summary - Epsilon-Lexicase Selection"
)  %>% kable_styling(latex_options = "scale_down") %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")

```

<!-- Statistical Analysis

For both experiments the best of the run models that are evolved using $\epsilon$-Lexicase Selection on average perform better as measured by their mean error score. This behaviour is consistent during both training and testing phases of the experiment

$\epsilon$-Lexicase based GP achieves an average minimum fitness of 3.166/1.576 (MSE/MAE) during model training and a score of 2.949/1.678 during the testing phase on previously unseen data. Suprisingly the models that are trained using MSE as a metric for fitness evaluation, on average, perform even better on unseen cases as they do on the initial training data.

In comparison, the models that are trained using tournament selection perform with an average minimum fitness of 10.9887/2.32173 (MSE/MAE) during model training and 11.146/2.196 during the testing phase. Again we observe an improved performance in testing error, this time models that are evolved using MAE for fitness perform better on testing data than on their initial training data.

The initial results seem support the two alternative hypothesis of this research project:

- $HA_{testing}$: T $\epsilon$-Lexicase Selection produces models that differ in their generalization behaviour if compared to models that are evolved using torunament selection

- $H0_{training}$: $\epsilon$-Lexicase Selection produces models that differ in their performance during model training if compared to models that are evolved using torunament selection

--->

## Inferential Statistics

Figure 1 visualizes the distribution of fitness scores achieved by both GP systems after finishing 100 full generations for all 50 runs of the experiment.

![Distribution of Errors](./plots/mean_error_boxplot_all.png)

... interpretation...

The four samples from figure 1 have been tested for normal distribution by computing the D’Agostino and Pearson test [@DAgostino1971AnOT] [@DAgostino1973TestsFD] to determine if they satisfy the requirements of the students t-test. The results are summarized in table 8. Since not all samples test positive for a normal distribution a Mann-Whitney U (MWU) ranksum test will be performed to test for statistical significance.

```{r, echo=FALSE, warning=FALSE}

fig.pos="H"

df <- read.csv("./tables/csv/normal_dist_test.csv")

knitr::kable(
  df,
  format = "latex",
  booktabs = TRUE,
  digits=3,
  caption="Normal Distribution Tests"
) %>% kable_styling(latex_options = "scale_down") %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")

```

The results of the MWU test are summarized in table 9.

```{r, echo=FALSE, warning=FALSE}

fig.pos="H"

df <- read.csv("./tables/csv/mwu_matrix_size.csv")

knitr::kable(
  df,
  format = "latex",
  booktabs = TRUE,
  digits=3,
  caption="Mean Error - P-Values"
) %>% kable_styling(latex_options = "scale_down") %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")

```

... interpretation...

To further examine the generalization behaviour based on both selection operators I plotted the mean fitness scores for both algorithms in figure 2.

![Mean Errors](./plots/mean_error_combined.png)

... interpretation...

In a final step I analyzed the growth behaviour for both GP systems. Figure 3 visualizes the average size of individuals inside the population as well as the average size of the best performing individual for each generation of the evolution.

![Mean Size](./plots/size_combined.png)
... interpretation...

Again, a MWU test is conducted to test for statistical differences in the underlying distribution of the sizes measured during the experiment. The results are summarized in table 10.

```{r, echo=FALSE, warning=FALSE}
fig.pos="H"

df <- read.csv("./tables/csv/mwu_matrix_size.csv")

knitr::kable(
  df,
  format = "latex",
  booktabs = TRUE,
  digits=3,
  caption="Mean Size - P-Values"
) %>% kable_styling(latex_options = "scale_down") %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")

```

# Conclusion

# Limitations and open questions

I faced two major difficulties in the preparation and evaluation of this research project:

  1. Configuration of evolutionary parameters
  2. Limited computational ressources

The combination of both issues might have resulted in a significant reduction in overall robustness of the results represented in sections 5 and 6. The performance of all evolutionary algorithms, including GP, can be highly dependant on a large number of parameters and implementation details. Although the series of experiments conducted in this project are comparably simple in nature, the results achieved might still be highly influenced by the GP configuration detailed in subsection 4.2.

To gain further insight into the differences in generalizaion behaviour and to adress the research question with a higher level of certainty, the experiment I conducted should be repeated for many different GP configurations. Some exemplary parameters that could drastically influence the results of this experiment include the population size, the total number of generations, the inclusion of elitism or the computation of $\epsilon$ in lexicase selection. Other factors whose influence on generalization behaviour could be important include different methods to compute an individuals fitness (e.g. mean absolute error or root mean squared error) and testing out different variations of the genetic operators for mutation and crossover.

Another obvious weakness of my project is that the experiment is only based on a single symbolic regression application, the prediction of the heating load of buildings based on a relativley small dataset. To gain further trust in the obtained results, the experiment should be repeated for other symbolic regression tasks and datasets. At the current state it remains unknown if the generalization behaviour of both selection operators might be highly problem specific and dependant on the dataset used for training and testing.

All limitations and shortcomings described above are also connected to the second limitation of this research project, the limited amount of computational ressources. The total computation time consumed to run the current experiments (50 runs, 2 algorithms, 100 generations, 500 individuals) has already been close to 24 hours on a modern, arm-based Apple-M1 system. Eventhough the overall computation time could probably be reduced by further optimization of the source code, e.g. using tools for parallel computation, the ressources necessary to repeat the whole experiment for different parameter configurations would certainly be beyond the scope of this research project.

\newpage

# References

### 2022-06-21 src/__old__/__oldold__/statistical_evaluation.py

@@ -1,0 +1,19 @@
+from scipy.stats import mannwhitneyu
+import pandas as pd
+
+
+def main():
+
+    # load results_50gens_500pop into pd.DataFrame
+    data_path = "../results_50gens_500pop/multiple_runs.csv"
+    df = pd.read_csv(
+        filepath_or_buffer=data_path,
+        sep=",",
+        header=0,
+    )
+
+    print(df.head())
+
+
+if __name__ == "__main__":
+    main()

### 2022-06-21 src/__old__/gp_symreg.py

@@ -1,0 +1,750 @@
+import math
+import operator
+import os
+from dataclasses import dataclass
+from random import random, randint
+from statistics import pstdev
+from sys import stderr
+from typing import Any, Tuple, Callable, List
+
+import networkx as nx
+import numpy as np
+import pandas as pd
+from deap import gp, tools, creator, base, algorithms
+from matplotlib import pyplot as plt
+from sklearn.model_selection import train_test_split as skl_train_test_split
+
+
+def get_datasets(training_split: float = 0.5) -> Tuple[pd.DataFrame, pd.DataFrame]:
+    """
+    Download the original database, splits and returns two randomly sampled datasets for training and testing
+
+    Usage:
+        0.0 < <training_split> < 1.0
+        size training dataset: N *       <training_split>
+        size testing dataset:  N *  (1 - <training_split>)
+    """
+
+    assert 0.0 < training_split < 1.0
+
+    FILENAME = "ENB2012_data.xlsx"
+    URL = "https://archive.ics.uci.edu/ml/machine-learning-databases/00242/"
+    DESTINATION = "../data"
+
+    def __download() -> None:
+        if not os.path.exists(f"{DESTINATION}/{FILENAME}"):
+            os.system(f"wget {URL}{FILENAME} -P {DESTINATION}")
+
+    def __split(
+            __training_split: float
+    ) -> Tuple[pd.DataFrame, pd.DataFrame]:
+        df = pd.read_excel(f"{DESTINATION}/{FILENAME}")
+        return skl_train_test_split(
+            df,
+            train_size=__training_split,
+            test_size=(1.0 - __training_split),
+        )
+
+    __download()
+    return __split(training_split)
+
+
+def plot_expression_tree(expr_tree, title: str, filepath: str) -> None:
+    """plots an expression tree and saves it as ../plots/<filename>.png"""
+    nodes, edges, labels = gp.graph(expr_tree)
+
+    g = nx.Graph()
+    g.add_nodes_from(nodes)
+    g.add_edges_from(edges)
+
+    pos = nx.drawing.nx_agraph.graphviz_layout(g, prog="dot")
+
+    nx.draw_networkx_nodes(g, pos)
+    nx.draw_networkx_edges(g, pos)
+    nx.draw_networkx_labels(g, pos, labels)
+
+    plt.title(title)
+    plt.savefig(filepath)
+
+
+def get_primitive_set() -> gp.PrimitiveSet:
+    """returns a basic deap.gp.PrimitiveSet object for symbolic regression without registering methods for selection
[…]

### 2022-06-21 src/__old__/main.py

@@ -1,0 +1,105 @@
+from copy import deepcopy
+from sys import argv
+
+from seminar.src.old_version.gp_symreg import *
+
+"""y
+Description:
+    Do a full run (Train and Test) for both algorithms 
+
+Usage:
+    ARGV[1]: Target Variable (y1 | y2)
+    ARGV[2]: Error Metric (absolute | squared) 
+"""
+
+
+def run(target_var: str, err_metric: str):
+    """Train, Test and log results_50gens_500pop for <target_var> using mean <err_metric> error for minimizing fitness function"""
+    # get database and split into two pd.DataFrames
+    TRAINING_DATASET, TESTING_DATASET = get_datasets(training_split=0.5)
+
+    # primitive set
+    PRIMITIVE_SET = get_primitive_set()
+
+    # deap toolbox
+    TOOLBOX = get_toolbox(
+        primitive_set=PRIMITIVE_SET,
+    )
+
+    # deap multi statistics
+    MULTI_STATISTICS = get_multi_statistics()
+
+    # gp configuration
+    EA_CONFIG = EaConfig(
+        population_size=50, #500
+        num_generations=10, #50
+        mutation_rate=0.05,
+        crossover_rate=0.2,
+        multi_statistics=MULTI_STATISTICS,
+        target_var=target_var,
+        err_metric=err_metric,
+    )
+
+    # train model for y1 using tournament selection
+    training_results_tournament = train_tournament(
+        toolbox=deepcopy(TOOLBOX),
+        ea_config=EA_CONFIG,
+        training_dataset=TRAINING_DATASET,
+        testing_dataset=TESTING_DATASET,
+        pset=deepcopy(PRIMITIVE_SET)
+    )
+
+    # # test model on unknown dataset
+    # testing_results_tournament = test_model(
+    #     training_results_tournament,
+    #     EA_CONFIG,
+    #     TESTING_DATASET,
+    #     PRIMITIVE_SET,
+    # )
+
+    # # log single run results_50gens_500pop
+    # log_single_run(
+    #     ea_results=training_results_tournament,
+    #     testing_results=testing_results_tournament,
+    #     selection_method="tournament"
+    # )
+
+    # train model for y1 using epsilon lexicase selection
+    training_results_epsilon_lexicase = train_epsilon_lexicase(
+        toolbox=deepcopy(TOOLBOX),
+        ea_config=EA_CONFIG,
+        training_dataset=TRAINING_DATASET,
[…]

### 2022-06-21 src/__old__/single_run.py

@@ -1,0 +1,120 @@
+from copy import deepcopy
+from sys import argv
+
+from seminar.src.old_version.gp_symreg import *
+
+
+def run(target_var: str, err_metric: str):
+    """Train, Test and log results_50gens_500pop for <target_var> using mean <err_metric> error for minimizing fitness function"""
+    # get database and split into two pd.DataFrames
+    TRAINING_DATASET, TESTING_DATASET = get_datasets(training_split=0.65)
+
+    # primitive set
+    PRIMITIVE_SET = get_primitive_set()
+
+    # deap toolbox
+    TOOLBOX = get_toolbox(
+        primitive_set=PRIMITIVE_SET,
+    )
+
+    # deap multi statistics
+    MULTI_STATISTICS = get_multi_statistics()
+
+    # gp configuration
+    EA_CONFIG = EaConfig(
+        population_size=500,
+        num_generations=40,
+        mutation_rate=0.05,
+        crossover_rate=0.2,
+        multi_statistics=MULTI_STATISTICS,
+        target_var=target_var,
+        err_metric=err_metric,
+    )
+
+    # train model for y1 using tournament selection
+    training_results_tournament = train_tournament(
+        toolbox=deepcopy(TOOLBOX),
+        ea_config=EA_CONFIG,
+        training_dataset=TRAINING_DATASET,
+    )
+
+    # test model on unknown dataset
+    testing_results_tournament = test_model(
+        training_results_tournament,
+        EA_CONFIG,
+        TESTING_DATASET,
+        PRIMITIVE_SET,
+    )
+
+    # train model for y1 using epsilon lexicase selection
+    training_results_epsilon_lexicase = train_epsilon_lexicase(
+        toolbox=deepcopy(TOOLBOX),
+        ea_config=EA_CONFIG,
+        training_dataset=TRAINING_DATASET,
+    )
+
+    # test model on unknown dataset
+    testing_results_epsilon_lexicase = test_model(
+        training_results_epsilon_lexicase,
+        EA_CONFIG,
+        TESTING_DATASET,
+        PRIMITIVE_SET,
+    )
+
+    full_results = FullResults(
+        algorithm_1="Tournament Selection",
+        training_results_1=training_results_tournament,
+        testing_results_1=testing_results_tournament,
+        algorithm_2="Epsilon-Lexicase Selection",
+        training_results_2=training_results_epsilon_lexicase,
+        testing_results_2=testing_results_tournament
+    )
+
+    # log results_50gens_500pop
+    directory = log_single_run(
[…]

### 2022-06-21 src/run_statistics.py
autogit

@@ -1,772 +1,0 @@
-#!/usr/bin/env python
-# coding: utf-8
-
-# # Statistics
-# 
-# This notebook needs to be executed to create all the plots and tables of statistical tests that are referenced in the main paper.Rmd file. Output is created from the csv files located at ../results/single_run
-# 
-# 
-# The initial results can be reproduced by running the shell script <run.zsh>
-# 
-# All results of this notebook are saved in ../docs/rmd/plots and ../docs/rmd/tables
-
-# In[185]:
-
-
-from scipy.stats import mannwhitneyu, normaltest
-import pandas as pd
-import numpy as np
-from math import sqrt
-from matplotlib import pyplot as plt
-from typing import Tuple, List
-from statistics import mean
-from dataclasses import dataclass
-import csv
-import os
-
-get_ipython().run_line_magic('cd', '~/github/geneticProgramming/seminar/src')
-
-
-# In[186]:
-
-
-# plotting setup
-
-get_ipython().run_line_magic('matplotlib', 'inline')
-
-FIGSIZE_INCHES =  14, 9
-FIGSIZE_INCHES_LARGE =  16, 10
-TITLE_FONT_SIZE = 17
-TITLE_FONT_SIZE_LARGE = 20
-
-
-# In[187]:
-
-
-# set output paths
-
-PATH = "../docs/rmd/"
-TABLE_PATH = f"{PATH}/tables"
-PLOT_PATH =  f"{PATH}/plots"
-
-jobs = (
-    'os.makedirs(f"{TABLE_PATH}/csv")',
-    'os.makedirs(f"{TABLE_PATH}/md")',
-    'os.makedirs(f"{PLOT_PATH}")'
-)
-
-for job in jobs:
-    try:
-        exec(job)
-    except FileExistsError:
-        pass
-
-# set input paths
-
-DIR_TOURNAMENT = "../results/single_run/e_lexicase"
-DIR_ELEXICASE = "../results/single_run/tournament"
-
-
-tournament_files = os.listdir(DIR_TOURNAMENT)
-elexicase_files = os.listdir(DIR_ELEXICASE)
-
-if not len(tournament_files) == len(elexicase_files):
-    print("Warning - Unequal number of records!\nVariable <TOTAL_RUNS> not set")
-    TOTAL_RUNS = None
-    
-else:
-    TOTAL_RUNS = len(tournament_files)
[…]

### 2022-06-22 docs/rmd/paper.Rmd
autogit

--- 2022-06-21 docs/rmd/paper_word.Rmd
@@
 title: |
-  Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression:
+  <center>Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression: </center>
+  <center>
+    <small>
     A comparison of $\epsilon$-Lexicase Selection and Tournament Selection
+     </small> 
+  </center>
+  <center>--------------------------------------------------------</center>
+  <center>Student: Roman Höhn</center>
+  <center>Date of Birth: 1991-04-14</center>
+  <center>Place of Birth: Wiesbaden, Hesse</center>
+  <center>Student ID: 2712497</center>
+  <center>Supervisor: David Wittenberg</center>
+  <center>--------------------------------------------------------</center>
+  <center>Research Project: Seminar Information Systems (03.996.3299)</center>
+  <center>Research Area: Computational Intelligence</center>
+  <center>Chair of Business Administration and Computer Science</center>
+  <center>Johannes Gutenberg University Mainz</center>
+  <center>Summerterm 2022</center>
+  |
 date: "Date of Submission: `r Sys.Date()`"
 output:
-  word_document:
-    reference_docx: "templates/template_ls.docx"
+  pdf_document:
+    keep_tex: true
+    extra_dependencies: ["float"]
     fig_caption: yes
     number_sections: true
-always_allow_html: true 
+    includes:
+      in_header: settings.tex
 fontsize: 12pt
-toc: yes
-toc_depth: 4
+#toc: yes
+# toc_depth: 4
 bibliography: bib/bib/ref.bib
 csl: bib/csl/harvard1.csl
-include-before: '`\newpage{}`{=latex}'
+#include-before: '`\newpage{} `{=latex}'
+include-before: '`\thispagestyle{empty} `{=latex}'
 ---
@@
 ```{r setup, include=FALSE}
-
 library(kableExtra)
 library(magrittr)
-
 knitr::opts_chunk$set(echo = TRUE)
 knitr::opts_chunk$set(fig.pos = "!H")
-
 ```
 
+\tableofcontents
+\thispagestyle{empty}
 \newpage
-
-# Abstract
-
-...
 
@@
 
-# References
+# I   References {.unnumbered #I}
+
+::: {#refs}
+:::
+\newpage
+
+# II  Statutory Declaration {.unnumbered #II}

### 2022-06-22 docs/rmd/paper.Rmd
autogit

--- 2022-06-21 docs/rmd/paper_word.Rmd
@@
 title: |
-  Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression:
+  <center>Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression: </center>
+  <center>
     A comparison of $\epsilon$-Lexicase Selection and Tournament Selection
+  </center>
+  <center>--------------------------------------------------------</center>
+  <center>Student: Roman Höhn</center>
+  <center>Date of Birth: 1991-04-14</center>
+  <center>Place of Birth: Wiesbaden, Hesse</center>
+  <center>Student ID: 2712497</center>
+  <center>Supervisor: David Wittenberg</center>
+  <center>--------------------------------------------------------</center>
+  <center>Research Project for Seminar Information Systems (03.996.3299)</center>
+  <center>Chair of Business Administration and Computer Science</center>
+  <center>Johannes Gutenberg University Mainz</center>
+  <center>Summerterm 2022</center>
+  |
 date: "Date of Submission: `r Sys.Date()`"
 output:
-  word_document:
-    reference_docx: "templates/template_ls.docx"
+  pdf_document:
+    keep_tex: true
+    extra_dependencies: ["float"]
     fig_caption: yes
     number_sections: true
-always_allow_html: true 
+    includes:
+      in_header: settings.tex
 fontsize: 12pt
-toc: yes
-toc_depth: 4
+# toc: yes
+# toc_depth: 4
 bibliography: bib/bib/ref.bib
 csl: bib/csl/harvard1.csl
-include-before: '`\newpage{}`{=latex}'
+# include-before: '`\newpage{} `{=latex}'
+include-before: '`\thispagestyle{empty} `{=latex}'
 ---
@@
 
-The basic algorithm for lexicase parent selection as described by @lexicase_first_desciption is given below:
+The first description^[@lexicase_first_desciption, p.4] of the lexicase parent selection algorithm by @lexicase_first_desciption is quoted below for reference:
 
@@
 
-## Descriptive Statistics
-
-Tables 6 and 7 summarize the results for all fitness scores collected over 50 total runs of the experiment^[All floating point numbers in text have been rounded to 3 decimal places].
+Tables 6 and 7 summarize the results for all fitness scores collected over 50 total runs of the experiment^[All floating point numbers in tables have been rounded to 3 decimal places if not stated otherwise].
 
@@
 
-<!-- Statistical Analysis
+Some interesting observation from tables 6 and 7 can be made by comparing the mean of the variables $min\_training\_error$ and $testing\_error$.
+Variable $min\_training\_error$ represents the fitness, measured by the MSE, of the best performing solution that has been found during a single run of the experiment while the variable $testing\_error$ represents the fitness of the same solution if evaluated for the unknown fitness cases inside the testing dataset.
+Tournament selection based GP achieves a mean $min\_training\_error$ of $16.483$ and a mean $testing\_error$ of $16.905$ while the GP system using $\epsilon$-Lexicase selection can only achieve a mean $min\_training\_error$ of $22.013$ and a mean $testing\_error$ of $22.456$.
+These results indicate an overall performance benefit for torunament selection based GP in the experiment.
+It should also be mentioned that the best performing model over all 50 runs, measured by the minimum of variables $min\_training\_error$ and $testing\_error$, has been achieved by tournament selection with an MSE of 7.066 for the training data and	7.847 for the testing data.
 
-For both experiments the best of the run models that are evolved using $\epsilon$-Lexicase Selection on average perform better as measured by their mean error score.
-This behaviour is consistent during both training and testing phases of the experiment
+If we compare the differences of $min\_training\_error$ and $testing\_error$ on the basis of each algorithm, we can observe that both tournament and $\epsilon$-lexicase achieve a slightly lower error on the training data than on the testing data.
+If we compute the total difference between the means of $min\_training\_error$ and $testing\_error$ for both algorithms tournament selection amounts to 0.422 and $\epsilon$-Lexicase selection to 0.443.
+Regarding our primary research question, the differences in generalization, the initial observations do not seem to indicate a significant difference in the the relative gap between the solutions performance on testing data and training data.
 
-$\epsilon$-Lexicase based GP achieves an average minimum fitness of 3.166/1.576 (MSE/MAE) during model training and a score of 2.949/1.678 during the testing phase on previously unseen data.
-Suprisingly the models that are trained using MSE as a metric for fitness evaluation, on average, perform even better on unseen cases as they do on the initial training data.
-
-In comparison, the models that are trained using tournament selection perform with an average minimum fitness of 10.9887/2.32173 (MSE/MAE) during model training and 11.146/2.196 during the testing phase.
-Again we observe an improved performance in testing error, this time models that are evolved using MAE for fitness perform better on testing data than on their initial training data.
-
-The initial results seem support the two alternative hypothesis of this research project:
-
-- $HA_{testing}$: T $\epsilon$-Lexicase Selection produces models that differ in their generalization behaviour if compared to models that are evolved using torunament selection
-
-- $H0_{training}$: $\epsilon$-Lexicase Selection produces models that differ in their performance during model training if compared to models that are evolved using torunament selection
-
---->
-
-## Inferential Statistics
-
-Figure 1 visualizes the distribution of fitness scores achieved by both GP systems after finishing 100 full generations for all 50 runs of the experiment.
+Figure 1 visualizes the distribution of $min\_training\_error$ and $testing\_error$ for both selection operators over the total 50 runs using a boxplot and summerizes the initial observations from tables 6 and 7 described above.
 
@@
 
-... interpretation...
-
 The four samples from figure 1 have been tested for normal distribution by computing the D’Agostino and Pearson test [@DAgostino1971AnOT] [@DAgostino1973TestsFD] to determine if they satisfy the requirements of the students t-test.
 The results are summarized in table 8.
-Since not all samples test positive for a normal distribution a Mann-Whitney U (MWU) ranksum test will be performed to test for statistical significance.
+Since not all samples test positive for a normal distribution a Mann-Whitney U (MWU) ranksum test will be performed to test for statistical significance between the four variables $tournament\_min\_training\_error$, $tournament\_testing\_error$, $e-lexicase\_min\_training\_error$ and $e-lexicase\_testing\_error$.
 
@@
 
-The results of the MWU test are summarized in table 9.
+Table 9 displays the resulting P-values from computing the MWU test in a matrix format. The results demonstrate a highly significant difference int
+
+> $H0_{1}$: The distribution underlying the samples of training/testing errors produced by torunament selection is the same as the distribution underlying samples of training/testing errors produced by $\epsilon$-lexicase selection
+
+> $H0_{2}$: The distribution underlying the samples of training errors produced by $\epsilon$-Lexicase/torunament selection is the same as the distribution underlying samples of testing errors produced by $\epsilon$-Lexicase/tournament selection
 
@@
 
-df <- read.csv("./tables/csv/mwu_matrix_size.csv")
+df <- read.csv("./tables/csv/mwu_matrix_error.csv")
 
@@
 
+The mean overall performance of tournament selection appears to be superior to that of $\epsilon$-Lexicase selection which could be considered contradictory evidence to the 
 ... interpretation...

### 2022-06-22 docs/rmd/paper.tex
autogit

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{float}
\usepackage[numbers]{natbib}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Analyzing the influence of Selection on Genetic Programming's
Generalization ability in Symbolic Regression:

A comparison of \(\epsilon\)-Lexicase Selection and Tournament Selection

--------------------------------------------------------

Student: Roman Höhn

Date of Birth: 1991-04-14

Place of Birth: Wiesbaden, Hesse

Student ID: 2712497

Supervisor: David Wittenberg

--------------------------------------------------------

Research Project for Seminar Information Systems (03.996.3299)

Chair of Business Administration and Computer Science

Johannes Gutenberg University Mainz

Summerterm 2022}
\author{}
\date{\vspace{-2.5em}Date of Submission: 2022-06-22}

\begin{document}
\maketitle

\thispagestyle{empty}

\tableofcontents
\thispagestyle{empty}
\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Genetic programming (GP), a subfield of evolutionary computation (EC),
is a metaheuristic that is used to search for computer programs that
solve a given problem by simulating the process of darwinian evolution.
The basic principle of GP is to gradually evolve solutions by repeatedly
selecting parent solutions from a randomized population of computer
programs based on a fitness metric. Then, genetic operators are applied
on the selected solutions to genereate new offspring candidate
solutions. By repeating this process over many generations, GP acts as a
guided search for high fitness solutions throughout the decision space.
A unique feature of GP among other evolutionary optimization procedures
is the possibility to evolve solutions of variable length.

The overall performance of GP can depend strongly on the choice of its
underlying operators, one crucial component in this is the operator for
parent selection. Tournament Selection is a commonly used selection
operator in EC and is the most used operator in GP systems (Fang and Li,
2010, p. 181). A parent solution is selected by randomly sampling \(k\)
individuals from the current population into a tournament pool and then
the solution with the highest fitness score from the tournament pool is
selected (Fang and Li, 2010, p. 182). Lexicase selection has been
suggested as an alternative to tournament selection that is not based on
aggregating fitness scores. It samples \(n\) test cases in random order
and then eliminates solutions from the selection pool on a per test case
basis if they are not performing on an elite level (Helmuth, Spector and
Matheson, 2015, p. 1).

Since regular Lexicase selection has been shown to perform suboptimal on
continous-valued optimization problems, a modified variation called
\(\epsilon\)-Lexicase selection has been suggested for symbolic
regression by La Cava et. al (2016). Here, \(\epsilon\)-Lexicase
selection has shown itself to outperform both in overall performance
while showing only negligible computational overhead (La Cava, Spector
and Danai, 2016, p. 747).

The goal of symbolic regression is to find a mathematical model for an
observed set of datapoints (Paris, Robilliard and Fonlupt, 2004, p.
794). Symbolic Regression has been one of the first GP applications and
to this day is an actively studied and highly relevant area of research
(Poli, Langdon and McPhee, 2008, p. 114). In most symbolic regression
problems, little to no a priori knowledge about the optimal form and
structure of the target function is available. The ability of GP to
optimize for model structure as well as for parameters has lead to it
being one of the most prevalent methods used in the domain of symbolic
regression (Paris, Robilliard and Fonlupt, 2004, p. 795).

An important quality of all supervised machine learning applications,
including GP, is the ability to not only optimize performance for the
test cases a model is trained on but to also perform well on previously
unseen cases, this is refered to as generalization. In most real world
applications of symbolic regression only a small subset of labeled data
is available for training. The aim is to produce a model that not only
accurately predicts the provided training data but can also predict
previously unseen cases with high precision (Gonçalves, 2016, p. 6). A
model that is extensivley optimized on the provided training data, may
overfit to this data sample which may lead to a decrease in
generalization.

This research project tries to answer the question if the usage of
\(\epsilon\)-Lexicase selection influences the generalization behaviour
of programs that are evolved using GP for symbolic regression if
compared to programs that are evolved using traditional tournament
selection.

\hypertarget{an-overview-of-the-current-state-of-research}{%
\section{An Overview of the current state of
research}\label{an-overview-of-the-current-state-of-research}}

\hypertarget{selection}{%
\subsection{Selection}\label{selection}}

In the first description of lexicase\footnote{The alternative, more
  descriptive name given by Spector (2012): global pool, uniform random
  sequence, elitist lexicase parent selection} selection for GP by
Spector (2012) it was suggested as a novel parent selection method for
target problems that are modal in nature. The author classified modal
problems as ``problems that qualitatively different modes of response
are required for inputs from different regions of the problem's
domain''(Spector, 2012, p. 1).

In a following article Lexicase selection has been proposed specifically
for the purpose of solving so called uncompromising problems with GP.
Helmuth, Spector and Matheson (2015) defined uncompromising problems as
problems that require the final solution to perform optimal on each of
the cases it is tested on, examples include symbolic regression, the
design of digital multipliers or finding terms in finite algebras. The
authors also provided evidence that lexicase selection can significantly
improves GP's ability to solve some uncompromising problems if compared
to selection methods that are based on aggregated fitness (Helmuth,
Spector and Matheson, 2015, p. 12).

Both articles argued that many of the problem domains that GP is
commonly used for are prevalent for problems that are modal/
uncompromising in nature. In comparison to other selection operators,
populations that are evolved using variations of the lexicase selection
operator show a very high degree of genetic diversity which might be a
key contributer to the improved performance (Helmuth, Spector and
Matheson, 2015, p. 1) (La Cava, Spector and Danai, 2016, p. 745). In
theory, Lexicase based selection should be more prone to select
solutions that are specialist (high fitness on a small subset of
training cases) than solutions that are generalist (high average fitness
among all training cases).

The first description\footnote{Spector (2012), p.4} of the lexicase
parent selection algorithm by Spector (2012) is quoted below for
reference:

\begin{verbatim}
Lexicase - Parent-Selection:

1. Initialize:
  (a) Set candidates to be the entire population.
  (b) Set cases to be a list of all of the fitness cases in random order.

2. Loop:
  (a) Set candidates to be the subset of the current candidates that have 
      exactly the best fitness of any individual currently in candidates for 
      the first case in cases.
  (b) If candidates or cases contains just a single element then return
      the first individual in candidates.
  (c) Otherwise remove the first case from cases and go to Loop.
\end{verbatim}

More recent research by La Cava, Spector and Danai (2016) suggested
\(\epsilon\)-Lexicase selection as a modified selection operator that
can improve overall GP performance if applied to continous-valued
symbolic regression tasks in comparison to tournament and standard
lexicase selection. La Cava, Spector and Danai (2016) argued that the
original concept behind lexicase selection as formulated by Spector
(2012) does not fit the requirements of real-world symbolic regression
tasks. The authors identified the pass condition of regular lexicase
selection as the main problem if applied to symbolic regression:
Individual solutions can only be selected if they perform on an elite
level but with noisey and continous-valued data it is very unlikely for
two individuals to achieve an exactly equal error on a test case, this
finally results in filtering out too many individuals during the
selection process(La Cava, Spector and Danai, 2016, p. 742).
\(\epsilon\)-Lexicase adresses this problem by introducing an
\(\epsilon\) parameter that specifies a range around the elite error,
individuals that perform inside this range pass the current selection
iteration. La Cava \emph{et al.} (2017) describe \(\epsilon\)-Lexicase
selection as a more ``relaxed version of lexicase selection''. Different
methods to configure the \(\epsilon\) Parameter have been explored, for
the limited scope of this project I will focus on the most promising
implementation of \(\epsilon\)-Lexicase selection that is automatically
adjusted on the basis of the median absolute deviation of errors inside
the selection pool (La Cava, Spector and Danai, 2016, p. 742) (La Cava
\emph{et al.}, 2017, p. 6).

The performance increase of \(\epsilon\)-Lexicase selection for symbolic
regression problems has been demonstrated and reported for many
benchmark problems which led to widespread adaption of it in symbolic
regression applications (La Cava, Spector and Danai, 2016, pp.
744--745).

In comparison, traditional GP selection methods such as tournament
selection most commonly compute the fitness of a program as the mean of
its error for each individual fitness case. One downside to this
approach is that the total amount of information is reduced from a wide
range of individual errors to a single metric. Helmuth, Spector and
Matheson (2015) suspected that this loss in overall information provided
to GP might reduce overall performance especially if applied to the
class of uncompromising problems that require the solution to perform
well on a wide range of diverse cases. Since tournament selection
selects individuals based on their average fitness across all test
cases, the resulting solutions should be expected to be more biased
towards being generalists instead of specialists.

The basic algorithm for tournament is given below for comparison (Fang
and Li, 2010, pp. 182--183):

\begin{verbatim}
Tournament - Parent-Selection:

k = tournament size

1. Sample:
  (a) Randomly sample k individuals from the current population
      into a tournament pool.
2. Select:
  (a) Compute the mean fitness for each individual inside the tournament pool
      based on all fitness cases.
  (b) Select individual from the tournament pool with the highest mean fitness.
\end{verbatim}

\hypertarget{generalization}{%
\subsection{Generalization}\label{generalization}}

Generalization, the ability of a model to perform well on previously
unseen cases, is one of the fundamental goals in most real world machine
learning applications. O'Neill \emph{et al.} (2010) raised awareness to
the fact, that the topic of generalization in GP has not gotten the
attention that other machine learning-based areas, e.g, deep learning,
have attributed to it. A similiar statement was proposed by Kushchu
(2002) in his review of research in generalization in GP. Both authors
adressed the need for additional research regarding the topic.

Kushchu (2002) critized that almost all applications published in the
initial GP literature by John Koza (Koza, 1992) are not using separate
training and testing datasets which might result in overfitting and a
poor overall generalization ability of the programs that are produced.
The author calls for a more widespread adoption of methods like
generational sampling of new training cases or the overall separation
into training and testing datasets to improve generalization in GP
(Kushchu, 2002, p. 10).

An aspect that is suspected to negatively correlate with GP's
generalization is overall size and in particular bloating of the
resulting programs. Bloating can be described as a growth in the total
size of a program that does not also improve it's performance in any
meaningful way (Silva and Vanneschi, 2009, p. 1). It has been widely
suspected that GP configurations that produce very large programs also
have a higher tendency to specialize on difficult or unusual test cases
which could result in lower generalization (Wang, Wagner and Rondinelli,
2019, p. 268). The minimum description length principle (MDLP) describes
this phenomenom by claiming that less complex models are more likely to
perform better at generalizing than more complex models that achieve a
comparable fitness during the training phase (O'Neill \emph{et al.},
2010, p. 349). However MDLP should be considered carefully, experiments
by Silva and Vanneschi (2009) on the prediction of bioavailability for
drugs demonstrated that GP based techniques can either bloat without
overfitting or overfit without bloating (Silva and Vanneschi, 2009, p.
8).

\hypertarget{symbolic-regression}{%
\subsection{Symbolic Regression}\label{symbolic-regression}}

The task of finding a mathematical function that fits a given set of
datapoints has been one of the first applications of GP in (Koza, 1992,
p. 238). Practical examples given by Koza (1992) included tasks such as
the discovery of scientific laws, finding solutions to differential and
integral equations or the programmatic compression of images.

Since it's inception, GP based symbolic regression has been widely used
in practical application ranging over a wide field of disciplines. A
well known example of an industrial use-case has been published by
Jordaan \emph{et al.} (2004). The authors used GP based symbolic
regression to derive nonlinear functions that could be used to
significantly increase the robustness of industrial sensors which
resulted in wide-spread adoption and a significant reduction in costs.

Many variations of GP based symbolic regression have been proposed to
improve overall performance, Wang, Wagner and Rondinelli (2019)
summarized the main characteristics and differences for 6 common
variations including procedures such as geometric semantic genetic
programming, cartesian genetic programming or GP-based relevance vector
machines. A large benchmarking study on symbolic regression published by
Orzechowski, Cava and Moore (2018) compared 4 GP-based procedures to
several state-of-the-art machine leaning techniques. The authors find
that GP, eventhough more time consuming, can achieve better results if
compared to the state-of-the-art machine learning algorithm
gradient-boosting.

\hypertarget{experimental-study}{%
\section{Experimental study}\label{experimental-study}}

\hypertarget{research-design}{%
\subsection{Research Design}\label{research-design}}

To study the influence of selection on generalization I selected a
dataset about energy efficiency in buildings that is part of the UC
Irvine Machine Learning Repository (Dua and Graff, 2017). The dataset
contains eight individual building attributes that map to two different
outcomes, heating and cooling load, for N=768 cases (Tsanas and Xifara,
2012).

Using two otherwise identical GP systems, one deploying tournament
selection and the other \(\epsilon\)-lexicase selection, the objective
is to find a computer program that best predicts the outcome variable
heating load \((Y1)\)\footnote{The symbolic regression is performed on
  one of the two provided outcome variables, the variable Cooling Load
  will be excluded.} of the buildings using a subset of the eight
building attributes \((X1,..,X8)\) for input. The specific meaning of
all attributes inside the dataset are described in table 1.

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating Dataset}
\centering
\begin{tabular}[t]{l|l}
\hline
\textbf{Variable} & \textbf{Description}\\
\hline
X1 & Relative Compactness\\
\hline
X2 & Surface Area\\
\hline
X3 & Wall Area\\
\hline
X4 & Roof Area\\
\hline
X5 & Overall Height\\
\hline
X6 & Orientation\\
\hline
X7 & Glazing Area\\
\hline
X8 & Glazing Area Distribution\\
\hline
y1 & Heating Load\\
\hline
y2 & Cooling Load\\
\hline
\end{tabular}
\end{table}

To measure the generalization ability of each model the dataset will be
randomly split in half, resulting in a training and testing dataset each
containing 384 individual cases. Each model will be evolved by
traditional GP using only the fitness cases present in the training
dataset. For each generation the highest fitness model of the current
population will be tested with the previously unseen fitness cases that
are part of the testing dataset. For each run of the experiment
statistics will be collected on fitness which will form the basis of my
further statistical analysis. Additional statistics will be collected
for the average length of the population and the length of the elite
program for each generation to explore differences in their distribution
and possible correlations to generalization.

Since GP is a stochastic optimization algorithm, the basic experiment
will be run for a total of 50 times to ensure a fair and meaningful
comparison based on a large number of runs for both algorithms. For each
run of the experiment the dataset will be randomly split in half as
described, both models are then trained and tested using the exact same
set of fitness cases.

The statistical analysis of the collected data will first focus on
examining the question if, on average, the usage of
\(\epsilon\)-Lexicase selection will result in models that perform
significantly different than models that are evolved using tournament
selection both.

\begin{quote}
\(H0_{1}\): The distribution underlying the samples of training/testing
errors produced by torunament selection is the same as the distribution
underlying samples of training/testing errors produced by
\(\epsilon\)-lexicase selection
\end{quote}

The next question of interest is, to examine if statistical significant
differences between the mean errors of training and testing data exist
for both selection operators:

\begin{quote}
\(H0_{2}\): The distribution underlying the samples of training errors
produced by \(\epsilon\)-Lexicase/torunament selection is the same as
the distribution underlying samples of testing errors produced by
\(\epsilon\)-Lexicase/tournament selection
\end{quote}

To gather further insight into the differences between both GP systems,
an additional test will be performed on the question if differences in
the total size of the resulting programs exist:

\begin{quote}
\(H0_{3}\): Size differences exist between the distribution underlying
the samples produced by \(\epsilon\)-Lexicase and the distribution
underlying samples of tournament selection
\end{quote}

All hypothesis will be tested using a level of significance of
\(\alpha=0.05\).

To further examine the difference in generalization behaviour, the mean
testing and training errors over each generation will be visualized for
both algorithms. The specific aim of this visualization is to explore
the mechanism of overfitting and to examine if differences between both
methods can be detected.

All GP experiments will be implemented by using the python programming
language in conjunction with DEAP, a framework for distributed
evolutionary algorithms that implements various tools and algorithms for
genetic programming (Fortin \emph{et al.}, 2012).

\hypertarget{gentetic-programming-configuration}{%
\subsection{Gentetic Programming
Configuration}\label{gentetic-programming-configuration}}

\hypertarget{evolutionary-parameters}{%
\subsubsection{Evolutionary parameters}\label{evolutionary-parameters}}

The basic evolutionary parameters for both systems are presented in
table 2. Tournament selection is used with a default tournament size of
\(3\) individuals while the \(\epsilon\) parameter in lexicase selection
is selected automatically as previously mentioned in subsection 3.1.

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-2}Evolutionary Parameters}
\centering
\begin{tabular}[t]{l|l}
\hline
\textbf{Parameter} & \textbf{Value}\\
\hline
Population Size & 500\\
\hline
Number of Generations & 100\\
\hline
Mutation Rate & 20\%\\
\hline
Crossover Rate & 80\%\\
\hline
Tournament Size & 3\\
\hline
Epsilon selection & automatic\\
\hline
Elite Size & 0\\
\hline
\end{tabular}
\end{table}

\hypertarget{fitness-evaluation}{%
\subsubsection{Fitness Evaluation}\label{fitness-evaluation}}

The fitness \(f\) for each model will be based on the mean squared error
(MSE) over all fitness cases for prediction and the empirically measured
values as described by eq.1.

\begin{equation}
\tag{eq. 1}
MSE = \frac{1}{n} * \sum_{i=1}^{n} (Y_i - \hat{Y_1})^2
\end{equation}

where:

\begin{itemize}
\tightlist
\item
  \(n\): total number of test cases
\item
  \(Y_i\): empirical value for case \(i\)
\item
  \(\hat{Y_1}\): predicted value for case \(i\).
\end{itemize}

The resulting fitness function \(f\) for an individual program \(i\) is
descibed by eq 2:

\begin{equation}
\tag{eq. 2}
 f(i, \tau ) = \frac{1}{N} * \sum_{t \epsilon \tau} (y_t - y\hat{}_t(i, x_t))^2 
\end{equation}

where\footnote{Naming and notation was adopted from La Cava, Spector and
  Danai (2016)}:

\begin{itemize}
\tightlist
\item
  \(\tau\): the set of \(N\) fitness cases
\item
  \(y_t\): empirical value of the target for case \(t\)
\item
  \(y\hat{}_t(i, x_t)\): predicted value for the target for case \(t\)
  by running the program \(i\) with the total set of input variables
  \(x_t\)
\end{itemize}

\hypertarget{primitive-set}{%
\subsubsection{Primitive Set}\label{primitive-set}}

The primitve set consists of the terminals that are listed in table 3
and the functions that are listed in table 4. To avoid runtime errors I
implemented protected version of the operators for division, natural
logarithm and square root (Koza, 1992, pp. 82--83). As suggested by Koza
(1992) I also included ephemeral constants to the set of terminals to
provide the evolutionay search the opportunity to explore and include
randomly generated constants.

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-3}Terminals}
\centering
\begin{tabular}[t]{l|l}
\hline
\textbf{Terminal} & \textbf{Description}\\
\hline
X1 & Relative Compactness\\
\hline
X2 & Surface Area\\
\hline
X3 & Wall Area\\
\hline
X4 & Roof Area\\
\hline
X5 & Overall Height\\
\hline
X6 & Orientation\\
\hline
X7 & Glazing Area\\
\hline
X8 & Glazing Area Distribution\\
\hline
random\_int & Ephemeral Constant (integer)\\
\hline
random\_float & Ephemeral Constant(float)\\
\hline
\end{tabular}
\end{table}

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-4}Functions}
\centering
\begin{tabular}[t]{l|r}
\hline
\textbf{Function} & \textbf{Arity}\\
\hline
Addition & 2\\
\hline
Subtraction & 2\\
\hline
Multiplication & 2\\
\hline
Negation & 1\\
\hline
Sine & 1\\
\hline
Cosine & 1\\
\hline
Protected Division & 2\\
\hline
Protected Natural Logarithm & 1\\
\hline
Protected Square Root & 1\\
\hline
\end{tabular}
\end{table}

\hypertarget{genetic-operators}{%
\subsubsection{Genetic Operators}\label{genetic-operators}}

GP operators are represented in table 5. Both genetic operators use a
static limit to control for the height of the reulting trees (Koza,
1992, p. 104). Individual programs are initialized by using the ramped
half-and-half method, 50\% of the population are created by using the
Growth algorithm and the remaining 50\% are created by using the Full
algorithm (Koza, 1992, p. 93).

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-5}GP Operators}
\centering
\begin{tabular}[t]{l|l|r}
\hline
\textbf{Operator} & \textbf{Implementation} & \textbf{Static.Height.Limit}\\
\hline
Initilization & Ramped Half/Half & 2\\
\hline
Crossover & One Point Crossover & 17\\
\hline
Mutation & Uniform Mutation & 17\\
\hline
\end{tabular}
\end{table}

The crossover operator implemented by DEAP randomly selects a crossover
point in each individual and exchanges each subtree with the point as
root between each individual (Fortin \emph{et al.}, 2012). Mutation also
randomly selects a point in the tree individual, it then replaces the
subtree below that point as a root by the expression generated using the
full grow initialization method (Fortin \emph{et al.}, 2012).

\hypertarget{results}{%
\section{Results}\label{results}}

Tables 6 and 7 summarize the results for all fitness scores collected
over 50 total runs of the experiment\footnote{All floating point numbers
  in tables have been rounded to 3 decimal places if not stated
  otherwise}.

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-6}Summary - Tournament Selection}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrrrrrrrr}
\toprule
\textbf{X} & \textbf{gen} & \textbf{nevals} & \textbf{mean\_training\_error} & \textbf{std\_training\_error} & \textbf{min\_training\_error} & \textbf{max\_training\_error} & \textbf{testing\_error} & \textbf{std\_testing\_error} & \textbf{avg\_size} & \textbf{elite\_size}\\
\midrule
count & 101.0 & 101.000 & 1.010000e+02 & 1.010000e+02 & 101.000 & 1.010000e+02 & 100.000 & 100.000 & 100.000 & 100.000\\
mean & 50.0 & 420.742 & 2.092614e+18 & 3.305357e+20 & 16.483 & 1.046309e+21 & 16.905 & 28.858 & 55.674 & 62.070\\
std & 29.3 & 8.115 & 2.098028e+19 & 3.313954e+21 & 12.514 & 1.049016e+22 & 11.117 & 15.732 & 34.504 & 36.850\\
min & 0.0 & 417.300 & 2.582030e+08 & 1.463390e+10 & 7.066 & 1.193720e+11 & 7.847 & 16.217 & 3.331 & 4.340\\
25\% & 25.0 & 418.780 & 2.151894e+10 & 3.044239e+12 & 8.479 & 1.070232e+13 & 9.350 & 18.344 & 25.133 & 28.195\\
\addlinespace
50\% & 50.0 & 420.000 & 3.494635e+11 & 5.511534e+13 & 11.918 & 1.747285e+14 & 12.874 & 22.715 & 57.299 & 66.400\\
75\% & 75.0 & 421.000 & 4.892936e+12 & 6.455802e+14 & 18.773 & 2.238973e+15 & 19.685 & 32.750 & 87.852 & 96.625\\
max & 100.0 & 500.000 & 2.108540e+20 & 3.330558e+22 & 77.243 & 1.054272e+23 & 62.693 & 95.282 & 109.761 & 115.340\\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-7}Summary - Epsilon-Lexicase Selection}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrrrrrrrr}
\toprule
\textbf{X} & \textbf{gen} & \textbf{nevals} & \textbf{mean\_training\_error} & \textbf{std\_training\_error} & \textbf{min\_training\_error} & \textbf{max\_training\_error} & \textbf{testing\_error} & \textbf{std\_testing\_error} & \textbf{avg\_size} & \textbf{elite\_size}\\
\midrule
count & 101.0 & 101.000 & 1.010000e+02 & 1.010000e+02 & 101.000 & 1.010000e+02 & 100.000 & 100.000 & 100.000 & 100.000\\
mean & 50.0 & 420.781 & 7.533781e+12 & 1.186094e+15 & 22.013 & 3.762638e+15 & 22.456 & 39.048 & 58.416 & 63.819\\
std & 29.3 & 8.109 & 5.900311e+13 & 9.313627e+15 & 16.670 & 2.950203e+16 & 15.904 & 19.225 & 34.515 & 36.356\\
min & 0.0 & 416.240 & 4.660413e+05 & 3.314323e+07 & 9.680 & 1.539125e+08 & 11.129 & 21.487 & 3.219 & 4.280\\
25\% & 25.0 & 418.900 & 2.528193e+08 & 2.727451e+10 & 11.316 & 1.192373e+11 & 12.089 & 24.860 & 28.233 & 29.615\\
\addlinespace
50\% & 50.0 & 420.040 & 2.196352e+09 & 2.647194e+11 & 14.478 & 1.098068e+12 & 15.223 & 30.057 & 64.689 & 71.440\\
75\% & 75.0 & 421.020 & 2.027020e+11 & 3.198174e+13 & 23.511 & 1.012976e+14 & 23.305 & 51.081 & 87.394 & 94.910\\
max & 100.0 & 500.000 & 5.788911e+14 & 9.137151e+16 & 74.622 & 2.894465e+17 & 77.518 & 110.244 & 108.785 & 115.660\\
\bottomrule
\end{tabular}}
\end{table}

Some interesting observation from tables 6 and 7 can be made by
comparing the mean of the variables \(min\_training\_error\) and
\(testing\_error\). Variable \(min\_training\_error\) represents the
fitness, measured by the MSE, of the best performing solution that has
been found during a single run of the experiment while the variable
\(testing\_error\) represents the fitness of the same solution if
evaluated for the unknown fitness cases inside the testing dataset.
Tournament selection based GP achieves a mean \(min\_training\_error\)
of \(16.483\) and a mean \(testing\_error\) of \(16.905\) while the GP
system using \(\epsilon\)-Lexicase selection can only achieve a mean
\(min\_training\_error\) of \(22.013\) and a mean \(testing\_error\) of
\(22.456\). These results indicate an overall performance benefit for
torunament selection based GP in the experiment. It should also be
mentioned that the best performing model over all 50 runs, measured by
the minimum of variables \(min\_training\_error\) and
\(testing\_error\), has been achieved by tournament selection with an
MSE of 7.066 for the training data and 7.847 for the testing data.

If we compare the differences of \(min\_training\_error\) and
\(testing\_error\) on the basis of each algorithm, we can observe that
both tournament and \(\epsilon\)-lexicase achieve a slightly lower error
on the training data than on the testing data. If we compute the total
difference between the means of \(min\_training\_error\) and
\(testing\_error\) for both algorithms tournament selection amounts to
0.422 and \(\epsilon\)-Lexicase selection to 0.443. Regarding our
primary research question, the differences in generalization, the
initial observations do not seem to indicate a significant difference in
the the relative gap between the solutions performance on testing data
and training data.

Figure 1 visualizes the distribution of \(min\_training\_error\) and
\(testing\_error\) for both selection operators over the total 50 runs
using a boxplot and summerizes the initial observations from tables 6
and 7 described above.

\begin{figure}
\centering
\includegraphics{./plots/mean_error_boxplot_all.png}
\caption{Distribution of Errors}
\end{figure}

The four samples from figure 1 have been tested for normal distribution
by computing the D'Agostino and Pearson test (D'Agostino, 1971)
(D'Agostino and Pearson, 1973) to determine if they satisfy the
requirements of the students t-test. The results are summarized in table
8. Since not all samples test positive for a normal distribution a
Mann-Whitney U (MWU) ranksum test will be performed to test for
statistical significance between the four variables
\(tournament\_min\_training\_error\), \(tournament\_testing\_error\),
\(e-lexicase\_min\_training\_error\) and \(e-lexicase\_testing\_error\).

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-8}Normal Distribution Tests}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrl}
\toprule
\textbf{sample} & \textbf{statistic} & \textbf{p.value} & \textbf{alpha} & \textbf{normal\_distributed}\\
\midrule
tournament\_min\_training\_error & 36.146 & 0.000 & 0.05 & False\\
e-lexicase\_min\_testing\_error & 1.575 & 0.455 & 0.05 & True\\
tournament\_testing\_error & 32.503 & 0.000 & 0.05 & False\\
e-lexicase\_testing\_error & 59.135 & 0.000 & 0.05 & False\\
\bottomrule
\end{tabular}}
\end{table}

The results of the MWU test are summarized in table 9.

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-9}Mean Error - P-Values}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrr}
\toprule
\textbf{ } & \textbf{tournament\_training\_errors} & \textbf{tournament\_testing\_errors} & \textbf{elexicase\_training\_errors} & \textbf{elexicase\_testing\_errors}\\
\midrule
tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
\bottomrule
\end{tabular}}
\end{table}

The mean overall performance of tournament selection appears to be
superior to that of \(\epsilon\)-Lexicase selection which could be
considered contradictory evidence to the \ldots{} interpretation\ldots{}

To further examine the generalization behaviour based on both selection
operators I plotted the mean fitness scores for both algorithms in
figure 2.

\begin{figure}
\centering
\includegraphics{./plots/mean_error_combined.png}
\caption{Mean Errors}
\end{figure}

\ldots{} interpretation\ldots{}

In a final step I analyzed the growth behaviour for both GP systems.
Figure 3 visualizes the average size of individuals inside the
population as well as the average size of the best performing individual
for each generation of the evolution.

\includegraphics{./plots/size_combined.png} \ldots{}
interpretation\ldots{}

Again, a MWU test is conducted to test for statistical differences in
the underlying distribution of the sizes measured during the experiment.
The results are summarized in table 10.

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-10}Mean Size - P-Values}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrr}
\toprule
\textbf{ } & \textbf{tournament\_elite\_size} & \textbf{elexicase\_elite\_size} & \textbf{tournament\_avg\_size} & \textbf{elexicase\_avg\_size}\\
\midrule
tournament\_elite\_size & 1.000 & 0.858 & 0.533 & 0.652\\
elexicase\_elite\_size & 0.858 & 1.000 & 0.764 & 0.682\\
tournament\_avg\_size & 0.533 & 0.764 & 1.000 & 0.992\\
elexicase\_avg\_size & 0.652 & 0.682 & 0.992 & 1.000\\
\bottomrule
\end{tabular}}
\end{table}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\hypertarget{limitations-and-open-questions}{%
\section{Limitations and open
questions}\label{limitations-and-open-questions}}

I faced two major difficulties in the preparation and evaluation of this
research project:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Configuration of evolutionary parameters
\item
  Limited computational ressources
\end{enumerate}

The combination of both issues might have resulted in a significant
reduction in overall robustness of the results represented in sections 5
and 6. The performance of all evolutionary algorithms, including GP, can
be highly dependant on a large number of parameters and implementation
details. Although the series of experiments conducted in this project
are comparably simple in nature, the results achieved might still be
highly influenced by the GP configuration detailed in subsection 4.2.

To gain further insight into the differences in generalizaion behaviour
and to adress the research question with a higher level of certainty,
the experiment I conducted should be repeated for many different GP
configurations. Some exemplary parameters that could drastically
influence the results of this experiment include the population size,
the total number of generations, the inclusion of elitism or the
computation of \(\epsilon\) in lexicase selection. Other factors whose
influence on generalization behaviour could be important include
different methods to compute an individuals fitness (e.g.~mean absolute
error or root mean squared error) and testing out different variations
of the genetic operators for mutation and crossover.

Another obvious weakness of my project is that the experiment is only
based on a single symbolic regression application, the prediction of the
heating load of buildings based on a relativley small dataset. To gain
further trust in the obtained results, the experiment should be repeated
for other symbolic regression tasks and datasets. At the current state
it remains unknown if the generalization behaviour of both selection
operators might be highly problem specific and dependant on the dataset
used for training and testing.

All limitations and shortcomings described above are also connected to
the second limitation of this research project, the limited amount of
computational ressources. The total computation time consumed to run the
current experiments (50 runs, 2 algorithms, 100 generations, 500
individuals) has already been close to 24 hours on a modern, arm-based
Apple-M1 system. Eventhough the overall computation time could probably
be reduced by further optimization of the source code, e.g.~using tools
for parallel computation, the ressources necessary to repeat the whole
experiment for different parameter configurations would certainly be
beyond the scope of this research project.

\newpage

\hypertarget{I}{%
\section*{I References}\label{I}}
\addcontentsline{toc}{section}{I References}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-DAgostino1971AnOT}{}}%
D'Agostino, R.B. (1971) {``An omnibus test of normality for moderate and
large size samples,''} \emph{Biometrika}, 58, pp. 341--348.

\leavevmode\vadjust pre{\hypertarget{ref-DAgostino1973TestsFD}{}}%
D'Agostino, R.B. and Pearson, E.S. (1973) {``Tests for departure from
normality,''} in.

\leavevmode\vadjust pre{\hypertarget{ref-Dua:2019}{}}%
Dua, D. and Graff, C. (2017) {``{UCI} machine learning repository.''}
University of California, Irvine, School of Information; Computer
Sciences. Available at: \url{http://archive.ics.uci.edu/ml}.

\leavevmode\vadjust pre{\hypertarget{ref-10.1007ux2f978-3-642-16493-4_19}{}}%
Fang, Y. and Li, J. (2010) {``A review of tournament selection in
genetic programming,''} in Cai, Z. et al. (eds.) \emph{Advances in
computation and intelligence}. Berlin, Heidelberg: Springer Berlin
Heidelberg, pp. 181--192.

\leavevmode\vadjust pre{\hypertarget{ref-DEAP_JMLR2012}{}}%
Fortin, F.-A. \emph{et al.} (2012) {``{DEAP}: Evolutionary algorithms
made easy,''} \emph{Journal of Machine Learning Research}, 13, pp.
2171--2175.

\leavevmode\vadjust pre{\hypertarget{ref-Gonalves2016AnEO}{}}%
Gonçalves, I. (2016) {``An exploration of generalization and overfitting
in genetic programming: Standard and geometric semantic approaches,''}
in.

\leavevmode\vadjust pre{\hypertarget{ref-6920034}{}}%
Helmuth, T., Spector, L. and Matheson, J. (2015) {``Solving
uncompromising problems with lexicase selection,''} \emph{IEEE
Transactions on Evolutionary Computation}, 19(5), pp. 630--643.
doi:\href{https://doi.org/10.1109/TEVC.2014.2362729}{10.1109/TEVC.2014.2362729}.

\leavevmode\vadjust pre{\hypertarget{ref-soft_sensor_gp}{}}%
Jordaan, E. \emph{et al.} (2004) {``Robust inferential sensors based on
ensemble of predictors generated by genetic programming,''} in Yao, X.
et al. (eds.) \emph{Parallel problem solving from nature - PPSN VIII}.
Berlin, Heidelberg: Springer Berlin Heidelberg, pp. 522--531.

\leavevmode\vadjust pre{\hypertarget{ref-koza_main}{}}%
Koza, J.R. (1992) \emph{Genetic programming: On the programming of
computers by means of natural selection}. Cambridge, MA, USA: MIT Press.
Available at: \url{http://mitpress.mit.edu/books/genetic-programming}.

\leavevmode\vadjust pre{\hypertarget{ref-generalisation_in_gp}{}}%
Kushchu, I. (2002) {``An evaluation of EvolutionaryGeneralisation in
genetic programming,''} \emph{Artificial Intelligence Review - AIR}, 18,
pp. 3--14.
doi:\href{https://doi.org/10.1023/A:1016379201230}{10.1023/A:1016379201230}.

\leavevmode\vadjust pre{\hypertarget{ref-https:ux2fux2fdoi.orgux2f10.48550ux2farxiv.1709.05394}{}}%
La Cava, W. \emph{et al.} (2017) {``A probabilistic and multi-objective
analysis of lexicase selection and epsilon-lexicase selection.''} arXiv.
doi:\href{https://doi.org/10.48550/ARXIV.1709.05394}{10.48550/ARXIV.1709.05394}.

\leavevmode\vadjust pre{\hypertarget{ref-epsilon_lexicase_main}{}}%
La Cava, W., Spector, L. and Danai, K. (2016) {``Epsilon-lexicase
selection for regression,''} in \emph{Proceedings of the genetic and
evolutionary computation conference 2016}. New York, NY, USA:
Association for Computing Machinery (GECCO '16), pp. 741--748.
doi:\href{https://doi.org/10.1145/2908812.2908898}{10.1145/2908812.2908898}.

\leavevmode\vadjust pre{\hypertarget{ref-open_issues_gp}{}}%
O'Neill, M. \emph{et al.} (2010) {``Open issues in genetic
programming,''} \emph{Genetic Programming and Evolvable Machines}, 11,
pp. 339--363.
doi:\href{https://doi.org/10.1007/s10710-010-9113-2}{10.1007/s10710-010-9113-2}.

\leavevmode\vadjust pre{\hypertarget{ref-Orzechowski_2018}{}}%
Orzechowski, P., Cava, W.L. and Moore, J.H. (2018) {``Where are we
now?''} in \emph{Proceedings of the genetic and evolutionary computation
conference}. {ACM}.
doi:\href{https://doi.org/10.1145/3205455.3205539}{10.1145/3205455.3205539}.

\leavevmode\vadjust pre{\hypertarget{ref-10.1007ux2f978-3-540-24621-3_22}{}}%
Paris, G., Robilliard, D. and Fonlupt, C. (2004) {``Exploring
overfitting in genetic programming,''} in Liardet, P. et al. (eds.)
\emph{Artificial evolution}. Berlin, Heidelberg: Springer Berlin
Heidelberg, pp. 267--277.

\leavevmode\vadjust pre{\hypertarget{ref-poli08:fieldguide}{}}%
Poli, R., Langdon, W.B. and McPhee, N.F. (2008) \emph{A field guide to
genetic programming}. Published via \texttt{http://lulu.com}; freely
available at \texttt{http://www.gp-field-guide.org.uk}. Available at:
\url{https://digitalcommons.morris.umn.edu/cgi/viewcontent.cgi?article=1001\&context=cs_facpubs}.

\leavevmode\vadjust pre{\hypertarget{ref-bloat_overfitting_gp}{}}%
Silva, S. and Vanneschi, L. (2009) {``Operator equalisation, bloat and
overfitting: A study on human oral bioavailability prediction,''} in
\emph{Proceedings of the 11th Annual Genetic and Evolutionary
Computation Conference, GECCO-2009}, pp. 1115--1122.
doi:\href{https://doi.org/10.1145/1569901.1570051}{10.1145/1569901.1570051}.

\leavevmode\vadjust pre{\hypertarget{ref-lexicase_first_desciption}{}}%
Spector, L. (2012) {``Assessment of problem modality by differential
performance of lexicase selection in genetic programming: A preliminary
report.''}
doi:\href{https://doi.org/10.1145/2330784.2330846}{10.1145/2330784.2330846}.

\leavevmode\vadjust pre{\hypertarget{ref-fe8fa39e88a040bbacba5a465c48043f}{}}%
Tsanas, A. and Xifara, A. (2012) {``Accurate quantitative estimation of
energy performance of residential buildings using statistical machine
learning tools,''} \emph{Energy and buildings}, 49, pp. 560--567.
doi:\href{https://doi.org/10.1016/j.enbuild.2012.03.003}{10.1016/j.enbuild.2012.03.003}.

\leavevmode\vadjust pre{\hypertarget{ref-wang_wagner_rondinelli_2019}{}}%
Wang, Y., Wagner, N. and Rondinelli, J.M. (2019) {``Symbolic regression
in materials science,''} \emph{MRS Communications}, 9(3), pp. 793--805.
doi:\href{https://doi.org/10.1557/mrc.2019.85}{10.1557/mrc.2019.85}.

\end{CSLReferences}

\newpage

\hypertarget{II}{%
\section*{II Statutory Declaration}\label{II}}
\addcontentsline{toc}{section}{II Statutory Declaration}

\end{document}

### 2022-06-22 docs/rmd/paper.Rmd
autogit

@@
 
-Table 9 displays the resulting P-values from computing the MWU test in a matrix format. The results demonstrate a highly significant difference int
-
-> $H0_{1}$: The distribution underlying the samples of training/testing errors produced by torunament selection is the same as the distribution underlying samples of training/testing errors produced by $\epsilon$-lexicase selection
-
-> $H0_{2}$: The distribution underlying the samples of training errors produced by $\epsilon$-Lexicase/torunament selection is the same as the distribution underlying samples of testing errors produced by $\epsilon$-Lexicase/tournament selection
+Table 9 displays the resulting p-values from computing the MWU test in a matrix format.
 
@@
 
-The mean overall performance of tournament selection appears to be superior to that of $\epsilon$-Lexicase selection which could be considered contradictory evidence to the 
-... interpretation...
+The results show two major findings:
 
-To further examine the generalization behaviour based on both selection operators I plotted the mean fitness scores for both algorithms in figure 2.
+1.
+The differences in mean fitness between tournament selection and $\epsilon$-Lexicase selection are highly statistical significant even if considering an $\alpha$ of 0.01.
+For the given symbolic regression task tournament selection achieves a better mean performance on both the training as well as the testing dataset than $\epsilon$-Lexicase selection.
+These findings contradict the hypothesis $H0_{1}$ descibed in subsection 3.1 and are contrary to some of the research that was summarized in subsections 2.1 and 2.3.[@epsilon_lexicase_main] [@https://doi.org/10.48550/arxiv.1709.05394].
+
+2.
+The differences in mean fitness between $min\_training\_error$ and $testing\_error$ are not statistically significant for both selection algorithms.
+If comparing the mean training error to the mean testing error, the p-values only computes to 0.309 for tournament selection and to 0.257 for $\epsilon$-Lexicase selection.
+These findings fail to reject $H0_{2}$, based on the obtained results we can not find evidence that both algorithms demonstrate a significantly different performance between the training data and the previously unseen testing data.
+
+To further examine the generalization behaviour of both selection operators I plotted the mean fitness scores for both algorithms over the total number of 100 generations in figure 2.
+Figure 2 shows that both algorithms show little visual evidence of overfitting which would result in a growing gap between between $min\_training\_error$ and $testing\_error$ in the final stages of the evolutionary optimization.
+For $\epsilon$-Lexicase selection a small gap can be observed starting at around the 90.th generation, while the minimum training error slowly converges a small increase in the testing error can be observed.
+Based on the limited amount of collected data, it is questionable if this small sign of overfitting can be interpreted as a weak indicator for an inferior generalization behaviour of $\epsilon$-Lexicase in comparison to tournament selection.
+
+Figure 2 also shows a visual representation of the described overall performance benefit of using tournament selection instead of $\epsilon$-Lexicase selection for the choosen symbolic regression task.
+It can be observed that at all stages of the evolution tournament selection based GP achieves a lower mean error for both datasets.
 
 ![Mean Errors](./plots/mean_error_combined.png)
-
-... interpretation...
 

### 2022-06-22 docs/rmd/paper.tex
autogit

@@
 
-The results of the MWU test are summarized in table 9.
+Table 9 displays the resulting p-values from computing the MWU test in a
+matrix format.
 
@@
 
-The mean overall performance of tournament selection appears to be
-superior to that of \(\epsilon\)-Lexicase selection which could be
-considered contradictory evidence to the \ldots{} interpretation\ldots{}
-
-To further examine the generalization behaviour based on both selection
-operators I plotted the mean fitness scores for both algorithms in
-figure 2.
+The results show two major findings:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  The differences in mean fitness between tournament selection and
+  \(\epsilon\)-Lexicase selection are highly statistical significant
+  even if considering an \(\alpha\) of 0.01. For the given symbolic
+  regression task tournament selection achieves a better mean
+  performance on both the training as well as the testing dataset than
+  \(\epsilon\)-Lexicase selection. These findings contradict the
+  hypothesis \(H0_{1}\) descibed in subsection 3.1 and are contrary to
+  some of the research that was summarized in subsections 2.1 and
+  2.3.(La Cava, Spector and Danai, 2016) (La Cava \emph{et al.}, 2017).
+\item
+  The differences in mean fitness between \(min\_training\_error\) and
+  \(testing\_error\) are not statistically significant for both
+  selection algorithms. If comparing the mean training error to the mean
+  testing error, the p-values only computes to 0.309 for tournament
+  selection and to 0.257 for \(\epsilon\)-Lexicase selection. These
+  findings fail to reject \(H0_{2}\), based on the obtained results we
+  can not find evidence that both algorithms demonstrate a significantly
+  different performance between the training data and the previously
+  unseen testing data.
+\end{enumerate}
+
+To further examine the generalization behaviour of both selection
+operators I plotted the mean fitness scores for both algorithms over the
+total number of 100 generations in figure 2. Figure 2 shows that both
+algorithms show little visual evidence of overfitting which would result
+in a growing gap between between \(min\_training\_error\) and
+\(testing\_error\) in the final stages of the evolutionary optimization.
+For \(\epsilon\)-Lexicase selection a small gap can be observed starting
+at around the 90.th generation, while the minimum training error slowly
+converges a small increase in the testing error can be observed. Based
+on the limited amount of collected data, it is questionable if this
+small sign of overfitting can be interpreted as a weak indicator for an
+inferior generalization behaviour of \(\epsilon\)-Lexicase in comparison
+to tournament selection.
+
+Figure 2 also shows a visual representation of the described overall
+performance benefit of using tournament selection instead of
+\(\epsilon\)-Lexicase selection for the choosen symbolic regression
+task. It can be observed that at all stages of the evolution tournament
+selection based GP achieves a lower mean error for both datasets.
 
@@
 \end{figure}
-
-\ldots{} interpretation\ldots{}
 

### 2022-06-22 docs/rmd/paper.Rmd
autogit

@@
 <!-- GP --->
-Genetic programming (GP), a subfield of evolutionary computation (EC), is a metaheuristic that is used to search for computer programs that solve a given problem by simulating the process of darwinian evolution.
+Genetic programming (GP), a sub field of evolutionary computation (EC), is a meta heuristic that is used to search for computer programs that solve a given problem by simulating the process of Darwinian evolution.
 The basic principle of GP is to gradually evolve solutions by repeatedly selecting parent solutions from a randomized population of computer programs based on a fitness metric.
-Then, genetic operators are applied on the selected solutions to genereate new offspring candidate solutions.
+Then, genetic operators are applied on the selected solutions to generate new offspring candidate solutions.
 By repeating this process over many generations, GP acts as a guided search for high fitness solutions throughout the decision space.
@@
 
-Since regular Lexicase selection has been shown to perform suboptimal on continous-valued optimization problems, a modified variation called $\epsilon$-Lexicase selection has been suggested for symbolic regression by La Cava et.
+Since regular Lexicase selection has been shown to perform sub optimal on continuous-valued optimization problems, a modified variation called $\epsilon$-Lexicase selection has been suggested for symbolic regression by La Cava et.
 al (2016).
@@
 <!-- Symbolic Regression --->
-The goal of symbolic regression is to find a mathematical model for an observed set of datapoints [@10.1007/978-3-540-24621-3_22, p.794].
+The goal of symbolic regression is to find a mathematical model for an observed set of data points [@10.1007/978-3-540-24621-3_22, p.794].
 Symbolic Regression has been one of the first GP applications and to this day is an actively studied and highly relevant area of research [@poli08:fieldguide, p.114].
-In most symbolic regression problems, little to no a priori knowledge about the optimal form and structure of the target function is available.
+In most symbolic regression problems, little to no prior knowledge about the optimal form and structure of the target function is available.
 The ability of GP to optimize for model structure as well as for parameters has lead to it being one of the most prevalent methods used in the domain of symbolic regression [@10.1007/978-3-540-24621-3_22, p.795].
@@
 <!-- Generalization --->
-An important quality of all supervised machine learning applications, including GP, is the ability to not only optimize performance for the test cases a model is trained on but to also perform well on previously unseen cases, this is refered to as generalization.
+An important quality of all supervised machine learning applications, including GP, is the ability to not only optimize performance for the test cases a model is trained on but to also perform well on previously unseen cases, this is referred to as generalization.
 In most real world applications of symbolic regression only a small subset of labeled data is available for training.
 The aim is to produce a model that not only accurately predicts the provided training data but can also predict previously unseen cases with high precision [@Gonalves2016AnEO, p.6].
-A model that is extensivley optimized on the provided training data, may overfit to this data sample which may lead to a decrease in generalization.
+A model that is extensively optimized on the provided training data, can become over fitted to this data sample which may lead to a decrease in generalization.
 
 <!-- Research Question --->
-This research project tries to answer the question if the usage of $\epsilon$-Lexicase selection influences the generalization behaviour of programs that are evolved using GP for symbolic regression if compared to programs that are evolved using traditional tournament selection.
+This research project tries to answer the question if the usage of $\epsilon$-Lexicase selection influences the generalization behavior of programs that are evolved using GP for symbolic regression if compared to programs that are evolved using traditional tournament selection.
 
@@
 In the first description of lexicase^[The alternative, more descriptive name given by @lexicase_first_desciption: global pool, uniform random sequence, elitist lexicase parent selection] selection for GP by @lexicase_first_desciption it was suggested as a novel parent selection method for target problems that are modal in nature.
-The author classified modal problems as "problems that qualitatively different modes of response are required for inputs from different regions of the problem's domain"[@lexicase_first_desciption, p.1].
+The author classifies modal problems as "problems that qualitatively different modes of response are required for inputs from different regions of the problem's domain"[@lexicase_first_desciption, p.1].
 
 In a following article Lexicase selection has been proposed specifically for the purpose of solving so called uncompromising problems with GP.
-@6920034 defined uncompromising problems as problems that require the final solution to perform optimal on each of the cases it is tested on, examples include symbolic regression, the design of digital multipliers or finding terms in finite algebras.
-The authors also provided evidence that lexicase selection can significantly improves GP's ability to solve some uncompromising problems if compared to selection methods that are based on aggregated fitness [@6920034, p.12].
+@6920034 defines uncompromising problems as problems that require the final solution to perform optimal on each of the cases it is tested on, examples include symbolic regression, the design of digital multipliers or finding terms in finite algebras.
+The authors also provide evidence that lexicase selection can significantly improves GP's ability to solve some uncompromising problems if compared to selection methods that are based on aggregated fitness [@6920034, p.12].
 
-Both articles argued that many of the problem domains that GP is commonly used for are prevalent for problems that are modal/ uncompromising in nature.
-In comparison to other selection operators, populations that are evolved using variations of the lexicase selection operator show a very high degree of genetic diversity which might be a key contributer to the improved performance [@6920034, p.1] [@epsilon_lexicase_main, p.745].
-In theory, Lexicase based selection should be more prone to select solutions that are specialist (high fitness on a small subset of training cases) than solutions that are generalist (high average fitness among all training cases).
+Both articles argue that many of the problem domains that GP is commonly used for are prevalent for problems that are modal/ uncompromising in nature.
+In comparison to other selection operators, populations that are evolved using variations of the lexicase selection operator show a very high degree of genetic diversity which might be a key contributor to the improved performance [@6920034, p.1] [@epsilon_lexicase_main, p.745].
+In theory, Lexicase based selection should be more prone to select solutions that are specialist (solutions that achieve high fitness on a smaller subset of unusual training cases) than solutions that are generalist (solutions that achieve high average fitness among all training cases).
 
-The first description^[@lexicase_first_desciption, p.4] of the lexicase parent selection algorithm by @lexicase_first_desciption is quoted below for reference:
+The first description^[@lexicase_first_desciption, p.4] of the lexicase parent selection algorithm by @lexicase_first_desciption is quoted verbatim below for reference:
 
@@
 
-More recent research by @epsilon_lexicase_main suggested $\epsilon$-Lexicase selection as a modified selection operator that can improve overall GP performance if applied to continous-valued symbolic regression tasks in comparison to tournament and standard lexicase selection.
-@epsilon_lexicase_main argued that the original concept behind lexicase selection as formulated by @lexicase_first_desciption does not fit the requirements of real-world symbolic regression tasks.
-The authors identified the pass condition of regular lexicase selection as the main problem if applied to symbolic regression: Individual solutions can only be selected if they perform on an elite level but with noisey and continous-valued data it is very unlikely for two individuals to achieve an exactly equal error on a test case, this finally results in filtering out too many individuals during the selection process[@epsilon_lexicase_main, p.742].
+More recent research by @epsilon_lexicase_main suggests $\epsilon$-Lexicase selection as a modified selection operator that can improve overall GP performance if applied to continuous-valued symbolic regression tasks in comparison to tournament and standard lexicase selection.
+@epsilon_lexicase_main argues that the original concept behind lexicase selection as formulated by @lexicase_first_desciption does not fit the requirements of real-world symbolic regression tasks.
+The authors identifies the pass condition of regular lexicase selection as the main problem if applied to symbolic regression: Individual solutions can only be selected if they perform on an elite level but with continuous-valued, and often noisy, data it is very unlikely for two individuals to achieve an exactly equal error for a given test case.
+This stringent pass condition can result in filtering out too many individuals during the selection process [@epsilon_lexicase_main, p.742].
 $\epsilon$-Lexicase adresses this problem by introducing an $\epsilon$ parameter that specifies a range around the elite error, individuals that perform inside this range pass the current selection iteration.
-@https://doi.org/10.48550/arxiv.1709.05394 describe $\epsilon$-Lexicase selection as a more "relaxed version of lexicase selection".
+@https://doi.org/10.48550/arxiv.1709.05394 describes $\epsilon$-Lexicase selection as a more "relaxed version of lexicase selection".
 Different methods to configure the $\epsilon$ Parameter have been explored, for the limited scope of this project I will focus on the most promising implementation of $\epsilon$-Lexicase selection that is automatically adjusted on the basis of the median absolute deviation of errors inside the selection pool [@epsilon_lexicase_main, p.742] [@https://doi.org/10.48550/arxiv.1709.05394, p.6].
@@
 
-In comparison, traditional GP selection methods such as tournament selection most commonly compute the fitness of a program as the mean of its error for each individual fitness case.
+In comparison, a traditional GP selection method like tournament selection most commonly computes the fitness of a program as the mean of its error for each individual fitness case.
 One downside to this approach is that the total amount of information is reduced from a wide range of individual errors to a single metric.
-@6920034 suspected that this loss in overall information provided to GP might reduce overall performance especially if applied to the class of uncompromising problems that require the solution to perform well on a wide range of diverse cases.
+@6920034 suspects that this loss in overall information provided to GP might reduce overall performance especially if applied to the class of uncompromising problems that require the solution to perform well on a wide range of diverse cases.
 Since tournament selection selects individuals based on their average fitness across all test cases, the resulting solutions should be expected to be more biased towards being generalists instead of specialists.
@@
 Generalization, the ability of a model to perform well on previously unseen cases, is one of the fundamental goals in most real world machine learning applications.
-@open_issues_gp raised awareness to the fact, that the topic of generalization in GP has not gotten the attention that other machine learning-based areas, e.g, deep learning, have attributed to it.
-A similiar statement was proposed by @generalisation_in_gp in his review of research in generalization in GP.
-Both authors adressed the need for additional research regarding the topic.
+@open_issues_gp raises awareness to the fact, that the topic of generalization in GP has not gotten the attention that other machine learning-based areas  have attributed to it.
+A similar statement has been proposed by @generalisation_in_gp in his review of research in generalization in GP.
+Both authors address the need for additional research regarding the topic.
 
-@generalisation_in_gp critized that almost all applications published in the initial GP literature by John Koza [@koza_main] are not using separate training and testing datasets which might result in overfitting and a poor overall generalization ability of the programs that are produced.
-The author calls for a more widespread adoption of methods like generational sampling of new training cases or the overall separation into training and testing datasets to improve generalization in GP [@generalisation_in_gp, p.10].
+@generalisation_in_gp criticizes that almost all applications published in the initial GP literature by @koza_main are not using separate training and testing data sets which might result in over fitting and a poor overall generalization ability of the programs that are produced.
+The author calls for a more widespread adoption of methods like generational sampling of new training cases or the overall separation into training and testing data sets to improve generalization in GP [@generalisation_in_gp, p.10].
 
@@
 268].
-The minimum description length principle (MDLP) describes this phenomenom by claiming that less complex models are more likely to perform better at generalizing than more complex models that achieve a comparable fitness during the training phase [@open_issues_gp, p.
+The minimum description length principle (MDLP) describes this phenomenon by claiming that less complex models are more likely to perform better at generalizing than more complex models that achieve a comparable fitness during the training phase [@open_issues_gp, p.
 349].
-However MDLP should be considered carefully, experiments by @bloat_overfitting_gp on the prediction of bioavailability for drugs demonstrated that GP based techniques can either bloat without overfitting or overfit without bloating [@bloat_overfitting_gp, p.
+However MDLP should be considered carefully, experiments by @bloat_overfitting_gp on the prediction of bio availability for drugs demonstrate that GP based techniques can either bloat without over fitting or over fit without bloating [@bloat_overfitting_gp, p.
 8].
@@
 
-The task of finding a mathematical function that fits a given set of datapoints has been one of the first applications of GP in [@koza_main, p. 238].
+The task of finding a mathematical function that fits a given set of data points has been one of the first applications of GP in [@koza_main, p. 238].
 Practical examples given by @koza_main included tasks such as the discovery of scientific laws, finding solutions to differential and integral equations or the programmatic compression of images.
 
-Since it's inception, GP based symbolic regression has been widely used in practical application ranging over a wide field of disciplines.
+Since it's inception, GP based symbolic regression has been widely used in practical applications ranging over a wide field of disciplines.
 A well known example of an industrial use-case has been published by @soft_sensor_gp.
-The authors used GP based symbolic regression to derive nonlinear functions that could be used to significantly increase the robustness of industrial sensors which resulted in wide-spread adoption and a significant reduction in costs.
+The authors use GP based symbolic regression to derive nonlinear functions that can be used to significantly increase the robustness of industrial sensors which resulted in wide-spread adoption and a significant reduction in costs.
 
-Many variations of GP based symbolic regression have been proposed to improve overall performance, @wang_wagner_rondinelli_2019 summarized the main characteristics and differences for 6 common variations including procedures such as geometric semantic genetic programming, cartesian genetic programming or GP-based relevance vector machines.
-A large benchmarking study on symbolic regression published by @Orzechowski_2018 compared 4 GP-based procedures to several state-of-the-art machine leaning techniques.
-The authors find that GP, eventhough more time consuming, can achieve better results if compared to the state-of-the-art machine learning algorithm gradient-boosting.
+Many variations of GP based symbolic regression have been proposed to improve overall performance, @wang_wagner_rondinelli_2019 summarizes the main characteristics and differences for 6 common variations including procedures such as geometric semantic genetic programming, Cartesian genetic programming or GP-based relevance vector machines.
+A large benchmarking study on symbolic regression published by @Orzechowski_2018 compares 4 GP-based procedures to several state-of-the-art machine leaning techniques.
+The authors demonstrate that GP, even though more time consuming, can achieve better results if compared to the state-of-the-art machine learning algorithm gradient-boosting.
 
@@
 
-<!-- Dataset and Task --->
-To study the influence of selection on generalization I selected a dataset about energy efficiency in buildings that is part of the UC Irvine Machine Learning Repository [@Dua:2019].
-The dataset contains eight individual building attributes that map to two different outcomes, heating and cooling load, for N=768 cases [@fe8fa39e88a040bbacba5a465c48043f].
+<!-- data set and Task --->
+To study the influence of selection on generalization I selected a data set about energy efficiency in buildings that is part of the UC Irvine Machine Learning Repository [@Dua:2019].
+The data set contains eight individual building attributes that map to two different outcomes, heating and cooling load, for N=768 cases [@fe8fa39e88a040bbacba5a465c48043f].
 
-Using two otherwise identical GP systems, one deploying tournament selection and the other $\epsilon$-lexicase selection, the objective is to find a computer program that best predicts the outcome variable heating load $(Y1)$^[The symbolic regression is performed on one of the two provided outcome variables, the variable Cooling Load will be excluded.] of the buildings using a subset of the eight building attributes $(X1,..,X8)$ for input.
-The specific meaning of all attributes inside the dataset are described in table 1.
+Using two otherwise identical GP systems, one deploying tournament selection and the other $\epsilon$-lexicase selection, the objective is to find a computer program that best predicts the outcome variable heating load $(Y1)$^[The symbolic regression is performed on one of the two provided outcome variables, the variable Cooling Load will be excluded.] of the buildings using a subset of the eight building attributes $(X1,..,X8)$ for it's input.
+The specific meaning of all attributes inside the data set are described in table 1.
 
@@
   digits=3,
-  caption="Overview - Energy Heating Dataset"
+  caption="Overview - Energy Heating data set"
 )    %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
@@
 <!-- Method:single run --->
-To measure the generalization ability of each model the dataset will be randomly split in half, resulting in a training and testing dataset each containing 384 individual cases.
-Each model will be evolved by traditional GP using only the fitness cases present in the training dataset.
-For each generation the highest fitness model of the current population will be tested with the previously unseen fitness cases that are part of the testing dataset.
+To measure the generalization ability of each model the data set will be randomly split in half, resulting in a training and testing data set each containing 384 individual cases.
+Each model will be evolved by traditional GP using only the fitness cases present in the training data set.
+For each generation the highest fitness model of the current population will be tested with the previously unseen fitness cases that are part of the testing data set.
 For each run of the experiment statistics will be collected on fitness which will form the basis of my further statistical analysis.
@@
 Since GP is a stochastic optimization algorithm, the basic experiment will be run for a total of 50 times to ensure a fair and meaningful comparison based on a large number of runs for both algorithms.
-For each run of the experiment the dataset will be randomly split in half as described, both models are then trained and tested using the exact same set of fitness cases.
+For each run of the experiment the data set will be randomly split in half as described, both models are then trained and tested using the exact same set of fitness cases.
 
@@
 
-To further examine the difference in generalization behaviour, the mean testing and training errors over each generation will be visualized for both algorithms.
-The specific aim of this visualization is to explore the mechanism of overfitting and to examine if differences between both methods can be detected.
+To further examine the difference in generalization behavior, the mean testing and training errors over each generation will be visualized for both algorithms.
+The specific aim of this visualization is to explore the mechanism of over fitting and to examine if differences between both methods can be detected.
 
@@
 
-The resulting fitness function $f$ for an individual program $i$ is descibed by eq 2:
+The resulting fitness function $f$ for an individual program $i$ is described by eq 2:
 
@@
 
-The primitve set consists of the terminals that are listed in table 3 and the functions that are listed in table 4.
-To avoid runtime errors I implemented protected version of the operators for division, natural logarithm and square root [@koza_main, p.82-83].
-As suggested by @koza_main I also included ephemeral constants to the set of terminals to provide the evolutionay search the opportunity to explore and include randomly generated constants.
+The primitive set consists of the terminals that are listed in table 3 and the functions that are listed in table 4.
+To avoid run time errors I implemented protected version of the operators for division, natural logarithm and square root [@koza_main, p.82-83].
+As suggested by @koza_main I also included ephemeral constants to the set of terminals to provide the evolutionary search the opportunity to explore and include randomly generated constants.
 
@@
 GP operators are represented in table 5.
-Both genetic operators use a static limit to control for the height of the reulting trees [@koza_main, p.104].
+Both genetic operators use a static limit to control for the height of the resulting trees [@koza_main, p.104].
 Individual programs are initialized by using the ramped half-and-half method, 50% of the population are created by using the Growth algorithm and the remaining 50% are created by using the Full algorithm [@koza_main, p.93].
@@
 
-The crossover operator implemented by DEAP randomly selects a crossover point in each individual and exchanges each subtree with the point as root between each individual [@DEAP_JMLR2012].
-Mutation also randomly selects a point in the tree individual, it then replaces the subtree below that point as a root by the expression generated using the full grow initialization method [@DEAP_JMLR2012].
+The crossover operator implemented by DEAP randomly selects a crossover point in each individual and exchanges each sub tree with the point as root between each individual [@DEAP_JMLR2012].
+Mutation also randomly selects a point in the tree individual, it then replaces the sub tree below that point as a root by the expression generated using the full grow initialization method [@DEAP_JMLR2012].
 
@@
 Some interesting observation from tables 6 and 7 can be made by comparing the mean of the variables $min\_training\_error$ and $testing\_error$.
-Variable $min\_training\_error$ represents the fitness, measured by the MSE, of the best performing solution that has been found during a single run of the experiment while the variable $testing\_error$ represents the fitness of the same solution if evaluated for the unknown fitness cases inside the testing dataset.
+Variable $min\_training\_error$ represents the fitness, measured by the MSE, of the best performing solution that has been found during a single run of the experiment while the variable $testing\_error$ represents the fitness of the same solution if evaluated for the unknown fitness cases inside the testing data set.
 Tournament selection based GP achieves a mean $min\_training\_error$ of $16.483$ and a mean $testing\_error$ of $16.905$ while the GP system using $\epsilon$-Lexicase selection can only achieve a mean $min\_training\_error$ of $22.013$ and a mean $testing\_error$ of $22.456$.
-These results indicate an overall performance benefit for torunament selection based GP in the experiment.
+These results indicate an overall performance benefit for tournament selection based GP in the experiment.
 It should also be mentioned that the best performing model over all 50 runs, measured by the minimum of variables $min\_training\_error$ and $testing\_error$, has been achieved by tournament selection with an MSE of 7.066 for the training data and	7.847 for the testing data.
@@
 
-Figure 1 visualizes the distribution of $min\_training\_error$ and $testing\_error$ for both selection operators over the total 50 runs using a boxplot and summerizes the initial observations from tables 6 and 7 described above.
+Figure 1 visualizes the distribution of $min\_training\_error$ and $testing\_error$ for both selection operators over the total 50 runs using a box plot and summarizes the initial observations from tables 6 and 7 described above.
 
@@
 The results are summarized in table 8.
-Since not all samples test positive for a normal distribution a Mann-Whitney U (MWU) ranksum test will be performed to test for statistical significance between the four variables $tournament\_min\_training\_error$, $tournament\_testing\_error$, $e-lexicase\_min\_training\_error$ and $e-lexicase\_testing\_error$.
+Since not all samples test positive for a normal distribution a Mann-Whitney U (MWU) rank sum test will be performed to test for statistical significance between the four variables $tournament\_min\_training\_error$, $tournament\_testing\_error$, $e-lexicase\_min\_training\_error$ and $e-lexicase\_testing\_error$.
 
@@
 The differences in mean fitness between tournament selection and $\epsilon$-Lexicase selection are highly statistical significant even if considering an $\alpha$ of 0.01.
-For the given symbolic regression task tournament selection achieves a better mean performance on both the training as well as the testing dataset than $\epsilon$-Lexicase selection.
-These findings contradict the hypothesis $H0_{1}$ descibed in subsection 3.1 and are contrary to some of the research that was summarized in subsections 2.1 and 2.3.[@epsilon_lexicase_main] [@https://doi.org/10.48550/arxiv.1709.05394].
+For the given symbolic regression task tournament selection achieves a better mean performance on both the training as well as the testing data set than $\epsilon$-Lexicase selection.
+These findings contradict the hypothesis $H0_{1}$ described in subsection 3.1 and are contrary to some of the research that was summarized in subsections 2.1 and 2.3.[@epsilon_lexicase_main] [@https://doi.org/10.48550/arxiv.1709.05394].
 
@@
 
-To further examine the generalization behaviour of both selection operators I plotted the mean fitness scores for both algorithms over the total number of 100 generations in figure 2.
-Figure 2 shows that both algorithms show little visual evidence of overfitting which would result in a growing gap between between $min\_training\_error$ and $testing\_error$ in the final stages of the evolutionary optimization.
+To further examine the generalization behavior of both selection operators I plotted the mean fitness scores for both algorithms over the total number of 100 generations in figure 2.
+Figure 2 shows that both algorithms show little visual evidence of over fitting which would result in a growing gap between between $min\_training\_error$ and $testing\_error$ in the final stages of the evolutionary optimization.
 For $\epsilon$-Lexicase selection a small gap can be observed starting at around the 90.th generation, while the minimum training error slowly converges a small increase in the testing error can be observed.
-Based on the limited amount of collected data, it is questionable if this small sign of overfitting can be interpreted as a weak indicator for an inferior generalization behaviour of $\epsilon$-Lexicase in comparison to tournament selection.
+Based on the limited amount of collected data, it is questionable if this small sign of over fitting can be interpreted as a weak indicator for an inferior generalization behavior of $\epsilon$-Lexicase in comparison to tournament selection.
 
-Figure 2 also shows a visual representation of the described overall performance benefit of using tournament selection instead of $\epsilon$-Lexicase selection for the choosen symbolic regression task.
-It can be observed that at all stages of the evolution tournament selection based GP achieves a lower mean error for both datasets.
+Figure 2 also shows a visual representation of the described overall performance benefit of using tournament selection instead of $\epsilon$-Lexicase selection for the chosen symbolic regression task.
+It can be observed that at all stages of the evolution tournament selection based GP achieves a lower mean error for both data sets.
 
@@
 
-In a final step I analyzed the growth behaviour for both GP systems.
+In a final step I analyzed the growth behavior for both GP systems.
 Figure 3 visualizes the average size of individuals inside the population as well as the average size of the best performing individual for each generation of the evolution.
@@
   1. Configuration of evolutionary parameters
-  2. Limited computational ressources
+  2. Limited computational resources
 
 The combination of both issues might have resulted in a significant reduction in overall robustness of the results represented in sections 5 and 6.
-The performance of all evolutionary algorithms, including GP, can be highly dependant on a large number of parameters and implementation details.
+The performance of all evolutionary algorithms, including GP, can be highly dependent on a large number of parameters and implementation details.
 Although the series of experiments conducted in this project are comparably simple in nature, the results achieved might still be highly influenced by the GP configuration detailed in subsection 4.2.
 
-To gain further insight into the differences in generalizaion behaviour and to adress the research question with a higher level of certainty, the experiment I conducted should be repeated for many different GP configurations.
+To gain further insight into the differences in generalization behavior and to adress the research question with a higher level of certainty, the experiment I conducted should be repeated for many different GP configurations.
 Some exemplary parameters that could drastically influence the results of this experiment include the population size, the total number of generations, the inclusion of elitism or the computation of $\epsilon$ in lexicase selection.
-Other factors whose influence on generalization behaviour could be important include different methods to compute an individuals fitness (e.g.
+Other factors whose influence on generalization behavior could be important include different methods to compute an individuals fitness (e.g.
 mean absolute error or root mean squared error) and testing out different variations of the genetic operators for mutation and crossover.
 
-Another obvious weakness of my project is that the experiment is only based on a single symbolic regression application, the prediction of the heating load of buildings based on a relativley small dataset.
-To gain further trust in the obtained results, the experiment should be repeated for other symbolic regression tasks and datasets.
-At the current state it remains unknown if the generalization behaviour of both selection operators might be highly problem specific and dependant on the dataset used for training and testing.
+Another obvious weakness of my project is that the experiment is only based on a single symbolic regression application, the prediction of the heating load of buildings based on a relatively small data set.
+To gain further trust in the obtained results, the experiment should be repeated for other symbolic regression tasks and data sets.
+At the current state it remains unknown if the generalization behavior of both selection operators might be highly problem specific and dependent on the data set used for training and testing.
 
-All limitations and shortcomings described above are also connected to the second limitation of this research project, the limited amount of computational ressources.
+All limitations and shortcomings described above are also connected to the second limitation of this research project, the limited amount of computational resources.
 The total computation time consumed to run the current experiments (50 runs, 2 algorithms, 100 generations, 500 individuals) has already been close to 24 hours on a modern, arm-based Apple-M1 system.
-Eventhough the overall computation time could probably be reduced by further optimization of the source code, e.g.
-using tools for parallel computation, the ressources necessary to repeat the whole experiment for different parameter configurations would certainly be beyond the scope of this research project.
+Even though the overall computation time could probably be reduced by further optimization of the source code, e.g.
+using tools for parallel computation, the resources necessary to repeat the whole experiment for different parameter configurations would certainly be beyond the scope of this research project.
 
@@
 :::
-\newpage
 
 # II  Statutory Declaration {.unnumbered #II}
+
+![Statutory Declaration](./misc/erklärung.pdf)\

### 2022-06-22 docs/rmd/paper.Rmd
rendered doc

@@
       in_header: settings.tex
+# indent: true
 fontsize: 12pt
@@
 # include-before: '`\newpage{} `{=latex}'
-include-before: '`\thispagestyle{empty} `{=latex}'
+include-before: '`\thispagestyle{empty} \newpage{}`{=latex}'
 ---
@@
 
-The first description^[@lexicase_first_desciption, p.4] of the lexicase parent selection algorithm by @lexicase_first_desciption is quoted verbatim below for reference:
+The first description of the lexicase parent selection algorithm is quoted verbatim below for reference [@lexicase_first_desciption, p.4]:
 
@@
 
-> $H0_{3}$: Size differences exist between the distribution underlying the samples produced by $\epsilon$-Lexicase and the distribution underlying samples of tournament selection
+> $H0_{3}$: No differences in program size exist between the distribution underlying the samples produced by $\epsilon$-Lexicase and the distribution underlying samples of tournament selection
 
@@
 
-In a final step I analyzed the growth behavior for both GP systems.
-Figure 3 visualizes the average size of individuals inside the population as well as the average size of the best performing individual for each generation of the evolution.
+In a final step I analyze the growth behavior for both GP systems to determine if differences in program size can be observed based on selection operators.
+Figure 3 visualizes the average size of individuals inside the population as well as the average size of the best performing individual for each generation and both GP systems.
+![Mean Size](./plots/size_subplotted.png)
 
-![Mean Size](./plots/size_combined.png)
-... interpretation...
+Both subplots display a very similar behavior that is common for GP, we observe a strong trend for solutions to grow in size with each additional generation that is concluded.
+Another observation  that can be made is that the best solutions of each generation almost always have a larger size than the population average.
+On first impression, the differences between both selection operators appears to be insignificant.
+Again, I conducted a MWU test to test for statistical differences in the underlying distribution of sizes measured during the experiment.
 
-Again, a MWU test is conducted to test for statistical differences in the underlying distribution of the sizes measured during the experiment. The results are summarized in table 10.
+The resulting p-values of the MWU test are summarized in table 10.
 
@@
 
+None of the differences in mean size are statistically significant.
+Again, based on the results obtained in my experiment, the evidence produced fails to reject the null hypothesis $H0_{3}$ that seletion has no significant effect on differences in program size.
+
 # Conclusion
@@
 
-To gain further insight into the differences in generalization behavior and to adress the research question with a higher level of certainty, the experiment I conducted should be repeated for many different GP configurations.
+To gain further insight into the differences in generalization behavior and to adress the research question with a higher level of certainty, the experiment I conducted should be repeated for different GP configurations.
 Some exemplary parameters that could drastically influence the results of this experiment include the population size, the total number of generations, the inclusion of elitism or the computation of $\epsilon$ in lexicase selection.
 Other factors whose influence on generalization behavior could be important include different methods to compute an individuals fitness (e.g.
-mean absolute error or root mean squared error) and testing out different variations of the genetic operators for mutation and crossover.
+mean absolute error or root mean squared error), testing different variations of the genetic operators for mutation and crossover and experimenting with different variations of the training:testing split size.
 
@@
 ![Statutory Declaration](./misc/erklärung.pdf)\
+
+```{zsh github, include=FALSE}
+git add * && git commit --m "rendered doc" && git push
+```

### 2022-06-22 docs/rmd/paper.tex
autogit
rendered doc

@@
 \author{}
-\date{\vspace{-2.5em}Date of Submission: 2022-06-22}
+\date{\vspace{-2.5em}Date of Submission: 2022-06-23}
 
@@
 
-\thispagestyle{empty}
+\thispagestyle{empty} \newpage{}
 
@@
 
-Genetic programming (GP), a subfield of evolutionary computation (EC),
-is a metaheuristic that is used to search for computer programs that
-solve a given problem by simulating the process of darwinian evolution.
+Genetic programming (GP), a sub field of evolutionary computation (EC),
+is a meta heuristic that is used to search for computer programs that
+solve a given problem by simulating the process of Darwinian evolution.
 The basic principle of GP is to gradually evolve solutions by repeatedly
@@
 programs based on a fitness metric. Then, genetic operators are applied
-on the selected solutions to genereate new offspring candidate
-solutions. By repeating this process over many generations, GP acts as a
-guided search for high fitness solutions throughout the decision space.
-A unique feature of GP among other evolutionary optimization procedures
-is the possibility to evolve solutions of variable length.
+on the selected solutions to generate new offspring candidate solutions.
+By repeating this process over many generations, GP acts as a guided
+search for high fitness solutions throughout the decision space. A
+unique feature of GP among other evolutionary optimization procedures is
+the possibility to evolve solutions of variable length.
 
@@
 
-Since regular Lexicase selection has been shown to perform suboptimal on
-continous-valued optimization problems, a modified variation called
+Since regular Lexicase selection has been shown to perform sub optimal
+on continuous-valued optimization problems, a modified variation called
 \(\epsilon\)-Lexicase selection has been suggested for symbolic
@@
 The goal of symbolic regression is to find a mathematical model for an
-observed set of datapoints (Paris, Robilliard and Fonlupt, 2004, p.
+observed set of data points (Paris, Robilliard and Fonlupt, 2004, p.
 794). Symbolic Regression has been one of the first GP applications and
@@
 (Poli, Langdon and McPhee, 2008, p. 114). In most symbolic regression
-problems, little to no a priori knowledge about the optimal form and
+problems, little to no prior knowledge about the optimal form and
 structure of the target function is available. The ability of GP to
@@
 test cases a model is trained on but to also perform well on previously
-unseen cases, this is refered to as generalization. In most real world
+unseen cases, this is referred to as generalization. In most real world
 applications of symbolic regression only a small subset of labeled data
@@
 previously unseen cases with high precision (Gonçalves, 2016, p. 6). A
-model that is extensivley optimized on the provided training data, may
-overfit to this data sample which may lead to a decrease in
+model that is extensively optimized on the provided training data, can
+become over fitted to this data sample which may lead to a decrease in
 generalization.
@@
 This research project tries to answer the question if the usage of
-\(\epsilon\)-Lexicase selection influences the generalization behaviour
+\(\epsilon\)-Lexicase selection influences the generalization behavior
 of programs that are evolved using GP for symbolic regression if
@@
 Spector (2012) it was suggested as a novel parent selection method for
-target problems that are modal in nature. The author classified modal
+target problems that are modal in nature. The author classifies modal
 problems as ``problems that qualitatively different modes of response
@@
 for the purpose of solving so called uncompromising problems with GP.
-Helmuth, Spector and Matheson (2015) defined uncompromising problems as
+Helmuth, Spector and Matheson (2015) defines uncompromising problems as
 problems that require the final solution to perform optimal on each of
@@
 design of digital multipliers or finding terms in finite algebras. The
-authors also provided evidence that lexicase selection can significantly
+authors also provide evidence that lexicase selection can significantly
 improves GP's ability to solve some uncompromising problems if compared
@@
 
-Both articles argued that many of the problem domains that GP is
-commonly used for are prevalent for problems that are modal/
-uncompromising in nature. In comparison to other selection operators,
-populations that are evolved using variations of the lexicase selection
-operator show a very high degree of genetic diversity which might be a
-key contributer to the improved performance (Helmuth, Spector and
-Matheson, 2015, p. 1) (La Cava, Spector and Danai, 2016, p. 745). In
-theory, Lexicase based selection should be more prone to select
-solutions that are specialist (high fitness on a small subset of
-training cases) than solutions that are generalist (high average fitness
-among all training cases).
-
-The first description\footnote{Spector (2012), p.4} of the lexicase
-parent selection algorithm by Spector (2012) is quoted below for
-reference:
+Both articles argue that many of the problem domains that GP is commonly
+used for are prevalent for problems that are modal/ uncompromising in
+nature. In comparison to other selection operators, populations that are
+evolved using variations of the lexicase selection operator show a very
+high degree of genetic diversity which might be a key contributor to the
+improved performance (Helmuth, Spector and Matheson, 2015, p. 1) (La
+Cava, Spector and Danai, 2016, p. 745). In theory, Lexicase based
+selection should be more prone to select solutions that are specialist
+(solutions that achieve high fitness on a smaller subset of unusual
+training cases) than solutions that are generalist (solutions that
+achieve high average fitness among all training cases).
+
+The first description of the lexicase parent selection algorithm is
+quoted verbatim below for reference (Spector, 2012, p. 4):
 
@@
 
-More recent research by La Cava, Spector and Danai (2016) suggested
+More recent research by La Cava, Spector and Danai (2016) suggests
 \(\epsilon\)-Lexicase selection as a modified selection operator that
-can improve overall GP performance if applied to continous-valued
+can improve overall GP performance if applied to continuous-valued
 symbolic regression tasks in comparison to tournament and standard
-lexicase selection. La Cava, Spector and Danai (2016) argued that the
+lexicase selection. La Cava, Spector and Danai (2016) argues that the
 original concept behind lexicase selection as formulated by Spector
 (2012) does not fit the requirements of real-world symbolic regression
-tasks. The authors identified the pass condition of regular lexicase
+tasks. The authors identifies the pass condition of regular lexicase
 selection as the main problem if applied to symbolic regression:
 Individual solutions can only be selected if they perform on an elite
-level but with noisey and continous-valued data it is very unlikely for
-two individuals to achieve an exactly equal error on a test case, this
-finally results in filtering out too many individuals during the
-selection process(La Cava, Spector and Danai, 2016, p. 742).
-\(\epsilon\)-Lexicase adresses this problem by introducing an
-\(\epsilon\) parameter that specifies a range around the elite error,
-individuals that perform inside this range pass the current selection
-iteration. La Cava \emph{et al.} (2017) describe \(\epsilon\)-Lexicase
-selection as a more ``relaxed version of lexicase selection''. Different
-methods to configure the \(\epsilon\) Parameter have been explored, for
-the limited scope of this project I will focus on the most promising
-implementation of \(\epsilon\)-Lexicase selection that is automatically
-adjusted on the basis of the median absolute deviation of errors inside
-the selection pool (La Cava, Spector and Danai, 2016, p. 742) (La Cava
-\emph{et al.}, 2017, p. 6).
+level but with continuous-valued, and often noisy, data it is very
+unlikely for two individuals to achieve an exactly equal error for a
+given test case. This stringent pass condition can result in filtering
+out too many individuals during the selection process (La Cava, Spector
+and Danai, 2016, p. 742). \(\epsilon\)-Lexicase adresses this problem by
+introducing an \(\epsilon\) parameter that specifies a range around the
+elite error, individuals that perform inside this range pass the current
+selection iteration. La Cava \emph{et al.} (2017) describes
+\(\epsilon\)-Lexicase selection as a more ``relaxed version of lexicase
+selection''. Different methods to configure the \(\epsilon\) Parameter
+have been explored, for the limited scope of this project I will focus
+on the most promising implementation of \(\epsilon\)-Lexicase selection
+that is automatically adjusted on the basis of the median absolute
+deviation of errors inside the selection pool (La Cava, Spector and
+Danai, 2016, p. 742) (La Cava \emph{et al.}, 2017, p. 6).
 
@@
 
-In comparison, traditional GP selection methods such as tournament
-selection most commonly compute the fitness of a program as the mean of
+In comparison, a traditional GP selection method like tournament
+selection most commonly computes the fitness of a program as the mean of
 its error for each individual fitness case. One downside to this
@@
 range of individual errors to a single metric. Helmuth, Spector and
-Matheson (2015) suspected that this loss in overall information provided
+Matheson (2015) suspects that this loss in overall information provided
 to GP might reduce overall performance especially if applied to the
@@
 unseen cases, is one of the fundamental goals in most real world machine
-learning applications. O'Neill \emph{et al.} (2010) raised awareness to
+learning applications. O'Neill \emph{et al.} (2010) raises awareness to
 the fact, that the topic of generalization in GP has not gotten the
-attention that other machine learning-based areas, e.g, deep learning,
-have attributed to it. A similiar statement was proposed by Kushchu
-(2002) in his review of research in generalization in GP. Both authors
-adressed the need for additional research regarding the topic.
-
-Kushchu (2002) critized that almost all applications published in the
-initial GP literature by John Koza (Koza, 1992) are not using separate
-training and testing datasets which might result in overfitting and a
-poor overall generalization ability of the programs that are produced.
-The author calls for a more widespread adoption of methods like
-generational sampling of new training cases or the overall separation
-into training and testing datasets to improve generalization in GP
-(Kushchu, 2002, p. 10).
+attention that other machine learning-based areas have attributed to it.
+A similar statement has been proposed by Kushchu (2002) in his review of
+research in generalization in GP. Both authors address the need for
+additional research regarding the topic.
+
+Kushchu (2002) criticizes that almost all applications published in the
+initial GP literature by Koza (1992) are not using separate training and
+testing data sets which might result in over fitting and a poor overall
+generalization ability of the programs that are produced. The author
+calls for a more widespread adoption of methods like generational
+sampling of new training cases or the overall separation into training
+and testing data sets to improve generalization in GP (Kushchu, 2002, p.
+10).
 
@@
 2019, p. 268). The minimum description length principle (MDLP) describes
-this phenomenom by claiming that less complex models are more likely to
+this phenomenon by claiming that less complex models are more likely to
 perform better at generalizing than more complex models that achieve a
@@
 2010, p. 349). However MDLP should be considered carefully, experiments
-by Silva and Vanneschi (2009) on the prediction of bioavailability for
-drugs demonstrated that GP based techniques can either bloat without
-overfitting or overfit without bloating (Silva and Vanneschi, 2009, p.
-8).
+by Silva and Vanneschi (2009) on the prediction of bio availability for
+drugs demonstrate that GP based techniques can either bloat without over
+fitting or over fit without bloating (Silva and Vanneschi, 2009, p. 8).
 
@@
 The task of finding a mathematical function that fits a given set of
-datapoints has been one of the first applications of GP in (Koza, 1992,
+data points has been one of the first applications of GP in (Koza, 1992,
 p. 238). Practical examples given by Koza (1992) included tasks such as
@@
 Since it's inception, GP based symbolic regression has been widely used
-in practical application ranging over a wide field of disciplines. A
+in practical applications ranging over a wide field of disciplines. A
 well known example of an industrial use-case has been published by
-Jordaan \emph{et al.} (2004). The authors used GP based symbolic
-regression to derive nonlinear functions that could be used to
+Jordaan \emph{et al.} (2004). The authors use GP based symbolic
+regression to derive nonlinear functions that can be used to
 significantly increase the robustness of industrial sensors which
@@
 improve overall performance, Wang, Wagner and Rondinelli (2019)
-summarized the main characteristics and differences for 6 common
+summarizes the main characteristics and differences for 6 common
 variations including procedures such as geometric semantic genetic
-programming, cartesian genetic programming or GP-based relevance vector
+programming, Cartesian genetic programming or GP-based relevance vector
 machines. A large benchmarking study on symbolic regression published by
-Orzechowski, Cava and Moore (2018) compared 4 GP-based procedures to
-several state-of-the-art machine leaning techniques. The authors find
-that GP, eventhough more time consuming, can achieve better results if
-compared to the state-of-the-art machine learning algorithm
+Orzechowski, Cava and Moore (2018) compares 4 GP-based procedures to
+several state-of-the-art machine leaning techniques. The authors
+demonstrate that GP, even though more time consuming, can achieve better
+results if compared to the state-of-the-art machine learning algorithm
 gradient-boosting.
@@
 
-To study the influence of selection on generalization I selected a
-dataset about energy efficiency in buildings that is part of the UC
-Irvine Machine Learning Repository (Dua and Graff, 2017). The dataset
-contains eight individual building attributes that map to two different
-outcomes, heating and cooling load, for N=768 cases (Tsanas and Xifara,
-2012).
+To study the influence of selection on generalization I selected a data
+set about energy efficiency in buildings that is part of the UC Irvine
+Machine Learning Repository (Dua and Graff, 2017). The data set contains
+eight individual building attributes that map to two different outcomes,
+heating and cooling load, for N=768 cases (Tsanas and Xifara, 2012).
 
@@
   will be excluded.} of the buildings using a subset of the eight
-building attributes \((X1,..,X8)\) for input. The specific meaning of
-all attributes inside the dataset are described in table 1.
+building attributes \((X1,..,X8)\) for it's input. The specific meaning
+of all attributes inside the data set are described in table 1.
 
@@
 
-\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating Dataset}
+\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating data set}
 \centering
@@
 
-To measure the generalization ability of each model the dataset will be
-randomly split in half, resulting in a training and testing dataset each
-containing 384 individual cases. Each model will be evolved by
-traditional GP using only the fitness cases present in the training
-dataset. For each generation the highest fitness model of the current
+To measure the generalization ability of each model the data set will be
+randomly split in half, resulting in a training and testing data set
+each containing 384 individual cases. Each model will be evolved by
+traditional GP using only the fitness cases present in the training data
+set. For each generation the highest fitness model of the current
 population will be tested with the previously unseen fitness cases that
-are part of the testing dataset. For each run of the experiment
+are part of the testing data set. For each run of the experiment
 statistics will be collected on fitness which will form the basis of my
@@
 comparison based on a large number of runs for both algorithms. For each
-run of the experiment the dataset will be randomly split in half as
+run of the experiment the data set will be randomly split in half as
 described, both models are then trained and tested using the exact same
@@
 \begin{quote}
-\(H0_{3}\): Size differences exist between the distribution underlying
-the samples produced by \(\epsilon\)-Lexicase and the distribution
-underlying samples of tournament selection
+\(H0_{3}\): No differences in program size exist between the
+distribution underlying the samples produced by \(\epsilon\)-Lexicase
+and the distribution underlying samples of tournament selection
 \end{quote}
@@
 
-To further examine the difference in generalization behaviour, the mean
+To further examine the difference in generalization behavior, the mean
 testing and training errors over each generation will be visualized for
 both algorithms. The specific aim of this visualization is to explore
-the mechanism of overfitting and to examine if differences between both
+the mechanism of over fitting and to examine if differences between both
 methods can be detected.
@@
 The resulting fitness function \(f\) for an individual program \(i\) is
-descibed by eq 2:
+described by eq 2:
 
@@
 
-The primitve set consists of the terminals that are listed in table 3
-and the functions that are listed in table 4. To avoid runtime errors I
+The primitive set consists of the terminals that are listed in table 3
+and the functions that are listed in table 4. To avoid run time errors I
 implemented protected version of the operators for division, natural
@@
 (1992) I also included ephemeral constants to the set of terminals to
-provide the evolutionay search the opportunity to explore and include
+provide the evolutionary search the opportunity to explore and include
 randomly generated constants.
@@
 GP operators are represented in table 5. Both genetic operators use a
-static limit to control for the height of the reulting trees (Koza,
+static limit to control for the height of the resulting trees (Koza,
 1992, p. 104). Individual programs are initialized by using the ramped
@@
 The crossover operator implemented by DEAP randomly selects a crossover
-point in each individual and exchanges each subtree with the point as
+point in each individual and exchanges each sub tree with the point as
 root between each individual (Fortin \emph{et al.}, 2012). Mutation also
 randomly selects a point in the tree individual, it then replaces the
-subtree below that point as a root by the expression generated using the
-full grow initialization method (Fortin \emph{et al.}, 2012).
+sub tree below that point as a root by the expression generated using
+the full grow initialization method (Fortin \emph{et al.}, 2012).
 
@@
 \(testing\_error\) represents the fitness of the same solution if
-evaluated for the unknown fitness cases inside the testing dataset.
+evaluated for the unknown fitness cases inside the testing data set.
 Tournament selection based GP achieves a mean \(min\_training\_error\)
@@
 \(22.456\). These results indicate an overall performance benefit for
-torunament selection based GP in the experiment. It should also be
+tournament selection based GP in the experiment. It should also be
 mentioned that the best performing model over all 50 runs, measured by
@@
 \(testing\_error\) for both selection operators over the total 50 runs
-using a boxplot and summerizes the initial observations from tables 6
+using a box plot and summarizes the initial observations from tables 6
 and 7 described above.
@@
 8. Since not all samples test positive for a normal distribution a
-Mann-Whitney U (MWU) ranksum test will be performed to test for
+Mann-Whitney U (MWU) rank sum test will be performed to test for
 statistical significance between the four variables
@@
   regression task tournament selection achieves a better mean
-  performance on both the training as well as the testing dataset than
+  performance on both the training as well as the testing data set than
   \(\epsilon\)-Lexicase selection. These findings contradict the
-  hypothesis \(H0_{1}\) descibed in subsection 3.1 and are contrary to
+  hypothesis \(H0_{1}\) described in subsection 3.1 and are contrary to
   some of the research that was summarized in subsections 2.1 and
@@
 
-To further examine the generalization behaviour of both selection
+To further examine the generalization behavior of both selection
 operators I plotted the mean fitness scores for both algorithms over the
 total number of 100 generations in figure 2. Figure 2 shows that both
-algorithms show little visual evidence of overfitting which would result
-in a growing gap between between \(min\_training\_error\) and
+algorithms show little visual evidence of over fitting which would
+result in a growing gap between between \(min\_training\_error\) and
 \(testing\_error\) in the final stages of the evolutionary optimization.
@@
 on the limited amount of collected data, it is questionable if this
-small sign of overfitting can be interpreted as a weak indicator for an
-inferior generalization behaviour of \(\epsilon\)-Lexicase in comparison
+small sign of over fitting can be interpreted as a weak indicator for an
+inferior generalization behavior of \(\epsilon\)-Lexicase in comparison
 to tournament selection.
@@
 performance benefit of using tournament selection instead of
-\(\epsilon\)-Lexicase selection for the choosen symbolic regression
-task. It can be observed that at all stages of the evolution tournament
-selection based GP achieves a lower mean error for both datasets.
+\(\epsilon\)-Lexicase selection for the chosen symbolic regression task.
+It can be observed that at all stages of the evolution tournament
+selection based GP achieves a lower mean error for both data sets.
 
@@
 
-In a final step I analyzed the growth behaviour for both GP systems.
-Figure 3 visualizes the average size of individuals inside the
-population as well as the average size of the best performing individual
-for each generation of the evolution.
-
-\includegraphics{./plots/size_combined.png} \ldots{}
-interpretation\ldots{}
-
-Again, a MWU test is conducted to test for statistical differences in
-the underlying distribution of the sizes measured during the experiment.
-The results are summarized in table 10.
+In a final step I analyze the growth behavior for both GP systems to
+determine if differences in program size can be observed based on
+selection operators. Figure 3 visualizes the average size of individuals
+inside the population as well as the average size of the best performing
+individual for each generation and both GP systems.
+\includegraphics{./plots/size_subplotted.png}
+
+Both subplots display a very similar behavior that is common for GP, we
+observe a strong trend for solutions to grow in size with each
+additional generation that is concluded. Another observation that can be
+made is that the best solutions of each generation almost always have a
+larger size than the population average. On first impression, the
+differences between both selection operators appears to be
+insignificant. Again, I conducted a MWU test to test for statistical
+differences in the underlying distribution of sizes measured during the
+experiment.
+
+The resulting p-values of the MWU test are summarized in table 10.
 
@@
 
+None of the differences in mean size are statistically significant.
+Again, based on the results obtained in my experiment, the evidence
+produced fails to reject the null hypothesis \(H0_{3}\) that seletion
+has no significant effect on differences in program size.
+
 \hypertarget{conclusion}{%
@@
 \item
-  Limited computational ressources
+  Limited computational resources
 \end{enumerate}
@@
 and 6. The performance of all evolutionary algorithms, including GP, can
-be highly dependant on a large number of parameters and implementation
+be highly dependent on a large number of parameters and implementation
 details. Although the series of experiments conducted in this project
@@
 
-To gain further insight into the differences in generalizaion behaviour
+To gain further insight into the differences in generalization behavior
 and to adress the research question with a higher level of certainty,
-the experiment I conducted should be repeated for many different GP
+the experiment I conducted should be repeated for different GP
 configurations. Some exemplary parameters that could drastically
@@
 computation of \(\epsilon\) in lexicase selection. Other factors whose
-influence on generalization behaviour could be important include
+influence on generalization behavior could be important include
 different methods to compute an individuals fitness (e.g.~mean absolute
-error or root mean squared error) and testing out different variations
-of the genetic operators for mutation and crossover.
+error or root mean squared error), testing different variations of the
+genetic operators for mutation and crossover and experimenting with
+different variations of the training:testing split size.
 
@@
 based on a single symbolic regression application, the prediction of the
-heating load of buildings based on a relativley small dataset. To gain
+heating load of buildings based on a relatively small data set. To gain
 further trust in the obtained results, the experiment should be repeated
-for other symbolic regression tasks and datasets. At the current state
-it remains unknown if the generalization behaviour of both selection
-operators might be highly problem specific and dependant on the dataset
+for other symbolic regression tasks and data sets. At the current state
+it remains unknown if the generalization behavior of both selection
+operators might be highly problem specific and dependent on the data set
 used for training and testing.
@@
 the second limitation of this research project, the limited amount of
-computational ressources. The total computation time consumed to run the
+computational resources. The total computation time consumed to run the
 current experiments (50 runs, 2 algorithms, 100 generations, 500
 individuals) has already been close to 24 hours on a modern, arm-based
-Apple-M1 system. Eventhough the overall computation time could probably
+Apple-M1 system. Even though the overall computation time could probably
 be reduced by further optimization of the source code, e.g.~using tools
-for parallel computation, the ressources necessary to repeat the whole
+for parallel computation, the resources necessary to repeat the whole
 experiment for different parameter configurations would certainly be
@@
 
-\newpage
-
 \hypertarget{II}{%
@@
 
+\includegraphics{./misc/erklärung.pdf}\\
+
 \end{document}

### 2022-06-23 docs/rmd/paper.Rmd
autogit

@@
   <center>Research Project for Seminar Information Systems (03.996.3299)</center>
-  <center>Chair of Business Administration and Computer Science</center>
+  <center>FB 03: Chair of Business Administration and Computer Science</center>
   <center>Johannes Gutenberg University Mainz</center>
@@
 All GP experiments will be implemented by using the python programming language in conjunction with DEAP, a framework for distributed evolutionary algorithms that implements various tools and algorithms for genetic programming [@DEAP_JMLR2012].
+A github repository that contains the full source code, tables, plots and other materials represented in this thesis can be accessed at [https://github.com/roman91DE/intelligent_information_systems_research_project](#https://github.com/roman91DE/intelligent_information_systems_research_project).
 
@@
 Figure 3 visualizes the average size of individuals inside the population as well as the average size of the best performing individual for each generation and both GP systems.
-![Mean Size](./plots/size_subplotted.png)
+\begin{figure}
+\centering
+\includegraphics{./plots/size_subplotted.png}
+\caption{Mean Size}
+\end{figure}
+
+<!-- ![Mean Size](./plots/size_subplotted.png)--->
 
@@
 # Conclusion
+
+...
 
@@
 ![Statutory Declaration](./misc/erklärung.pdf)\
-
-```{zsh github, include=FALSE}
-git add * && git commit --m "rendered doc" && git push
-```

### 2022-06-23 docs/rmd/paper.tex
autogit

@@
 
-Chair of Business Administration and Computer Science
+FB 03: Chair of Business Administration and Computer Science
 
@@
 evolutionary algorithms that implements various tools and algorithms for
-genetic programming (Fortin \emph{et al.}, 2012).
+genetic programming (Fortin \emph{et al.}, 2012). A github repository
+that contains the full source code, tables, plots and other materials
+represented in this thesis can be accessed at
+\protect\hyperlink{https:ux2fux2fgithub.comux2froman91DEux2fintelligent_information_systems_research_project}{https://github.com/roman91DE/intelligent\_information\_systems\_research\_project}
 
@@
 individual for each generation and both GP systems.
+
+\begin{figure}
+\centering
 \includegraphics{./plots/size_subplotted.png}
+\caption{Mean Size}
+\end{figure}
 
@@
 
+\ldots{}
+
 \hypertarget{limitations-and-open-questions}{%

### 2022-06-23 docs/rmd/paper.tex
autogit

@@
 represented in this thesis can be accessed at
-\protect\hyperlink{https:ux2fux2fgithub.comux2froman91DEux2fintelligent_information_systems_research_project}{https://github.com/roman91DE/intelligent\_information\_systems\_research\_project}
+\protect\hyperlink{https:ux2fux2fgithub.comux2froman91DEux2fintelligent_information_systems_research_project}{https://github.com/roman91DE/intelligent\_information\_systems\_research\_project}.
 

### 2022-06-23 docs/rmd/paper.Rmd
autogit

@@
   <center>
-    A comparison of $\epsilon$-Lexicase Selection and Tournament Selection
+    A comparison of $\epsilon$-lexicase Selection and Tournament Selection
   </center>
@@
 
+\listoftables
+\listoffigures
+\thispagestyle{empty}
+\newpage
+
 # Introduction
@@
 
-Since regular Lexicase selection has been shown to perform sub optimal on continuous-valued optimization problems, a modified variation called $\epsilon$-Lexicase selection has been suggested for symbolic regression by La Cava et.
+Since regular Lexicase selection has been shown to perform sub optimal on continuous-valued optimization problems, a modified variation called $\epsilon$-lexicase selection has been suggested for symbolic regression by La Cava et.
 al (2016).
-Here, $\epsilon$-Lexicase selection has shown itself to outperform both in overall performance while showing only negligible computational overhead [@epsilon_lexicase_main, p.747].
+Here, $\epsilon$-lexicase selection has shown itself to outperform both in overall performance while showing only negligible computational overhead [@epsilon_lexicase_main, p.747].
 
@@
 <!-- Research Question --->
-This research project tries to answer the question if the usage of $\epsilon$-Lexicase selection influences the generalization behavior of programs that are evolved using GP for symbolic regression if compared to programs that are evolved using traditional tournament selection.
+This research project tries to answer the question if the usage of $\epsilon$-lexicase selection influences the generalization behavior of programs that are evolved using GP for symbolic regression if compared to programs that are evolved using traditional tournament selection.
 
@@
 
-More recent research by @epsilon_lexicase_main suggests $\epsilon$-Lexicase selection as a modified selection operator that can improve overall GP performance if applied to continuous-valued symbolic regression tasks in comparison to tournament and standard lexicase selection.
+More recent research by @epsilon_lexicase_main suggests $\epsilon$-lexicase selection as a modified selection operator that can improve overall GP performance if applied to continuous-valued symbolic regression tasks in comparison to tournament and standard lexicase selection.
 @epsilon_lexicase_main argues that the original concept behind lexicase selection as formulated by @lexicase_first_desciption does not fit the requirements of real-world symbolic regression tasks.
@@
 This stringent pass condition can result in filtering out too many individuals during the selection process [@epsilon_lexicase_main, p.742].
-$\epsilon$-Lexicase adresses this problem by introducing an $\epsilon$ parameter that specifies a range around the elite error, individuals that perform inside this range pass the current selection iteration.
-@https://doi.org/10.48550/arxiv.1709.05394 describes $\epsilon$-Lexicase selection as a more "relaxed version of lexicase selection".
-Different methods to configure the $\epsilon$ Parameter have been explored, for the limited scope of this project I will focus on the most promising implementation of $\epsilon$-Lexicase selection that is automatically adjusted on the basis of the median absolute deviation of errors inside the selection pool [@epsilon_lexicase_main, p.742] [@https://doi.org/10.48550/arxiv.1709.05394, p.6].
+$\epsilon$-lexicase adresses this problem by introducing an $\epsilon$ parameter that specifies a range around the elite error, individuals that perform inside this range pass the current selection iteration.
+@https://doi.org/10.48550/arxiv.1709.05394 describes $\epsilon$-lexicase selection as a more "relaxed version of lexicase selection".
+Different methods to configure the $\epsilon$ Parameter have been explored, for the limited scope of this project I will focus on the most promising implementation of $\epsilon$-lexicase selection that is automatically adjusted on the basis of the median absolute deviation of errors inside the selection pool [@epsilon_lexicase_main, p.742] [@https://doi.org/10.48550/arxiv.1709.05394, p.6].
 
-The performance increase of $\epsilon$-Lexicase selection for symbolic regression problems has been demonstrated and reported for many benchmark problems which led to widespread adaption of it in symbolic regression applications [@epsilon_lexicase_main, p.744-745].
+The performance increase of $\epsilon$-lexicase selection for symbolic regression problems has been demonstrated and reported for many benchmark problems which led to widespread adaption of it in symbolic regression applications [@epsilon_lexicase_main, p.744-745].
 
@@
 
-The statistical analysis of the collected data will first focus on examining the question if, on average, the usage of $\epsilon$-Lexicase selection will result in models that perform significantly different than models that are evolved using tournament selection both.
+The statistical analysis of the collected data will first focus on examining the question if, on average, the usage of $\epsilon$-lexicase selection will result in models that perform significantly different than models that are evolved using tournament selection both.
 
@@
 
-> $H0_{2}$: The distribution underlying the samples of training errors produced by $\epsilon$-Lexicase/torunament selection is the same as the distribution underlying samples of testing errors produced by $\epsilon$-Lexicase/tournament selection
+> $H0_{2}$: The distribution underlying the samples of training errors produced by $\epsilon$-lexicase/torunament selection is the same as the distribution underlying samples of testing errors produced by $\epsilon$-lexicase/tournament selection
 
@@
 
-> $H0_{3}$: No differences in program size exist between the distribution underlying the samples produced by $\epsilon$-Lexicase and the distribution underlying samples of tournament selection
+> $H0_{3}$: No differences in program size exist between the distribution underlying the samples produced by $\epsilon$-lexicase and the distribution underlying samples of tournament selection
 
@@
 Variable $min\_training\_error$ represents the fitness, measured by the MSE, of the best performing solution that has been found during a single run of the experiment while the variable $testing\_error$ represents the fitness of the same solution if evaluated for the unknown fitness cases inside the testing data set.
-Tournament selection based GP achieves a mean $min\_training\_error$ of $16.483$ and a mean $testing\_error$ of $16.905$ while the GP system using $\epsilon$-Lexicase selection can only achieve a mean $min\_training\_error$ of $22.013$ and a mean $testing\_error$ of $22.456$.
+Tournament selection based GP achieves a mean $min\_training\_error$ of $16.483$ and a mean $testing\_error$ of $16.905$ while the GP system using $\epsilon$-lexicase selection can only achieve a mean $min\_training\_error$ of $22.013$ and a mean $testing\_error$ of $22.456$.
 These results indicate an overall performance benefit for tournament selection based GP in the experiment.
@@
 If we compare the differences of $min\_training\_error$ and $testing\_error$ on the basis of each algorithm, we can observe that both tournament and $\epsilon$-lexicase achieve a slightly lower error on the training data than on the testing data.
-If we compute the total difference between the means of $min\_training\_error$ and $testing\_error$ for both algorithms tournament selection amounts to 0.422 and $\epsilon$-Lexicase selection to 0.443.
+If we compute the total difference between the means of $min\_training\_error$ and $testing\_error$ for both algorithms tournament selection amounts to 0.422 and $\epsilon$-lexicase selection to 0.443.
 Regarding our primary research question, the differences in generalization, the initial observations do not seem to indicate a significant difference in the the relative gap between the solutions performance on testing data and training data.
@@
 1.
-The differences in mean fitness between tournament selection and $\epsilon$-Lexicase selection are highly statistical significant even if considering an $\alpha$ of 0.01.
-For the given symbolic regression task tournament selection achieves a better mean performance on both the training as well as the testing data set than $\epsilon$-Lexicase selection.
+The differences in mean fitness between tournament selection and $\epsilon$-lexicase selection are highly statistical significant even if considering an $\alpha$ of 0.01.
+For the given symbolic regression task tournament selection achieves a better mean performance on both the training as well as the testing data set than $\epsilon$-lexicase selection.
 These findings contradict the hypothesis $H0_{1}$ described in subsection 3.1 and are contrary to some of the research that was summarized in subsections 2.1 and 2.3.[@epsilon_lexicase_main] [@https://doi.org/10.48550/arxiv.1709.05394].
@@
 The differences in mean fitness between $min\_training\_error$ and $testing\_error$ are not statistically significant for both selection algorithms.
-If comparing the mean training error to the mean testing error, the p-values only computes to 0.309 for tournament selection and to 0.257 for $\epsilon$-Lexicase selection.
+If comparing the mean training error to the mean testing error, the p-values only computes to 0.309 for tournament selection and to 0.257 for $\epsilon$-lexicase selection.
 These findings fail to reject $H0_{2}$, based on the obtained results we can not find evidence that both algorithms demonstrate a significantly different performance between the training data and the previously unseen testing data.
@@
 Figure 2 shows that both algorithms show little visual evidence of over fitting which would result in a growing gap between between $min\_training\_error$ and $testing\_error$ in the final stages of the evolutionary optimization.
-For $\epsilon$-Lexicase selection a small gap can be observed starting at around the 90.th generation, while the minimum training error slowly converges a small increase in the testing error can be observed.
-Based on the limited amount of collected data, it is questionable if this small sign of over fitting can be interpreted as a weak indicator for an inferior generalization behavior of $\epsilon$-Lexicase in comparison to tournament selection.
+For $\epsilon$-lexicase selection a small gap can be observed starting at around the 90.th generation, while the minimum training error slowly converges a small increase in the testing error can be observed.
+Based on the limited amount of collected data, it is questionable if this small sign of over fitting can be interpreted as a weak indicator for an inferior generalization behavior of $\epsilon$-lexicase in comparison to tournament selection.
 
-Figure 2 also shows a visual representation of the described overall performance benefit of using tournament selection instead of $\epsilon$-Lexicase selection for the chosen symbolic regression task.
+Figure 2 also shows a visual representation of the described overall performance benefit of using tournament selection instead of $\epsilon$-lexicase selection for the chosen symbolic regression task.
 It can be observed that at all stages of the evolution tournament selection based GP achieves a lower mean error for both data sets.
@@
 None of the differences in mean size are statistically significant.
-Again, based on the results obtained in my experiment, the evidence produced fails to reject the null hypothesis $H0_{3}$ that seletion has no significant effect on differences in program size.
+Again, based on the results obtained in my experiment, the evidence produced fails to reject the null hypothesis $H0_{3}$ that selection has no significant effect on differences in program size.
+It should be noted that the configuration of both GP algorithm did include two mechanisms that place limitations on the growth behavior of the individual solutions to avoid excessive bloating (see sub section 3.2.4).
 
@@
 
-...
+The objective of this research project has been to answer the question if the choice between two commonly used parent selection operators, tournament selection and $\epsilon$-lexicase selection, significantly influences the generalization behavior of GP for symbolic regression tasks.
+
+The results obtained from my experiment did not demonstrate any statistically significant differences in generalization, both GP systems achieve a slightly lower mean error if evaluated for the training data sets than if evaluated for the unseen fitness cases of the testing data set.
+The mean differences between testing and training errors did not show statistical significance differences for both algorithms (summarized in table 9).
+
+To further explore differences in generalization behavior the mean for both algorithms training and testing error over all generations for all 50 independent runs was visualized and analyzed (see figure 2).
+A very small and probably insignificant gap did show for $\epsilon$-lexicase selection in the final phase of the evolutionary search, in the grand scheme of things this approach also did not yield any clear evidence of differences in generalization.
+
+A final attempt to explore possible differences between both GP systems has been to explore differences in growth behavior as an indicator for generalization.
+This idea was based on the wide spread hypothesis that increases in overall program size might be an indication of over fitting to the training data, the concept has been discussed in sub section 2.2.
+Again, based on my results I could not find statistically significant differences in the growth behavior between both selection operators.
+
+An interesting finding of my analysis is, that the overall performance of tournament selection has shown to be significantly better than that of $\epsilon$-lexicase selection for the selected symbolic regression problem.
+This finding was rather unexpected based on the research of @epsilon_lexicase_main and @https://doi.org/10.48550/arxiv.1709.05394 that primarily inspired this research project in the first place.
 
@@
 
-# I   References {.unnumbered #I}
+# I - References {.unnumbered #I}
 
@@
 
-# II  Statutory Declaration {.unnumbered #II}
+# II - Statutory Declaration {.unnumbered #II}
 
-![Statutory Declaration](./misc/erklärung.pdf)\
+![Statutory Declaration](./private/erklaerung.pdf)\

### 2022-06-23 docs/rmd/paper.tex
autogit

@@
 
-A comparison of \(\epsilon\)-Lexicase Selection and Tournament Selection
+A comparison of \(\epsilon\)-lexicase Selection and Tournament Selection
 
@@
 \tableofcontents
+\thispagestyle{empty}
+\newpage
+
+\listoftables
+\listoffigures
 \thispagestyle{empty}
@@
 on continuous-valued optimization problems, a modified variation called
-\(\epsilon\)-Lexicase selection has been suggested for symbolic
-regression by La Cava et. al (2016). Here, \(\epsilon\)-Lexicase
+\(\epsilon\)-lexicase selection has been suggested for symbolic
+regression by La Cava et. al (2016). Here, \(\epsilon\)-lexicase
 selection has shown itself to outperform both in overall performance
@@
 This research project tries to answer the question if the usage of
-\(\epsilon\)-Lexicase selection influences the generalization behavior
+\(\epsilon\)-lexicase selection influences the generalization behavior
 of programs that are evolved using GP for symbolic regression if
@@
 More recent research by La Cava, Spector and Danai (2016) suggests
-\(\epsilon\)-Lexicase selection as a modified selection operator that
+\(\epsilon\)-lexicase selection as a modified selection operator that
 can improve overall GP performance if applied to continuous-valued
@@
 out too many individuals during the selection process (La Cava, Spector
-and Danai, 2016, p. 742). \(\epsilon\)-Lexicase adresses this problem by
+and Danai, 2016, p. 742). \(\epsilon\)-lexicase adresses this problem by
 introducing an \(\epsilon\) parameter that specifies a range around the
@@
 selection iteration. La Cava \emph{et al.} (2017) describes
-\(\epsilon\)-Lexicase selection as a more ``relaxed version of lexicase
+\(\epsilon\)-lexicase selection as a more ``relaxed version of lexicase
 selection''. Different methods to configure the \(\epsilon\) Parameter
 have been explored, for the limited scope of this project I will focus
-on the most promising implementation of \(\epsilon\)-Lexicase selection
+on the most promising implementation of \(\epsilon\)-lexicase selection
 that is automatically adjusted on the basis of the median absolute
@@
 
-The performance increase of \(\epsilon\)-Lexicase selection for symbolic
+The performance increase of \(\epsilon\)-lexicase selection for symbolic
 regression problems has been demonstrated and reported for many
@@
 examining the question if, on average, the usage of
-\(\epsilon\)-Lexicase selection will result in models that perform
+\(\epsilon\)-lexicase selection will result in models that perform
 significantly different than models that are evolved using tournament
@@
 \(H0_{2}\): The distribution underlying the samples of training errors
-produced by \(\epsilon\)-Lexicase/torunament selection is the same as
+produced by \(\epsilon\)-lexicase/torunament selection is the same as
 the distribution underlying samples of testing errors produced by
-\(\epsilon\)-Lexicase/tournament selection
+\(\epsilon\)-lexicase/tournament selection
 \end{quote}
@@
 \(H0_{3}\): No differences in program size exist between the
-distribution underlying the samples produced by \(\epsilon\)-Lexicase
+distribution underlying the samples produced by \(\epsilon\)-lexicase
 and the distribution underlying samples of tournament selection
@@
 of \(16.483\) and a mean \(testing\_error\) of \(16.905\) while the GP
-system using \(\epsilon\)-Lexicase selection can only achieve a mean
+system using \(\epsilon\)-lexicase selection can only achieve a mean
 \(min\_training\_error\) of \(22.013\) and a mean \(testing\_error\) of
@@
 \(testing\_error\) for both algorithms tournament selection amounts to
-0.422 and \(\epsilon\)-Lexicase selection to 0.443. Regarding our
+0.422 and \(\epsilon\)-lexicase selection to 0.443. Regarding our
 primary research question, the differences in generalization, the
@@
   The differences in mean fitness between tournament selection and
-  \(\epsilon\)-Lexicase selection are highly statistical significant
+  \(\epsilon\)-lexicase selection are highly statistical significant
   even if considering an \(\alpha\) of 0.01. For the given symbolic
@@
   performance on both the training as well as the testing data set than
-  \(\epsilon\)-Lexicase selection. These findings contradict the
+  \(\epsilon\)-lexicase selection. These findings contradict the
   hypothesis \(H0_{1}\) described in subsection 3.1 and are contrary to
@@
   testing error, the p-values only computes to 0.309 for tournament
-  selection and to 0.257 for \(\epsilon\)-Lexicase selection. These
+  selection and to 0.257 for \(\epsilon\)-lexicase selection. These
   findings fail to reject \(H0_{2}\), based on the obtained results we
@@
 \(testing\_error\) in the final stages of the evolutionary optimization.
-For \(\epsilon\)-Lexicase selection a small gap can be observed starting
+For \(\epsilon\)-lexicase selection a small gap can be observed starting
 at around the 90.th generation, while the minimum training error slowly
@@
 small sign of over fitting can be interpreted as a weak indicator for an
-inferior generalization behavior of \(\epsilon\)-Lexicase in comparison
+inferior generalization behavior of \(\epsilon\)-lexicase in comparison
 to tournament selection.
@@
 performance benefit of using tournament selection instead of
-\(\epsilon\)-Lexicase selection for the chosen symbolic regression task.
+\(\epsilon\)-lexicase selection for the chosen symbolic regression task.
 It can be observed that at all stages of the evolution tournament
@@
 Again, based on the results obtained in my experiment, the evidence
-produced fails to reject the null hypothesis \(H0_{3}\) that seletion
-has no significant effect on differences in program size.
+produced fails to reject the null hypothesis \(H0_{3}\) that selection
+has no significant effect on differences in program size. It should be
+noted that the configuration of both GP algorithm did include two
+mechanisms that place limitations on the growth behavior of the
+individual solutions to avoid excessive bloating (see sub section
+3.2.4).
 
@@
 
-\ldots{}
+The objective of this research project has been to answer the question
+if the choice between two commonly used parent selection operators,
+tournament selection and \(\epsilon\)-lexicase selection, significantly
+influences the generalization behavior of GP for symbolic regression
+tasks.
+
+The results obtained from my experiment did not demonstrate any
+statistically significant differences in generalization, both GP systems
+achieve a slightly lower mean error if evaluated for the training data
+sets than if evaluated for the unseen fitness cases of the testing data
+set. The mean differences between testing and training errors did not
+show statistical significance differences for both algorithms
+(summarized in table 9).
+
+To further explore differences in generalization behavior the mean for
+both algorithms training and testing error over all generations for all
+50 independent runs was visualized and analyzed (see figure 2). A very
+small and probably insignificant gap did show for \(\epsilon\)-lexicase
+selection in the final phase of the evolutionary search, in the grand
+scheme of things this approach also did not yield any clear evidence of
+differences in generalization.
+
+A final attempt to explore possible differences between both GP systems
+has been to explore differences in growth behavior as an indicator for
+generalization. This idea was based on the wide spread hypothesis that
+increases in overall program size might be an indication of over fitting
+to the training data, the concept has been discussed in sub section 2.2.
+Again, based on my results I could not find statistically significant
+differences in the growth behavior between both selection operators.
+
+An interesting finding of my analysis is, that the overall performance
+of tournament selection has shown to be significantly better than that
+of \(\epsilon\)-lexicase selection for the selected symbolic regression
+problem. This finding was rather unexpected based on the research of La
+Cava, Spector and Danai (2016) and La Cava \emph{et al.} (2017) that
+primarily inspired this research project in the first place.
 
@@
 \hypertarget{I}{%
-\section*{I References}\label{I}}
-\addcontentsline{toc}{section}{I References}
+\section*{I - References}\label{I}}
+\addcontentsline{toc}{section}{I - References}
 
@@
 \hypertarget{II}{%
-\section*{II Statutory Declaration}\label{II}}
-\addcontentsline{toc}{section}{II Statutory Declaration}
-
-\includegraphics{./misc/erklärung.pdf}\\
+\section*{II - Statutory Declaration}\label{II}}
+\addcontentsline{toc}{section}{II - Statutory Declaration}
+
+\includegraphics{./private/erklaerung.pdf}\\
 

### 2022-06-23 docs/rmd/paper.Rmd
autogit

@@
 All GP experiments will be implemented by using the python programming language in conjunction with DEAP, a framework for distributed evolutionary algorithms that implements various tools and algorithms for genetic programming [@DEAP_JMLR2012].
-A github repository that contains the full source code, tables, plots and other materials represented in this thesis can be accessed at [https://github.com/roman91DE/intelligent_information_systems_research_project](#https://github.com/roman91DE/intelligent_information_systems_research_project).
+A repository that contains the full source code, tables, plots and other materials represented in this thesis is available at github.com^[https://github.com/roman91DE/intelligent_information_systems_research_project].
 
@@
 
-<!-- ![Mean Size](./plots/size_subplotted.png)--->
+<!-- ![Mean Size](./plots/size_sub plotted.png)--->
 
@@
 
-To gain further insight into the differences in generalization behavior and to adress the research question with a higher level of certainty, the experiment I conducted should be repeated for different GP configurations.
+To gain further insight into the differences in generalization behavior and to address the research question with a higher level of certainty, the experiment I conducted should be repeated for different GP configurations.
 Some exemplary parameters that could drastically influence the results of this experiment include the population size, the total number of generations, the inclusion of elitism or the computation of $\epsilon$ in lexicase selection.

### 2022-06-23 docs/rmd/paper.tex
autogit

@@
 evolutionary algorithms that implements various tools and algorithms for
-genetic programming (Fortin \emph{et al.}, 2012). A github repository
-that contains the full source code, tables, plots and other materials
-represented in this thesis can be accessed at
-\protect\hyperlink{https:ux2fux2fgithub.comux2froman91DEux2fintelligent_information_systems_research_project}{https://github.com/roman91DE/intelligent\_information\_systems\_research\_project}.
+genetic programming (Fortin \emph{et al.}, 2012). A repository that
+contains the full source code, tables, plots and other materials
+represented in this thesis is available at github.com\footnote{\url{https://github.com/roman91DE/intelligent_information_systems_research_project}}.
 
@@
 To gain further insight into the differences in generalization behavior
-and to adress the research question with a higher level of certainty,
+and to address the research question with a higher level of certainty,
 the experiment I conducted should be repeated for different GP

### 2022-06-23 docs/rmd/paper.Rmd
autogit

@@
 This research project tries to answer the question if the usage of $\epsilon$-lexicase selection influences the generalization behavior of programs that are evolved using GP for symbolic regression if compared to programs that are evolved using traditional tournament selection.
+
+<!-- Vorgehen --->
+Section 2 will give a brief summary and discussion of topic related scientific research.
+Section 3 describes the design of the experiment as well as GP specific implementation.
+The detailed evaluation of all relevant results and the summarized conclusions are given in section 3 and 4.
+Section 5 addresses limitation and open questions of this research project.
 

### 2022-06-23 docs/rmd/paper.tex
autogit

@@
 
+Section 2 will give a brief summary and discussion of topic related
+scientific research. Section 3 describes the design of the experiment as
+well as GP specific implementation. The detailed evaluation of all
+relevant results and the summarized conclusions are given in section 3
+and 4. Section 5 addresses limitation and open questions of this
+research project.
+
 \hypertarget{an-overview-of-the-current-state-of-research}{%

### 2022-06-23 docs/rmd/paper.Rmd
autogit

--- 2022-06-23 docs/rmd/paper.Rmd
@@
 Tournament Selection is a commonly used selection operator in EC and is the most used operator in GP systems [@10.1007/978-3-642-16493-4_19, p.181].
-A parent solution is selected by randomly sampling $k$ individuals from the current population into a tournament pool and then the solution with the highest fitness score from the tournament pool is selected [@10.1007/978-3-642-16493-4_19, p.182].
+A parent solution is selected by randomly sampling $k$ individuals from the current population into a tournament pool and then the solution with the highest fitness score from the pool is selected [@10.1007/978-3-642-16493-4_19, p.182].
 Lexicase selection has been suggested as an alternative to tournament selection that is not based on aggregating fitness scores.
@@
 The aim is to produce a model that not only accurately predicts the provided training data but can also predict previously unseen cases with high precision [@Gonalves2016AnEO, p.6].
-A model that is extensively optimized on the provided training data, can become over fitted to this data sample which may lead to a decrease in generalization.
+A model that is extensively optimized on training data, can become over fitted to the data sample which may lead to a decrease in generalization.
 
@@
 This research project tries to answer the question if the usage of $\epsilon$-lexicase selection influences the generalization behavior of programs that are evolved using GP for symbolic regression if compared to programs that are evolved using traditional tournament selection.
+
+<!-- Vorgehen --->
+Section 2 will give a brief summary and discussion of topic related scientific research.
+Section 3 describes the design of the experiment as well as GP specific implementation details.
+The statistical  evaluation of all relevant results and the summarized conclusions are given in section 3 and 4.
+Section 5 addresses limitation and open questions of this research project.
 
@@
 
-Both articles argue that many of the problem domains that GP is commonly used for are prevalent for problems that are modal/ uncompromising in nature.
+Both articles argue that many of the problem domains that GP is commonly used for are prevalent for problems that are modal or uncompromising in nature.
 In comparison to other selection operators, populations that are evolved using variations of the lexicase selection operator show a very high degree of genetic diversity which might be a key contributor to the improved performance [@6920034, p.1] [@epsilon_lexicase_main, p.745].
@@
 @epsilon_lexicase_main argues that the original concept behind lexicase selection as formulated by @lexicase_first_desciption does not fit the requirements of real-world symbolic regression tasks.
-The authors identifies the pass condition of regular lexicase selection as the main problem if applied to symbolic regression: Individual solutions can only be selected if they perform on an elite level but with continuous-valued, and often noisy, data it is very unlikely for two individuals to achieve an exactly equal error for a given test case.
+The authors identifies the pass condition of regular lexicase selection as the main problem if applied to symbolic regression: Individual solutions can only be selected if they perform on an elite level but with continuous-valued, and often noisy, data it is very unlikely for two individuals to achieve an exact equal error for a given test case.
 This stringent pass condition can result in filtering out too many individuals during the selection process [@epsilon_lexicase_main, p.742].
@@
       based on all fitness cases.
-  (b) Select individual from the tournament pool with the highest mean fitness.
+  (b) Select individual from the tournament pool with the highest mean fitness score.
 ```
@@
 Generalization, the ability of a model to perform well on previously unseen cases, is one of the fundamental goals in most real world machine learning applications.
-@open_issues_gp raises awareness to the fact, that the topic of generalization in GP has not gotten the attention that other machine learning-based areas  have attributed to it.
+@open_issues_gp raises awareness to the fact, that the topic of generalization in GP has not gotten the attention that other machine learning-based areas attributed to it.
 A similar statement has been proposed by @generalisation_in_gp in his review of research in generalization in GP.
@@
 An aspect that is suspected to negatively correlate with GP's generalization is overall size and in particular bloating of the resulting programs.
-Bloating can be described as a growth in the total size of a program that does not also improve it's performance in any meaningful way [@bloat_overfitting_gp, p.
+Bloating can be described as a growth in the total size of a program that does not improve its performance in any meaningful way [@bloat_overfitting_gp, p.
 1].
@@
 349].
-However MDLP should be considered carefully, experiments by @bloat_overfitting_gp on the prediction of bio availability for drugs demonstrate that GP based techniques can either bloat without over fitting or over fit without bloating [@bloat_overfitting_gp, p.
+However MDLP should be considered with caution, experiments by @bloat_overfitting_gp on the prediction of bio availability for drugs demonstrate that GP based techniques can either bloat without over fitting or over fit without bloating [@bloat_overfitting_gp, p.
 8].
@@
 
-Since it's inception, GP based symbolic regression has been widely used in practical applications ranging over a wide field of disciplines.
+Since its inception, GP based symbolic regression has been widely used in practical applications ranging over a wide field of disciplines.
 A well known example of an industrial use-case has been published by @soft_sensor_gp.
-The authors use GP based symbolic regression to derive nonlinear functions that can be used to significantly increase the robustness of industrial sensors which resulted in wide-spread adoption and a significant reduction in costs.
+The authors use GP based symbolic regression to derive nonlinear functions that can be used to significantly increase the robustness of industrial sensors.
+This resulted in wide-spread adoption and a significant reduction in costs.
 
@@
 
-Using two otherwise identical GP systems, one deploying tournament selection and the other $\epsilon$-lexicase selection, the objective is to find a computer program that best predicts the outcome variable heating load $(Y1)$^[The symbolic regression is performed on one of the two provided outcome variables, the variable Cooling Load will be excluded.] of the buildings using a subset of the eight building attributes $(X1,..,X8)$ for it's input.
+Using two otherwise identical GP systems, one deploying tournament selection and the other $\epsilon$-lexicase selection, the objective is to find a computer program that best predicts the outcome variable heating load $(Y1)$^[The symbolic regression is performed on one of the two provided outcome variables, the variable Cooling Load will be excluded.] of the buildings using a subset of the eight building attributes $(X1,..,X8)$ for its input.
 The specific meaning of all attributes inside the data set are described in table 1.
@@
 
-The statistical analysis of the collected data will first focus on examining the question if, on average, the usage of $\epsilon$-lexicase selection will result in models that perform significantly different than models that are evolved using tournament selection both.
+The statistical analysis of the collected data will first focus on examining the question if, on average, the usage of $\epsilon$-lexicase selection will result in models that perform significantly different than models that are evolved using tournament selection:
 
@@
 
-## Gentetic Programming Configuration
+## Genetic Programming Configuration
 
@@
 
-The primitive set consists of the terminals that are listed in table 3 and the functions that are listed in table 4.
+The primitive set consists of the terminals listed in table 3 and the functions listed in table 4.
 To avoid run time errors I implemented protected version of the operators for division, natural logarithm and square root [@koza_main, p.82-83].
@@
 Other factors whose influence on generalization behavior could be important include different methods to compute an individuals fitness (e.g.
-mean absolute error or root mean squared error), testing different variations of the genetic operators for mutation and crossover and experimenting with different variations of the training:testing split size.
+mean absolute error or root mean squared error), testing different variations of the genetic operators for mutation and crossover and experimenting with different variations of the training:testing data set split size.
 
-Another obvious weakness of my project is that the experiment is only based on a single symbolic regression application, the prediction of the heating load of buildings based on a relatively small data set.
+Another obvious weakness of this project is that the experiment is only based on a single symbolic regression application, the prediction of the heating load of buildings based on a relatively small data set.
 To gain further trust in the obtained results, the experiment should be repeated for other symbolic regression tasks and data sets.
-At the current state it remains unknown if the generalization behavior of both selection operators might be highly problem specific and dependent on the data set used for training and testing.
+It remains unknown if the generalization behavior of both selection operators might be highly problem specific and dependent on the data set used for training and testing.
 

### 2022-06-23 docs/rmd/paper.tex
autogit

--- 2022-06-23 docs/rmd/paper.tex
@@
 individuals from the current population into a tournament pool and then
-the solution with the highest fitness score from the tournament pool is
-selected (Fang and Li, 2010, p. 182). Lexicase selection has been
-suggested as an alternative to tournament selection that is not based on
-aggregating fitness scores. It samples \(n\) test cases in random order
-and then eliminates solutions from the selection pool on a per test case
-basis if they are not performing on an elite level (Helmuth, Spector and
+the solution with the highest fitness score from the pool is selected
+(Fang and Li, 2010, p. 182). Lexicase selection has been suggested as an
+alternative to tournament selection that is not based on aggregating
+fitness scores. It samples \(n\) test cases in random order and then
+eliminates solutions from the selection pool on a per test case basis if
+they are not performing on an elite level (Helmuth, Spector and
 Matheson, 2015, p. 1).
@@
 previously unseen cases with high precision (Gonçalves, 2016, p. 6). A
-model that is extensively optimized on the provided training data, can
-become over fitted to this data sample which may lead to a decrease in
+model that is extensively optimized on training data, can become over
+fitted to the data sample which may lead to a decrease in
 generalization.
@@
 selection.
+
+Section 2 will give a brief summary and discussion of topic related
+scientific research. Section 3 describes the design of the experiment as
+well as GP specific implementation details. The statistical evaluation
+of all relevant results and the summarized conclusions are given in
+section 3 and 4. Section 5 addresses limitation and open questions of
+this research project.
 
@@
 Both articles argue that many of the problem domains that GP is commonly
-used for are prevalent for problems that are modal/ uncompromising in
+used for are prevalent for problems that are modal or uncompromising in
 nature. In comparison to other selection operators, populations that are
@@
 level but with continuous-valued, and often noisy, data it is very
-unlikely for two individuals to achieve an exactly equal error for a
-given test case. This stringent pass condition can result in filtering
-out too many individuals during the selection process (La Cava, Spector
-and Danai, 2016, p. 742). \(\epsilon\)-lexicase adresses this problem by
+unlikely for two individuals to achieve an exact equal error for a given
+test case. This stringent pass condition can result in filtering out too
+many individuals during the selection process (La Cava, Spector and
+Danai, 2016, p. 742). \(\epsilon\)-lexicase adresses this problem by
 introducing an \(\epsilon\) parameter that specifies a range around the
@@
       based on all fitness cases.
-  (b) Select individual from the tournament pool with the highest mean fitness.
+  (b) Select individual from the tournament pool with the highest mean fitness score.
 \end{verbatim}
@@
 the fact, that the topic of generalization in GP has not gotten the
-attention that other machine learning-based areas have attributed to it.
-A similar statement has been proposed by Kushchu (2002) in his review of
+attention that other machine learning-based areas attributed to it. A
+similar statement has been proposed by Kushchu (2002) in his review of
 research in generalization in GP. Both authors address the need for
@@
 resulting programs. Bloating can be described as a growth in the total
-size of a program that does not also improve it's performance in any
+size of a program that does not improve its performance in any
 meaningful way (Silva and Vanneschi, 2009, p. 1). It has been widely
@@
 comparable fitness during the training phase (O'Neill \emph{et al.},
-2010, p. 349). However MDLP should be considered carefully, experiments
-by Silva and Vanneschi (2009) on the prediction of bio availability for
-drugs demonstrate that GP based techniques can either bloat without over
-fitting or over fit without bloating (Silva and Vanneschi, 2009, p. 8).
+2010, p. 349). However MDLP should be considered with caution,
+experiments by Silva and Vanneschi (2009) on the prediction of bio
+availability for drugs demonstrate that GP based techniques can either
+bloat without over fitting or over fit without bloating (Silva and
+Vanneschi, 2009, p. 8).
 
@@
 
-Since it's inception, GP based symbolic regression has been widely used
+Since its inception, GP based symbolic regression has been widely used
 in practical applications ranging over a wide field of disciplines. A
@@
 regression to derive nonlinear functions that can be used to
-significantly increase the robustness of industrial sensors which
+significantly increase the robustness of industrial sensors. This
 resulted in wide-spread adoption and a significant reduction in costs.
@@
   will be excluded.} of the buildings using a subset of the eight
-building attributes \((X1,..,X8)\) for it's input. The specific meaning
+building attributes \((X1,..,X8)\) for its input. The specific meaning
 of all attributes inside the data set are described in table 1.
@@
 significantly different than models that are evolved using tournament
-selection both.
+selection:
 
@@
 
-\hypertarget{gentetic-programming-configuration}{%
-\subsection{Gentetic Programming
-Configuration}\label{gentetic-programming-configuration}}
+\hypertarget{genetic-programming-configuration}{%
+\subsection{Genetic Programming
+Configuration}\label{genetic-programming-configuration}}
 
@@
 
-The primitive set consists of the terminals that are listed in table 3
-and the functions that are listed in table 4. To avoid run time errors I
-implemented protected version of the operators for division, natural
-logarithm and square root (Koza, 1992, pp. 82--83). As suggested by Koza
-(1992) I also included ephemeral constants to the set of terminals to
-provide the evolutionary search the opportunity to explore and include
-randomly generated constants.
+The primitive set consists of the terminals listed in table 3 and the
+functions listed in table 4. To avoid run time errors I implemented
+protected version of the operators for division, natural logarithm and
+square root (Koza, 1992, pp. 82--83). As suggested by Koza (1992) I also
+included ephemeral constants to the set of terminals to provide the
+evolutionary search the opportunity to explore and include randomly
+generated constants.
 
@@
 genetic operators for mutation and crossover and experimenting with
-different variations of the training:testing split size.
-
-Another obvious weakness of my project is that the experiment is only
+different variations of the training:testing data set split size.
+
+Another obvious weakness of this project is that the experiment is only
 based on a single symbolic regression application, the prediction of the
@@
 further trust in the obtained results, the experiment should be repeated
-for other symbolic regression tasks and data sets. At the current state
-it remains unknown if the generalization behavior of both selection
-operators might be highly problem specific and dependent on the data set
-used for training and testing.
+for other symbolic regression tasks and data sets. It remains unknown if
+the generalization behavior of both selection operators might be highly
+problem specific and dependent on the data set used for training and
+testing.
 

### 2022-06-23 docs/rmd/paper.Rmd
autogit

@@
 
-@generalisation_in_gp criticizes that almost all applications published in the initial GP literature by @koza_main are not using separate training and testing data sets which might result in over fitting and a poor overall generalization ability of the programs that are produced.
+@generalisation_in_gp criticizes that almost all applications published in the initial GP literature by @koza_main are not using separate training and testing data sets which might result in overfitting and a poor overall generalization ability of the programs that are produced.
 The author calls for a more widespread adoption of methods like generational sampling of new training cases or the overall separation into training and testing data sets to improve generalization in GP [@generalisation_in_gp, p.10].
@@
 349].
-However MDLP should be considered with caution, experiments by @bloat_overfitting_gp on the prediction of bio availability for drugs demonstrate that GP based techniques can either bloat without over fitting or over fit without bloating [@bloat_overfitting_gp, p.
+However MDLP should be considered with caution, experiments by @bloat_overfitting_gp on the prediction of bio availability for drugs demonstrate that GP based techniques can either bloat without overfitting or over fit without bloating [@bloat_overfitting_gp, p.
 8].
@@
 To further examine the difference in generalization behavior, the mean testing and training errors over each generation will be visualized for both algorithms.
-The specific aim of this visualization is to explore the mechanism of over fitting and to examine if differences between both methods can be detected.
+The specific aim of this visualization is to explore the mechanism of overfitting and to examine if differences between both methods can be detected.
 
@@
 
-Some interesting observation from tables 6 and 7 can be made by comparing the mean of the variables $min\_training\_error$ and $testing\_error$.
-Variable $min\_training\_error$ represents the fitness, measured by the MSE, of the best performing solution that has been found during a single run of the experiment while the variable $testing\_error$ represents the fitness of the same solution if evaluated for the unknown fitness cases inside the testing data set.
+A few interesting observations from tables 6 and 7 can be made by comparing the mean of the variables $min\_training\_error$ and $testing\_error$.
+Variable $min\_training\_error$ represents the fitness, measured by the MSE, of the best performing solution that has been found during a single run of the experiment.
+The variable $testing\_error$ represents the fitness of the same solution if evaluated for the unknown fitness cases inside the testing data set.
 Tournament selection based GP achieves a mean $min\_training\_error$ of $16.483$ and a mean $testing\_error$ of $16.905$ while the GP system using $\epsilon$-lexicase selection can only achieve a mean $min\_training\_error$ of $22.013$ and a mean $testing\_error$ of $22.456$.
-These results indicate an overall performance benefit for tournament selection based GP in the experiment.
+These results indicate an overall performance benefit for the tournament selection based GP in the experiment.
 It should also be mentioned that the best performing model over all 50 runs, measured by the minimum of variables $min\_training\_error$ and $testing\_error$, has been achieved by tournament selection with an MSE of 7.066 for the training data and	7.847 for the testing data.
 
-If we compare the differences of $min\_training\_error$ and $testing\_error$ on the basis of each algorithm, we can observe that both tournament and $\epsilon$-lexicase achieve a slightly lower error on the training data than on the testing data.
-If we compute the total difference between the means of $min\_training\_error$ and $testing\_error$ for both algorithms tournament selection amounts to 0.422 and $\epsilon$-lexicase selection to 0.443.
-Regarding our primary research question, the differences in generalization, the initial observations do not seem to indicate a significant difference in the the relative gap between the solutions performance on testing data and training data.
+If we compare the differences of $min\_training\_error$ and $testing\_error$ on the basis of each algorithm, we can observe that both tournament and $\epsilon$-lexicase achieve a slightly lower fitness on the training data than on the testing data.
+We compute the total difference between the means of $min\_training\_error$ and $testing\_error$ for both algorithms tournament selection amounts to 0.422 and $\epsilon$-lexicase selection to 0.443.
+The initial observations do not seem to indicate a significant difference in the the relative gap between the solutions' performance on testing data and training data.
 
@@
 2.
-The differences in mean fitness between $min\_training\_error$ and $testing\_error$ are not statistically significant for both selection algorithms.
-If comparing the mean training error to the mean testing error, the p-values only computes to 0.309 for tournament selection and to 0.257 for $\epsilon$-lexicase selection.
+The differences in mean fitness between $min\_training\_error$ and $testing\_error$ are not statistically significant for neither selection algorithms.
+When comparing the mean training error to the mean testing error, the p-values only compute to 0.309 for tournament selection and to 0.257 for $\epsilon$-lexicase selection.
 These findings fail to reject $H0_{2}$, based on the obtained results we can not find evidence that both algorithms demonstrate a significantly different performance between the training data and the previously unseen testing data.
@@
 To further examine the generalization behavior of both selection operators I plotted the mean fitness scores for both algorithms over the total number of 100 generations in figure 2.
-Figure 2 shows that both algorithms show little visual evidence of over fitting which would result in a growing gap between between $min\_training\_error$ and $testing\_error$ in the final stages of the evolutionary optimization.
-For $\epsilon$-lexicase selection a small gap can be observed starting at around the 90.th generation, while the minimum training error slowly converges a small increase in the testing error can be observed.
-Based on the limited amount of collected data, it is questionable if this small sign of over fitting can be interpreted as a weak indicator for an inferior generalization behavior of $\epsilon$-lexicase in comparison to tournament selection.
+Figure 2 shows that both algorithms show little visual evidence of overfitting which would result in a growing gap between between $min\_training\_error$ and $testing\_error$ in the final stages of the evolutionary optimization.
+For $\epsilon$-lexicase selection a small gap can be observed starting approximately at the 90th generation, while the minimum training error slowly converges a small increase in the testing error.
+Based on the limited amount of collected data, it is questionable if this small sign of overfitting can be interpreted as a weak indicator for an inferior generalization behavior of $\epsilon$-lexicase in comparison to tournament selection.
 
@@
 
-Both subplots display a very similar behavior that is common for GP, we observe a strong trend for solutions to grow in size with each additional generation that is concluded.
+Both subplots display a very similar behavior that is common for GP.
+We observe a strong trend for solutions to grow in size with each additional generation that is concluded.
 Another observation  that can be made is that the best solutions of each generation almost always have a larger size than the population average.
-On first impression, the differences between both selection operators appears to be insignificant.
-Again, I conducted a MWU test to test for statistical differences in the underlying distribution of sizes measured during the experiment.
+On first impression, the differences between both selection operators appear to be insignificant.
+Again, I conducted a MWU test to examine statistical differences in the underlying distribution of sizes measured during the experiment.
 
@@
 Again, based on the results obtained in my experiment, the evidence produced fails to reject the null hypothesis $H0_{3}$ that selection has no significant effect on differences in program size.
-It should be noted that the configuration of both GP algorithm did include two mechanisms that place limitations on the growth behavior of the individual solutions to avoid excessive bloating (see sub section 3.2.4).
+It should be noted that the configuration of both GP algorithms did include two mechanisms that place limitations on the growth behavior of the individual solutions to avoid excessive bloating (see sub section 3.2.4).
 
@@
 
-The objective of this research project has been to answer the question if the choice between two commonly used parent selection operators, tournament selection and $\epsilon$-lexicase selection, significantly influences the generalization behavior of GP for symbolic regression tasks.
+The objective of this research project is to answer the question if the choice between two commonly used parent selection operators, tournament selection and $\epsilon$-lexicase selection, significantly influences the generalization behavior of GP for symbolic regression tasks.
 
-The results obtained from my experiment did not demonstrate any statistically significant differences in generalization, both GP systems achieve a slightly lower mean error if evaluated for the training data sets than if evaluated for the unseen fitness cases of the testing data set.
-The mean differences between testing and training errors did not show statistical significance differences for both algorithms (summarized in table 9).
+The results obtained from my experiment do not demonstrate any statistically significant differences in generalization.
+Both GP systems achieve a slightly lower mean error if evaluated for the training data sets than if evaluated for the unseen fitness cases of the testing data set.
+The mean differences between testing and training errors do not show statistically significant differences for neither algorithms (summarized in table 9).
 
-To further explore differences in generalization behavior the mean for both algorithms training and testing error over all generations for all 50 independent runs was visualized and analyzed (see figure 2).
-A very small and probably insignificant gap did show for $\epsilon$-lexicase selection in the final phase of the evolutionary search, in the grand scheme of things this approach also did not yield any clear evidence of differences in generalization.
+To further explore differences in generalization behavior the mean for both algorithms training and testing error over all generations for all 50 independent runs is visualized and analyzed (see figure 2).
+A very small and probably insignificant gap does show for $\epsilon$-lexicase selection in the final phase of the evolutionary search.
+Comprehensively, this approach also does not yield any clear evidence of differences in generalization.
 
-A final attempt to explore possible differences between both GP systems has been to explore differences in growth behavior as an indicator for generalization.
-This idea was based on the wide spread hypothesis that increases in overall program size might be an indication of over fitting to the training data, the concept has been discussed in sub section 2.2.
-Again, based on my results I could not find statistically significant differences in the growth behavior between both selection operators.
+A final attempt to explore possible differences between both GP systems is to explore differences in growth behavior as an indicator for generalization.
+This idea is based on the wide spread hypothesis that increases in overall program size might be an indication of overfitting to the training data.
+This concept has been discussed in sub section 2.2.
+The results do not show statistically significant differences in the growth behavior between both selection operators.
 
-An interesting finding of my analysis is, that the overall performance of tournament selection has shown to be significantly better than that of $\epsilon$-lexicase selection for the selected symbolic regression problem.
-This finding was rather unexpected based on the research of @epsilon_lexicase_main and @https://doi.org/10.48550/arxiv.1709.05394 that primarily inspired this research project in the first place.
+An interesting finding of this analysis is, that the overall performance of tournament selection shows to be notably higher than the performance of $\epsilon$-lexicase selection for the selected symbolic regression problem.
+This finding is rather unexpected based on the research of @epsilon_lexicase_main and @https://doi.org/10.48550/arxiv.1709.05394 that is the primary source of inspiration for this research project.
 

### 2022-06-23 docs/rmd/paper.tex
autogit

@@
 initial GP literature by Koza (1992) are not using separate training and
-testing data sets which might result in over fitting and a poor overall
+testing data sets which might result in overfitting and a poor overall
 generalization ability of the programs that are produced. The author
@@
 availability for drugs demonstrate that GP based techniques can either
-bloat without over fitting or over fit without bloating (Silva and
+bloat without overfitting or over fit without bloating (Silva and
 Vanneschi, 2009, p. 8).
@@
 both algorithms. The specific aim of this visualization is to explore
-the mechanism of over fitting and to examine if differences between both
+the mechanism of overfitting and to examine if differences between both
 methods can be detected.
@@
 
-Some interesting observation from tables 6 and 7 can be made by
+A few interesting observations from tables 6 and 7 can be made by
 comparing the mean of the variables \(min\_training\_error\) and
@@
 fitness, measured by the MSE, of the best performing solution that has
-been found during a single run of the experiment while the variable
+been found during a single run of the experiment. The variable
 \(testing\_error\) represents the fitness of the same solution if
@@
 \(22.456\). These results indicate an overall performance benefit for
-tournament selection based GP in the experiment. It should also be
+the tournament selection based GP in the experiment. It should also be
 mentioned that the best performing model over all 50 runs, measured by
@@
 \(testing\_error\) on the basis of each algorithm, we can observe that
-both tournament and \(\epsilon\)-lexicase achieve a slightly lower error
-on the training data than on the testing data. If we compute the total
-difference between the means of \(min\_training\_error\) and
+both tournament and \(\epsilon\)-lexicase achieve a slightly lower
+fitness on the training data than on the testing data. We compute the
+total difference between the means of \(min\_training\_error\) and
 \(testing\_error\) for both algorithms tournament selection amounts to
-0.422 and \(\epsilon\)-lexicase selection to 0.443. Regarding our
-primary research question, the differences in generalization, the
-initial observations do not seem to indicate a significant difference in
-the the relative gap between the solutions performance on testing data
-and training data.
+0.422 and \(\epsilon\)-lexicase selection to 0.443. The initial
+observations do not seem to indicate a significant difference in the the
+relative gap between the solutions' performance on testing data and
+training data.
 
@@
   The differences in mean fitness between \(min\_training\_error\) and
-  \(testing\_error\) are not statistically significant for both
-  selection algorithms. If comparing the mean training error to the mean
-  testing error, the p-values only computes to 0.309 for tournament
+  \(testing\_error\) are not statistically significant for neither
+  selection algorithms. When comparing the mean training error to the
+  mean testing error, the p-values only compute to 0.309 for tournament
   selection and to 0.257 for \(\epsilon\)-lexicase selection. These
@@
 total number of 100 generations in figure 2. Figure 2 shows that both
-algorithms show little visual evidence of over fitting which would
-result in a growing gap between between \(min\_training\_error\) and
+algorithms show little visual evidence of overfitting which would result
+in a growing gap between between \(min\_training\_error\) and
 \(testing\_error\) in the final stages of the evolutionary optimization.
 For \(\epsilon\)-lexicase selection a small gap can be observed starting
-at around the 90.th generation, while the minimum training error slowly
-converges a small increase in the testing error can be observed. Based
-on the limited amount of collected data, it is questionable if this
-small sign of over fitting can be interpreted as a weak indicator for an
-inferior generalization behavior of \(\epsilon\)-lexicase in comparison
-to tournament selection.
+approximately at the 90th generation, while the minimum training error
+slowly converges a small increase in the testing error. Based on the
+limited amount of collected data, it is questionable if this small sign
+of overfitting can be interpreted as a weak indicator for an inferior
+generalization behavior of \(\epsilon\)-lexicase in comparison to
+tournament selection.
 
@@
 
-Both subplots display a very similar behavior that is common for GP, we
+Both subplots display a very similar behavior that is common for GP. We
 observe a strong trend for solutions to grow in size with each
@@
 larger size than the population average. On first impression, the
-differences between both selection operators appears to be
-insignificant. Again, I conducted a MWU test to test for statistical
-differences in the underlying distribution of sizes measured during the
-experiment.
+differences between both selection operators appear to be insignificant.
+Again, I conducted a MWU test to examine statistical differences in the
+underlying distribution of sizes measured during the experiment.
 
@@
 has no significant effect on differences in program size. It should be
-noted that the configuration of both GP algorithm did include two
+noted that the configuration of both GP algorithms did include two
 mechanisms that place limitations on the growth behavior of the
@@
 
-The objective of this research project has been to answer the question
-if the choice between two commonly used parent selection operators,
-tournament selection and \(\epsilon\)-lexicase selection, significantly
-influences the generalization behavior of GP for symbolic regression
-tasks.
-
-The results obtained from my experiment did not demonstrate any
-statistically significant differences in generalization, both GP systems
+The objective of this research project is to answer the question if the
+choice between two commonly used parent selection operators, tournament
+selection and \(\epsilon\)-lexicase selection, significantly influences
+the generalization behavior of GP for symbolic regression tasks.
+
+The results obtained from my experiment do not demonstrate any
+statistically significant differences in generalization. Both GP systems
 achieve a slightly lower mean error if evaluated for the training data
 sets than if evaluated for the unseen fitness cases of the testing data
-set. The mean differences between testing and training errors did not
-show statistical significance differences for both algorithms
+set. The mean differences between testing and training errors do not
+show statistically significant differences for neither algorithms
 (summarized in table 9).
@@
 both algorithms training and testing error over all generations for all
-50 independent runs was visualized and analyzed (see figure 2). A very
-small and probably insignificant gap did show for \(\epsilon\)-lexicase
-selection in the final phase of the evolutionary search, in the grand
-scheme of things this approach also did not yield any clear evidence of
+50 independent runs is visualized and analyzed (see figure 2). A very
+small and probably insignificant gap does show for \(\epsilon\)-lexicase
+selection in the final phase of the evolutionary search.
+Comprehensively, this approach also does not yield any clear evidence of
 differences in generalization.
@@
 A final attempt to explore possible differences between both GP systems
-has been to explore differences in growth behavior as an indicator for
-generalization. This idea was based on the wide spread hypothesis that
-increases in overall program size might be an indication of over fitting
-to the training data, the concept has been discussed in sub section 2.2.
-Again, based on my results I could not find statistically significant
-differences in the growth behavior between both selection operators.
-
-An interesting finding of my analysis is, that the overall performance
-of tournament selection has shown to be significantly better than that
+is to explore differences in growth behavior as an indicator for
+generalization. This idea is based on the wide spread hypothesis that
+increases in overall program size might be an indication of overfitting
+to the training data. This concept has been discussed in sub section
+2.2. The results do not show statistically significant differences in
+the growth behavior between both selection operators.
+
+An interesting finding of this analysis is, that the overall performance
+of tournament selection shows to be notably higher than the performance
 of \(\epsilon\)-lexicase selection for the selected symbolic regression
-problem. This finding was rather unexpected based on the research of La
-Cava, Spector and Danai (2016) and La Cava \emph{et al.} (2017) that
-primarily inspired this research project in the first place.
+problem. This finding is rather unexpected based on the research of La
+Cava, Spector and Danai (2016) and La Cava \emph{et al.} (2017) that is
+the primary source of inspiration for this research project.
 

### 2022-06-25 docs/rmd/presentation/slides.Rmd
autogit

---
title: "Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression"
subtitle: "A comparison of epsilon-lexicase Selection and Tournament Selection"
author: "Roman Hoehn"
affiliation: "Johannes Gutenberg University Mainz"
date: "2022-06-29"
output:
  beamer_presentation:
    #theme: "Darmstadt"
    colortheme: "beaver"
    fonttheme: "structurebold"
    keep_tex: true
    fig_caption: yes
    number_sections: true
    slide_level: 2
toc: yes
toc_depth: 1
bibliography: ../bib/bib/ref.bib
csl: ../bib/csl/harvard1.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction

...

## Genetic Programming

...

## Parent Selection

...

## Symbolic Regression

...

## Generalization

...

# Experimental Study

...

## Research Design

...

## Genetic Programming Configuration

...

# Results

...

# Conclusions

...

# Limitations and open Questions

...

# References

### 2022-06-25 docs/rmd/presentation/slides.tex
autogit

--- 2022-06-23 docs/rmd/paper.tex
@@
 \documentclass[
-  12pt,
-]{article}
+  ignorenonframetext,
+]{beamer}
+\usepackage{pgfpages}
+\setbeamertemplate{caption}[numbered]
+\setbeamertemplate{caption label separator}{: }
+\setbeamercolor{caption name}{fg=normal text.fg}
+\beamertemplatenavigationsymbolsempty
+% Prevent slide breaks in the middle of a paragraph
+\widowpenalties 1 10000
+\raggedbottom
+\setbeamertemplate{part page}{
+  \centering
+  \begin{beamercolorbox}[sep=16pt,center]{part title}
+    \usebeamerfont{part title}\insertpart\par
+  \end{beamercolorbox}
+}
+\setbeamertemplate{section page}{
+  \centering
+  \begin{beamercolorbox}[sep=12pt,center]{part title}
+    \usebeamerfont{section title}\insertsection\par
+  \end{beamercolorbox}
+}
+\setbeamertemplate{subsection page}{
+  \centering
+  \begin{beamercolorbox}[sep=8pt,center]{part title}
+    \usebeamerfont{subsection title}\insertsubsection\par
+  \end{beamercolorbox}
+}
+\AtBeginPart{
+  \frame{\partpage}
+}
+\AtBeginSection{
+  \ifbibliography
+  \else
+    \frame{\sectionpage}
+  \fi
+}
+\AtBeginSubsection{
+  \frame{\subsectionpage}
+}
 \usepackage{amsmath,amssymb}
@@
 \fi
+\usecolortheme{beaver}
+\usefonttheme{structurebold}
 % Use upquote if available, for straight quotes in verbatim environments
@@
 \usepackage{xcolor}
-\usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
+\newif\ifbibliography
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 \setcounter{secnumdepth}{5}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\usepackage{float}
-\usepackage[numbers]{natbib}
-\let\origfigure\figure
-\let\endorigfigure\endfigure
-\renewenvironment{figure}[1][2] {
-    \expandafter\origfigure\expandafter[H]
-} {
-    \endorigfigure
-}
-\usepackage{float}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 \ifLuaTeX
@@
 \hypersetup{
+  pdftitle={Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression},
+  pdfauthor={Roman Hoehn},
   hidelinks,
@@
 \title{Analyzing the influence of Selection on Genetic Programming's
-Generalization ability in Symbolic Regression:
-
-A comparison of \(\epsilon\)-lexicase Selection and Tournament Selection
-
---------------------------------------------------------
-
-Student: Roman Höhn
-
-Date of Birth: 1991-04-14
-
-Place of Birth: Wiesbaden, Hesse
-
-Student ID: 2712497
-
-Supervisor: David Wittenberg
-
---------------------------------------------------------
-
-Research Project for Seminar Information Systems (03.996.3299)
-
-FB 03: Chair of Business Administration and Computer Science
-
-Johannes Gutenberg University Mainz
-
-Summerterm 2022}
-\author{}
-\date{\vspace{-2.5em}Date of Submission: 2022-06-23}
+Generalization ability in Symbolic Regression}
+\subtitle{A comparison of epsilon-lexicase Selection and Tournament
+Selection}
+\author{Roman Hoehn}
+\date{2022-06-29}
 
 \begin{document}
-\maketitle
+\frame{\titlepage}
 
-\thispagestyle{empty} \newpage{}
-
-\tableofcontents
-\thispagestyle{empty}
-\newpage
-
-\listoftables
-\listoffigures
-\thispagestyle{empty}
-\newpage
-
+\begin{frame}[allowframebreaks]
+  \tableofcontents[hideallsubsections]
+\end{frame}
 \hypertarget{introduction}{%
@@
 
-Genetic programming (GP), a sub field of evolutionary computation (EC),
-is a meta heuristic that is used to search for computer programs that
-solve a given problem by simulating the process of Darwinian evolution.
-The basic principle of GP is to gradually evolve solutions by repeatedly
-selecting parent solutions from a randomized population of computer
-programs based on a fitness metric. Then, genetic operators are applied
-on the selected solutions to generate new offspring candidate solutions.
-By repeating this process over many generations, GP acts as a guided
-search for high fitness solutions throughout the decision space. A
-unique feature of GP among other evolutionary optimization procedures is
-the possibility to evolve solutions of variable length.
+\begin{frame}{Introduction}
+\ldots{}
+\end{frame}
 
-The overall performance of GP can depend strongly on the choice of its
-underlying operators, one crucial component in this is the operator for
-parent selection. Tournament Selection is a commonly used selection
-operator in EC and is the most used operator in GP systems (Fang and Li,
-2010, p. 181). A parent solution is selected by randomly sampling \(k\)
-individuals from the current population into a tournament pool and then
-the solution with the highest fitness score from the pool is selected
-(Fang and Li, 2010, p. 182). Lexicase selection has been suggested as an
-alternative to tournament selection that is not based on aggregating
-fitness scores. It samples \(n\) test cases in random order and then
-eliminates solutions from the selection pool on a per test case basis if
-they are not performing on an elite level (Helmuth, Spector and
-Matheson, 2015, p. 1).
+\begin{frame}{Genetic Programming}
+\protect\hypertarget{genetic-programming}{}
+\ldots{}
+\end{frame}
 
-Since regular Lexicase selection has been shown to perform sub optimal
-on continuous-valued optimization problems, a modified variation called
-\(\epsilon\)-lexicase selection has been suggested for symbolic
-regression by La Cava et. al (2016). Here, \(\epsilon\)-lexicase
-selection has shown itself to outperform both in overall performance
-while showing only negligible computational overhead (La Cava, Spector
-and Danai, 2016, p. 747).
+\begin{frame}{Parent Selection}
+\protect\hypertarget{parent-selection}{}
+\ldots{}
+\end{frame}
 
-The goal of symbolic regression is to find a mathematical model for an
-observed set of data points (Paris, Robilliard and Fonlupt, 2004, p.
-794). Symbolic Regression has been one of the first GP applications and
-to this day is an actively studied and highly relevant area of research
-(Poli, Langdon and McPhee, 2008, p. 114). In most symbolic regression
-problems, little to no prior knowledge about the optimal form and
-structure of the target function is available. The ability of GP to
-optimize for model structure as well as for parameters has lead to it
-being one of the most prevalent methods used in the domain of symbolic
-regression (Paris, Robilliard and Fonlupt, 2004, p. 795).
+\begin{frame}{Symbolic Regression}
+\protect\hypertarget{symbolic-regression}{}
+\ldots{}
+\end{frame}
 
-An important quality of all supervised machine learning applications,
-including GP, is the ability to not only optimize performance for the
-test cases a model is trained on but to also perform well on previously
-unseen cases, this is referred to as generalization. In most real world
-applications of symbolic regression only a small subset of labeled data
-is available for training. The aim is to produce a model that not only
-accurately predicts the provided training data but can also predict
-previously unseen cases with high precision (Gonçalves, 2016, p. 6). A
-model that is extensively optimized on training data, can become over
-fitted to the data sample which may lead to a decrease in
-generalization.
-
-This research project tries to answer the question if the usage of
-\(\epsilon\)-lexicase selection influences the generalization behavior
-of programs that are evolved using GP for symbolic regression if
-compared to programs that are evolved using traditional tournament
-selection.
-
-Section 2 will give a brief summary and discussion of topic related
-scientific research. Section 3 describes the design of the experiment as
-well as GP specific implementation details. The statistical evaluation
-of all relevant results and the summarized conclusions are given in
-section 3 and 4. Section 5 addresses limitation and open questions of
-this research project.
-
-\hypertarget{an-overview-of-the-current-state-of-research}{%
-\section{An Overview of the current state of
-research}\label{an-overview-of-the-current-state-of-research}}
-
-\hypertarget{selection}{%
-\subsection{Selection}\label{selection}}
-
-In the first description of lexicase\footnote{The alternative, more
-  descriptive name given by Spector (2012): global pool, uniform random
-  sequence, elitist lexicase parent selection} selection for GP by
-Spector (2012) it was suggested as a novel parent selection method for
-target problems that are modal in nature. The author classifies modal
-problems as ``problems that qualitatively different modes of response
-are required for inputs from different regions of the problem's
-domain''(Spector, 2012, p. 1).
-
-In a following article Lexicase selection has been proposed specifically
-for the purpose of solving so called uncompromising problems with GP.
-Helmuth, Spector and Matheson (2015) defines uncompromising problems as
-problems that require the final solution to perform optimal on each of
-the cases it is tested on, examples include symbolic regression, the
-design of digital multipliers or finding terms in finite algebras. The
-authors also provide evidence that lexicase selection can significantly
-improves GP's ability to solve some uncompromising problems if compared
-to selection methods that are based on aggregated fitness (Helmuth,
-Spector and Matheson, 2015, p. 12).
-
-Both articles argue that many of the problem domains that GP is commonly
-used for are prevalent for problems that are modal or uncompromising in
-nature. In comparison to other selection operators, populations that are
-evolved using variations of the lexicase selection operator show a very
-high degree of genetic diversity which might be a key contributor to the
-improved performance (Helmuth, Spector and Matheson, 2015, p. 1) (La
-Cava, Spector and Danai, 2016, p. 745). In theory, Lexicase based
-selection should be more prone to select solutions that are specialist
-(solutions that achieve high fitness on a smaller subset of unusual
-training cases) than solutions that are generalist (solutions that
-achieve high average fitness among all training cases).
-
-The first description of the lexicase parent selection algorithm is
-quoted verbatim below for reference (Spector, 2012, p. 4):
-
-\begin{verbatim}
-Lexicase - Parent-Selection:
-
-1. Initialize:
-  (a) Set candidates to be the entire population.
-  (b) Set cases to be a list of all of the fitness cases in random order.
-
-2. Loop:
-  (a) Set candidates to be the subset of the current candidates that have 
-      exactly the best fitness of any individual currently in candidates for 
-      the first case in cases.
-  (b) If candidates or cases contains just a single element then return
-      the first individual in candidates.
-  (c) Otherwise remove the first case from cases and go to Loop.
-\end{verbatim}
-
-More recent research by La Cava, Spector and Danai (2016) suggests
-\(\epsilon\)-lexicase selection as a modified selection operator that
-can improve overall GP performance if applied to continuous-valued
-symbolic regression tasks in comparison to tournament and standard
-lexicase selection. La Cava, Spector and Danai (2016) argues that the
-original concept behind lexicase selection as formulated by Spector
-(2012) does not fit the requirements of real-world symbolic regression
-tasks. The authors identifies the pass condition of regular lexicase
-selection as the main problem if applied to symbolic regression:
-Individual solutions can only be selected if they perform on an elite
-level but with continuous-valued, and often noisy, data it is very
-unlikely for two individuals to achieve an exact equal error for a given
-test case. This stringent pass condition can result in filtering out too
-many individuals during the selection process (La Cava, Spector and
-Danai, 2016, p. 742). \(\epsilon\)-lexicase adresses this problem by
-introducing an \(\epsilon\) parameter that specifies a range around the
-elite error, individuals that perform inside this range pass the current
-selection iteration. La Cava \emph{et al.} (2017) describes
-\(\epsilon\)-lexicase selection as a more ``relaxed version of lexicase
-selection''. Different methods to configure the \(\epsilon\) Parameter
-have been explored, for the limited scope of this project I will focus
-on the most promising implementation of \(\epsilon\)-lexicase selection
-that is automatically adjusted on the basis of the median absolute
-deviation of errors inside the selection pool (La Cava, Spector and
-Danai, 2016, p. 742) (La Cava \emph{et al.}, 2017, p. 6).
-
-The performance increase of \(\epsilon\)-lexicase selection for symbolic
-regression problems has been demonstrated and reported for many
-benchmark problems which led to widespread adaption of it in symbolic
-regression applications (La Cava, Spector and Danai, 2016, pp.
-744--745).
-
-In comparison, a traditional GP selection method like tournament
-selection most commonly computes the fitness of a program as the mean of
-its error for each individual fitness case. One downside to this
-approach is that the total amount of information is reduced from a wide
-range of individual errors to a single metric. Helmuth, Spector and
-Matheson (2015) suspects that this loss in overall information provided
-to GP might reduce overall performance especially if applied to the
-class of uncompromising problems that require the solution to perform
-well on a wide range of diverse cases. Since tournament selection
-selects individuals based on their average fitness across all test
-cases, the resulting solutions should be expected to be more biased
-towards being generalists instead of specialists.
-
-The basic algorithm for tournament is given below for comparison (Fang
-and Li, 2010, pp. 182--183):
-
-\begin{verbatim}
-Tournament - Parent-Selection:
-
-k = tournament size
-
-1. Sample:
-  (a) Randomly sample k individuals from the current population
-      into a tournament pool.
-2. Select:
-  (a) Compute the mean fitness for each individual inside the tournament pool
-      based on all fitness cases.
-  (b) Select individual from the tournament pool with the highest mean fitness score.
-\end{verbatim}
-
-\hypertarget{generalization}{%
-\subsection{Generalization}\label{generalization}}
-
-Generalization, the ability of a model to perform well on previously
-unseen cases, is one of the fundamental goals in most real world machine
-learning applications. O'Neill \emph{et al.} (2010) raises awareness to
-the fact, that the topic of generalization in GP has not gotten the
-attention that other machine learning-based areas attributed to it. A
-similar statement has been proposed by Kushchu (2002) in his review of
-research in generalization in GP. Both authors address the need for
-additional research regarding the topic.
-
-Kushchu (2002) criticizes that almost all applications published in the
-initial GP literature by Koza (1992) are not using separate training and
-testing data sets which might result in overfitting and a poor overall
-generalization ability of the programs that are produced. The author
-calls for a more widespread adoption of methods like generational
-sampling of new training cases or the overall separation into training
-and testing data sets to improve generalization in GP (Kushchu, 2002, p.
-10).
-
-An aspect that is suspected to negatively correlate with GP's
-generalization is overall size and in particular bloating of the
-resulting programs. Bloating can be described as a growth in the total
-size of a program that does not improve its performance in any
-meaningful way (Silva and Vanneschi, 2009, p. 1). It has been widely
-suspected that GP configurations that produce very large programs also
-have a higher tendency to specialize on difficult or unusual test cases
-which could result in lower generalization (Wang, Wagner and Rondinelli,
-2019, p. 268). The minimum description length principle (MDLP) describes
-this phenomenon by claiming that less complex models are more likely to
-perform better at generalizing than more complex models that achieve a
-comparable fitness during the training phase (O'Neill \emph{et al.},
-2010, p. 349). However MDLP should be considered with caution,
-experiments by Silva and Vanneschi (2009) on the prediction of bio
-availability for drugs demonstrate that GP based techniques can either
-bloat without overfitting or over fit without bloating (Silva and
-Vanneschi, 2009, p. 8).
-
-\hypertarget{symbolic-regression}{%
-\subsection{Symbolic Regression}\label{symbolic-regression}}
-
-The task of finding a mathematical function that fits a given set of
-data points has been one of the first applications of GP in (Koza, 1992,
-p. 238). Practical examples given by Koza (1992) included tasks such as
-the discovery of scientific laws, finding solutions to differential and
-integral equations or the programmatic compression of images.
-
-Since its inception, GP based symbolic regression has been widely used
-in practical applications ranging over a wide field of disciplines. A
-well known example of an industrial use-case has been published by
-Jordaan \emph{et al.} (2004). The authors use GP based symbolic
-regression to derive nonlinear functions that can be used to
-significantly increase the robustness of industrial sensors. This
-resulted in wide-spread adoption and a significant reduction in costs.
-
-Many variations of GP based symbolic regression have been proposed to
-improve overall performance, Wang, Wagner and Rondinelli (2019)
-summarizes the main characteristics and differences for 6 common
-variations including procedures such as geometric semantic genetic
-programming, Cartesian genetic programming or GP-based relevance vector
-machines. A large benchmarking study on symbolic regression published by
-Orzechowski, Cava and Moore (2018) compares 4 GP-based procedures to
-several state-of-the-art machine leaning techniques. The authors
-demonstrate that GP, even though more time consuming, can achieve better
-results if compared to the state-of-the-art machine learning algorithm
-gradient-boosting.
+\begin{frame}{Generalization}
+\protect\hypertarget{generalization}{}
+\ldots{}
+\end{frame}
 
 \hypertarget{experimental-study}{%
-\section{Experimental study}\label{experimental-study}}
+\section{Experimental Study}\label{experimental-study}}
 
-\hypertarget{research-design}{%
-\subsection{Research Design}\label{research-design}}
+\begin{frame}{Experimental Study}
+\ldots{}
+\end{frame}
 
-To study the influence of selection on generalization I selected a data
-set about energy efficiency in buildings that is part of the UC Irvine
-Machine Learning Repository (Dua and Graff, 2017). The data set contains
-eight individual building attributes that map to two different outcomes,
-heating and cooling load, for N=768 cases (Tsanas and Xifara, 2012).
+\begin{frame}{Research Design}
+\protect\hypertarget{research-design}{}
+\ldots{}
+\end{frame}
 
-Using two otherwise identical GP systems, one deploying tournament
-selection and the other \(\epsilon\)-lexicase selection, the objective
-is to find a computer program that best predicts the outcome variable
-heating load \((Y1)\)\footnote{The symbolic regression is performed on
-  one of the two provided outcome variables, the variable Cooling Load
-  will be excluded.} of the buildings using a subset of the eight
-building attributes \((X1,..,X8)\) for its input. The specific meaning
-of all attributes inside the data set are described in table 1.
-
-\begin{table}[!h]
-
-\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating data set}
-\centering
-\begin{tabular}[t]{l|l}
-\hline
-\textbf{Variable} & \textbf{Description}\\
-\hline
-X1 & Relative Compactness\\
-\hline
-X2 & Surface Area\\
-\hline
-X3 & Wall Area\\
-\hline
-X4 & Roof Area\\
-\hline
-X5 & Overall Height\\
-\hline
-X6 & Orientation\\
-\hline
-X7 & Glazing Area\\
-\hline
-X8 & Glazing Area Distribution\\
-\hline
-y1 & Heating Load\\
-\hline
-y2 & Cooling Load\\
-\hline
-\end{tabular}
-\end{table}
-
-To measure the generalization ability of each model the data set will be
-randomly split in half, resulting in a training and testing data set
-each containing 384 individual cases. Each model will be evolved by
-traditional GP using only the fitness cases present in the training data
-set. For each generation the highest fitness model of the current
-population will be tested with the previously unseen fitness cases that
-are part of the testing data set. For each run of the experiment
-statistics will be collected on fitness which will form the basis of my
-further statistical analysis. Additional statistics will be collected
-for the average length of the population and the length of the elite
-program for each generation to explore differences in their distribution
-and possible correlations to generalization.
-
-Since GP is a stochastic optimization algorithm, the basic experiment
-will be run for a total of 50 times to ensure a fair and meaningful
-comparison based on a large number of runs for both algorithms. For each
-run of the experiment the data set will be randomly split in half as
-described, both models are then trained and tested using the exact same
-set of fitness cases.
-
-The statistical analysis of the collected data will first focus on
-examining the question if, on average, the usage of
-\(\epsilon\)-lexicase selection will result in models that perform
-significantly different than models that are evolved using tournament
-selection:
-
-\begin{quote}
-\(H0_{1}\): The distribution underlying the samples of training/testing
-errors produced by torunament selection is the same as the distribution
-underlying samples of training/testing errors produced by
-\(\epsilon\)-lexicase selection
-\end{quote}
-
-The next question of interest is, to examine if statistical significant
-differences between the mean errors of training and testing data exist
-for both selection operators:
-
-\begin{quote}
-\(H0_{2}\): The distribution underlying the samples of training errors
-produced by \(\epsilon\)-lexicase/torunament selection is the same as
-the distribution underlying samples of testing errors produced by
-\(\epsilon\)-lexicase/tournament selection
-\end{quote}
-
-To gather further insight into the differences between both GP systems,
-an additional test will be performed on the question if differences in
-the total size of the resulting programs exist:
-
-\begin{quote}
-\(H0_{3}\): No differences in program size exist between the
-distribution underlying the samples produced by \(\epsilon\)-lexicase
-and the distribution underlying samples of tournament selection
-\end{quote}
-
-All hypothesis will be tested using a level of significance of
-\(\alpha=0.05\).
-
-To further examine the difference in generalization behavior, the mean
-testing and training errors over each generation will be visualized for
-both algorithms. The specific aim of this visualization is to explore
-the mechanism of overfitting and to examine if differences between both
-methods can be detected.
-
-All GP experiments will be implemented by using the python programming
-language in conjunction with DEAP, a framework for distributed
-evolutionary algorithms that implements various tools and algorithms for
-genetic programming (Fortin \emph{et al.}, 2012). A repository that
-contains the full source code, tables, plots and other materials
-represented in this thesis is available at github.com\footnote{\url{https://github.com/roman91DE/intelligent_information_systems_research_project}}.
-
-\hypertarget{genetic-programming-configuration}{%
-\subsection{Genetic Programming
-Configuration}\label{genetic-programming-configuration}}
-
-\hypertarget{evolutionary-parameters}{%
-\subsubsection{Evolutionary parameters}\label{evolutionary-parameters}}
-
-The basic evolutionary parameters for both systems are presented in
-table 2. Tournament selection is used with a default tournament size of
-\(3\) individuals while the \(\epsilon\) parameter in lexicase selection
-is selected automatically as previously mentioned in subsection 3.1.
-
-\begin{table}[!h]
-
-\caption{\label{tab:unnamed-chunk-2}Evolutionary Parameters}
-\centering
-\begin{tabular}[t]{l|l}
-\hline
-\textbf{Parameter} & \textbf{Value}\\
-\hline
-Population Size & 500\\
-\hline
-Number of Generations & 100\\
-\hline
-Mutation Rate & 20\%\\
-\hline
-Crossover Rate & 80\%\\
-\hline
-Tournament Size & 3\\
-\hline
-Epsilon selection & automatic\\
-\hline
-Elite Size & 0\\
-\hline
-\end{tabular}
-\end{table}
-
-\hypertarget{fitness-evaluation}{%
-\subsubsection{Fitness Evaluation}\label{fitness-evaluation}}
-
-The fitness \(f\) for each model will be based on the mean squared error
-(MSE) over all fitness cases for prediction and the empirically measured
-values as described by eq.1.
-
-\begin{equation}
-\tag{eq. 1}
-MSE = \frac{1}{n} * \sum_{i=1}^{n} (Y_i - \hat{Y_1})^2
-\end{equation}
-
-where:
-
-\begin{itemize}
-\tightlist
-\item
-  \(n\): total number of test cases
-\item
-  \(Y_i\): empirical value for case \(i\)
-\item
-  \(\hat{Y_1}\): predicted value for case \(i\).
-\end{itemize}
-
-The resulting fitness function \(f\) for an individual program \(i\) is
-described by eq 2:
-
-\begin{equation}
-\tag{eq. 2}
- f(i, \tau ) = \frac{1}{N} * \sum_{t \epsilon \tau} (y_t - y\hat{}_t(i, x_t))^2 
-\end{equation}
-
-where\footnote{Naming and notation was adopted from La Cava, Spector and
-  Danai (2016)}:
-
-\begin{itemize}
-\tightlist
-\item
-  \(\tau\): the set of \(N\) fitness cases
-\item
-  \(y_t\): empirical value of the target for case \(t\)
-\item
-  \(y\hat{}_t(i, x_t)\): predicted value for the target for case \(t\)
-  by running the program \(i\) with the total set of input variables
-  \(x_t\)
-\end{itemize}
-
-\hypertarget{primitive-set}{%
-\subsubsection{Primitive Set}\label{primitive-set}}
-
-The primitive set consists of the terminals listed in table 3 and the
-functions listed in table 4. To avoid run time errors I implemented
-protected version of the operators for division, natural logarithm and
-square root (Koza, 1992, pp. 82--83). As suggested by Koza (1992) I also
-included ephemeral constants to the set of terminals to provide the
-evolutionary search the opportunity to explore and include randomly
-generated constants.
-
-\begin{table}[!h]
-
-\caption{\label{tab:unnamed-chunk-3}Terminals}
-\centering
-\begin{tabular}[t]{l|l}
-\hline
-\textbf{Terminal} & \textbf{Description}\\
-\hline
-X1 & Relative Compactness\\
-\hline
-X2 & Surface Area\\
-\hline
-X3 & Wall Area\\
-\hline
-X4 & Roof Area\\
-\hline
-X5 & Overall Height\\
-\hline
-X6 & Orientation\\
-\hline
-X7 & Glazing Area\\
-\hline
-X8 & Glazing Area Distribution\\
-\hline
-random\_int & Ephemeral Constant (integer)\\
-\hline
-random\_float & Ephemeral Constant(float)\\
-\hline
-\end{tabular}
-\end{table}
-
-\begin{table}[!h]
-
-\caption{\label{tab:unnamed-chunk-4}Functions}
-\centering
-\begin{tabular}[t]{l|r}
-\hline
-\textbf{Function} & \textbf{Arity}\\
-\hline
-Addition & 2\\
-\hline
-Subtraction & 2\\
-\hline
-Multiplication & 2\\
-\hline
-Negation & 1\\
-\hline
-Sine & 1\\
-\hline
-Cosine & 1\\
-\hline
-Protected Division & 2\\
-\hline
-Protected Natural Logarithm & 1\\
-\hline
-Protected Square Root & 1\\
-\hline
-\end{tabular}
-\end{table}
-
-\hypertarget{genetic-operators}{%
-\subsubsection{Genetic Operators}\label{genetic-operators}}
-
-GP operators are represented in table 5. Both genetic operators use a
-static limit to control for the height of the resulting trees (Koza,
-1992, p. 104). Individual programs are initialized by using the ramped
-half-and-half method, 50\% of the population are created by using the
-Growth algorithm and the remaining 50\% are created by using the Full
-algorithm (Koza, 1992, p. 93).
-
-\begin{table}[!h]
-
-\caption{\label{tab:unnamed-chunk-5}GP Operators}
-\centering
-\begin{tabular}[t]{l|l|r}
-\hline
-\textbf{Operator} & \textbf{Implementation} & \textbf{Static.Height.Limit}\\
-\hline
-Initilization & Ramped Half/Half & 2\\
-\hline
-Crossover & One Point Crossover & 17\\
-\hline
-Mutation & Uniform Mutation & 17\\
-\hline
-\end{tabular}
-\end{table}
-
-The crossover operator implemented by DEAP randomly selects a crossover
-point in each individual and exchanges each sub tree with the point as
-root between each individual (Fortin \emph{et al.}, 2012). Mutation also
-randomly selects a point in the tree individual, it then replaces the
-sub tree below that point as a root by the expression generated using
-the full grow initialization method (Fortin \emph{et al.}, 2012).
+\begin{frame}{Genetic Programming Configuration}
+\protect\hypertarget{genetic-programming-configuration}{}
+\ldots{}
+\end{frame}
 
@@
 
-Tables 6 and 7 summarize the results for all fitness scores collected
-over 50 total runs of the experiment\footnote{All floating point numbers
-  in tables have been rounded to 3 decimal places if not stated
-  otherwise}.
+\begin{frame}{Results}
+\ldots{}
+\end{frame}
 
-\begin{table}[!h]
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
 
-\caption{\label{tab:unnamed-chunk-6}Summary - Tournament Selection}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lrrrrrrrrrr}
-\toprule
-\textbf{X} & \textbf{gen} & \textbf{nevals} & \textbf{mean\_training\_error} & \textbf{std\_training\_error} & \textbf{min\_training\_error} & \textbf{max\_training\_error} & \textbf{testing\_error} & \textbf{std\_testing\_error} & \textbf{avg\_size} & \textbf{elite\_size}\\
-\midrule
-count & 101.0 & 101.000 & 1.010000e+02 & 1.010000e+02 & 101.000 & 1.010000e+02 & 100.000 & 100.000 & 100.000 & 100.000\\
-mean & 50.0 & 420.742 & 2.092614e+18 & 3.305357e+20 & 16.483 & 1.046309e+21 & 16.905 & 28.858 & 55.674 & 62.070\\
-std & 29.3 & 8.115 & 2.098028e+19 & 3.313954e+21 & 12.514 & 1.049016e+22 & 11.117 & 15.732 & 34.504 & 36.850\\
-min & 0.0 & 417.300 & 2.582030e+08 & 1.463390e+10 & 7.066 & 1.193720e+11 & 7.847 & 16.217 & 3.331 & 4.340\\
-25\% & 25.0 & 418.780 & 2.151894e+10 & 3.044239e+12 & 8.479 & 1.070232e+13 & 9.350 & 18.344 & 25.133 & 28.195\\
-\addlinespace
-50\% & 50.0 & 420.000 & 3.494635e+11 & 5.511534e+13 & 11.918 & 1.747285e+14 & 12.874 & 22.715 & 57.299 & 66.400\\
-75\% & 75.0 & 421.000 & 4.892936e+12 & 6.455802e+14 & 18.773 & 2.238973e+15 & 19.685 & 32.750 & 87.852 & 96.625\\
-max & 100.0 & 500.000 & 2.108540e+20 & 3.330558e+22 & 77.243 & 1.054272e+23 & 62.693 & 95.282 & 109.761 & 115.340\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\begin{table}[!h]
-
-\caption{\label{tab:unnamed-chunk-7}Summary - Epsilon-Lexicase Selection}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lrrrrrrrrrr}
-\toprule
-\textbf{X} & \textbf{gen} & \textbf{nevals} & \textbf{mean\_training\_error} & \textbf{std\_training\_error} & \textbf{min\_training\_error} & \textbf{max\_training\_error} & \textbf{testing\_error} & \textbf{std\_testing\_error} & \textbf{avg\_size} & \textbf{elite\_size}\\
-\midrule
-count & 101.0 & 101.000 & 1.010000e+02 & 1.010000e+02 & 101.000 & 1.010000e+02 & 100.000 & 100.000 & 100.000 & 100.000\\
-mean & 50.0 & 420.781 & 7.533781e+12 & 1.186094e+15 & 22.013 & 3.762638e+15 & 22.456 & 39.048 & 58.416 & 63.819\\
-std & 29.3 & 8.109 & 5.900311e+13 & 9.313627e+15 & 16.670 & 2.950203e+16 & 15.904 & 19.225 & 34.515 & 36.356\\
-min & 0.0 & 416.240 & 4.660413e+05 & 3.314323e+07 & 9.680 & 1.539125e+08 & 11.129 & 21.487 & 3.219 & 4.280\\
-25\% & 25.0 & 418.900 & 2.528193e+08 & 2.727451e+10 & 11.316 & 1.192373e+11 & 12.089 & 24.860 & 28.233 & 29.615\\
-\addlinespace
-50\% & 50.0 & 420.040 & 2.196352e+09 & 2.647194e+11 & 14.478 & 1.098068e+12 & 15.223 & 30.057 & 64.689 & 71.440\\
-75\% & 75.0 & 421.020 & 2.027020e+11 & 3.198174e+13 & 23.511 & 1.012976e+14 & 23.305 & 51.081 & 87.394 & 94.910\\
-max & 100.0 & 500.000 & 5.788911e+14 & 9.137151e+16 & 74.622 & 2.894465e+17 & 77.518 & 110.244 & 108.785 & 115.660\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-A few interesting observations from tables 6 and 7 can be made by
-comparing the mean of the variables \(min\_training\_error\) and
-\(testing\_error\). Variable \(min\_training\_error\) represents the
-fitness, measured by the MSE, of the best performing solution that has
-been found during a single run of the experiment. The variable
-\(testing\_error\) represents the fitness of the same solution if
-evaluated for the unknown fitness cases inside the testing data set.
-Tournament selection based GP achieves a mean \(min\_training\_error\)
-of \(16.483\) and a mean \(testing\_error\) of \(16.905\) while the GP
-system using \(\epsilon\)-lexicase selection can only achieve a mean
-\(min\_training\_error\) of \(22.013\) and a mean \(testing\_error\) of
-\(22.456\). These results indicate an overall performance benefit for
-the tournament selection based GP in the experiment. It should also be
-mentioned that the best performing model over all 50 runs, measured by
-the minimum of variables \(min\_training\_error\) and
-\(testing\_error\), has been achieved by tournament selection with an
-MSE of 7.066 for the training data and 7.847 for the testing data.
-
-If we compare the differences of \(min\_training\_error\) and
-\(testing\_error\) on the basis of each algorithm, we can observe that
-both tournament and \(\epsilon\)-lexicase achieve a slightly lower
-fitness on the training data than on the testing data. We compute the
-total difference between the means of \(min\_training\_error\) and
-\(testing\_error\) for both algorithms tournament selection amounts to
-0.422 and \(\epsilon\)-lexicase selection to 0.443. The initial
-observations do not seem to indicate a significant difference in the the
-relative gap between the solutions' performance on testing data and
-training data.
-
-Figure 1 visualizes the distribution of \(min\_training\_error\) and
-\(testing\_error\) for both selection operators over the total 50 runs
-using a box plot and summarizes the initial observations from tables 6
-and 7 described above.
-
-\begin{figure}
-\centering
-\includegraphics{./plots/mean_error_boxplot_all.png}
-\caption{Distribution of Errors}
-\end{figure}
-
-The four samples from figure 1 have been tested for normal distribution
-by computing the D'Agostino and Pearson test (D'Agostino, 1971)
-(D'Agostino and Pearson, 1973) to determine if they satisfy the
-requirements of the students t-test. The results are summarized in table
-8. Since not all samples test positive for a normal distribution a
-Mann-Whitney U (MWU) rank sum test will be performed to test for
-statistical significance between the four variables
-\(tournament\_min\_training\_error\), \(tournament\_testing\_error\),
-\(e-lexicase\_min\_training\_error\) and \(e-lexicase\_testing\_error\).
-
-\begin{table}[!h]
-
-\caption{\label{tab:unnamed-chunk-8}Normal Distribution Tests}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lrrrl}
-\toprule
-\textbf{sample} & \textbf{statistic} & \textbf{p.value} & \textbf{alpha} & \textbf{normal\_distributed}\\
-\midrule
-tournament\_min\_training\_error & 36.146 & 0.000 & 0.05 & False\\
-e-lexicase\_min\_testing\_error & 1.575 & 0.455 & 0.05 & True\\
-tournament\_testing\_error & 32.503 & 0.000 & 0.05 & False\\
-e-lexicase\_testing\_error & 59.135 & 0.000 & 0.05 & False\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-Table 9 displays the resulting p-values from computing the MWU test in a
-matrix format.
-
-\begin{table}[!h]
-
-\caption{\label{tab:unnamed-chunk-9}Mean Error - P-Values}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lrrrr}
-\toprule
-\textbf{ } & \textbf{tournament\_training\_errors} & \textbf{tournament\_testing\_errors} & \textbf{elexicase\_training\_errors} & \textbf{elexicase\_testing\_errors}\\
-\midrule
-tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
-tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
-elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
-elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-The results show two major findings:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  The differences in mean fitness between tournament selection and
-  \(\epsilon\)-lexicase selection are highly statistical significant
-  even if considering an \(\alpha\) of 0.01. For the given symbolic
-  regression task tournament selection achieves a better mean
-  performance on both the training as well as the testing data set than
-  \(\epsilon\)-lexicase selection. These findings contradict the
-  hypothesis \(H0_{1}\) described in subsection 3.1 and are contrary to
-  some of the research that was summarized in subsections 2.1 and
-  2.3.(La Cava, Spector and Danai, 2016) (La Cava \emph{et al.}, 2017).
-\item
-  The differences in mean fitness between \(min\_training\_error\) and
-  \(testing\_error\) are not statistically significant for neither
-  selection algorithms. When comparing the mean training error to the
-  mean testing error, the p-values only compute to 0.309 for tournament
-  selection and to 0.257 for \(\epsilon\)-lexicase selection. These
-  findings fail to reject \(H0_{2}\), based on the obtained results we
-  can not find evidence that both algorithms demonstrate a significantly
-  different performance between the training data and the previously
-  unseen testing data.
-\end{enumerate}
-
-To further examine the generalization behavior of both selection
-operators I plotted the mean fitness scores for both algorithms over the
-total number of 100 generations in figure 2. Figure 2 shows that both
-algorithms show little visual evidence of overfitting which would result
-in a growing gap between between \(min\_training\_error\) and
-\(testing\_error\) in the final stages of the evolutionary optimization.
-For \(\epsilon\)-lexicase selection a small gap can be observed starting
-approximately at the 90th generation, while the minimum training error
-slowly converges a small increase in the testing error. Based on the
-limited amount of collected data, it is questionable if this small sign
-of overfitting can be interpreted as a weak indicator for an inferior
-generalization behavior of \(\epsilon\)-lexicase in comparison to
-tournament selection.
-
-Figure 2 also shows a visual representation of the described overall
-performance benefit of using tournament selection instead of
-\(\epsilon\)-lexicase selection for the chosen symbolic regression task.
-It can be observed that at all stages of the evolution tournament
-selection based GP achieves a lower mean error for both data sets.
-
-\begin{figure}
-\centering
-\includegraphics{./plots/mean_error_combined.png}
-\caption{Mean Errors}
-\end{figure}
-
-In a final step I analyze the growth behavior for both GP systems to
-determine if differences in program size can be observed based on
-selection operators. Figure 3 visualizes the average size of individuals
-inside the population as well as the average size of the best performing
-individual for each generation and both GP systems.
-
-\begin{figure}
-\centering
-\includegraphics{./plots/size_subplotted.png}
-\caption{Mean Size}
-\end{figure}
-
-Both subplots display a very similar behavior that is common for GP. We
-observe a strong trend for solutions to grow in size with each
-additional generation that is concluded. Another observation that can be
-made is that the best solutions of each generation almost always have a
-larger size than the population average. On first impression, the
-differences between both selection operators appear to be insignificant.
-Again, I conducted a MWU test to examine statistical differences in the
-underlying distribution of sizes measured during the experiment.
-
-The resulting p-values of the MWU test are summarized in table 10.
-
-\begin{table}[!h]
-
-\caption{\label{tab:unnamed-chunk-10}Mean Size - P-Values}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lrrrr}
-\toprule
-\textbf{ } & \textbf{tournament\_elite\_size} & \textbf{elexicase\_elite\_size} & \textbf{tournament\_avg\_size} & \textbf{elexicase\_avg\_size}\\
-\midrule
-tournament\_elite\_size & 1.000 & 0.858 & 0.533 & 0.652\\
-elexicase\_elite\_size & 0.858 & 1.000 & 0.764 & 0.682\\
-tournament\_avg\_size & 0.533 & 0.764 & 1.000 & 0.992\\
-elexicase\_avg\_size & 0.652 & 0.682 & 0.992 & 1.000\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-None of the differences in mean size are statistically significant.
-Again, based on the results obtained in my experiment, the evidence
-produced fails to reject the null hypothesis \(H0_{3}\) that selection
-has no significant effect on differences in program size. It should be
-noted that the configuration of both GP algorithms did include two
-mechanisms that place limitations on the growth behavior of the
-individual solutions to avoid excessive bloating (see sub section
-3.2.4).
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-The objective of this research project is to answer the question if the
-choice between two commonly used parent selection operators, tournament
-selection and \(\epsilon\)-lexicase selection, significantly influences
-the generalization behavior of GP for symbolic regression tasks.
-
-The results obtained from my experiment do not demonstrate any
-statistically significant differences in generalization. Both GP systems
-achieve a slightly lower mean error if evaluated for the training data
-sets than if evaluated for the unseen fitness cases of the testing data
-set. The mean differences between testing and training errors do not
-show statistically significant differences for neither algorithms
-(summarized in table 9).
-
-To further explore differences in generalization behavior the mean for
-both algorithms training and testing error over all generations for all
-50 independent runs is visualized and analyzed (see figure 2). A very
-small and probably insignificant gap does show for \(\epsilon\)-lexicase
-selection in the final phase of the evolutionary search.
-Comprehensively, this approach also does not yield any clear evidence of
-differences in generalization.
-
-A final attempt to explore possible differences between both GP systems
-is to explore differences in growth behavior as an indicator for
-generalization. This idea is based on the wide spread hypothesis that
-increases in overall program size might be an indication of overfitting
-to the training data. This concept has been discussed in sub section
-2.2. The results do not show statistically significant differences in
-the growth behavior between both selection operators.
-
-An interesting finding of this analysis is, that the overall performance
-of tournament selection shows to be notably higher than the performance
-of \(\epsilon\)-lexicase selection for the selected symbolic regression
-problem. This finding is rather unexpected based on the research of La
-Cava, Spector and Danai (2016) and La Cava \emph{et al.} (2017) that is
-the primary source of inspiration for this research project.
+\begin{frame}{Conclusions}
+\ldots{}
+\end{frame}
 
@@
 \section{Limitations and open
-questions}\label{limitations-and-open-questions}}
+Questions}\label{limitations-and-open-questions}}
 
-I faced two major difficulties in the preparation and evaluation of this
-research project:
+\begin{frame}{Limitations and open Questions}
+\ldots{}
+\end{frame}
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Configuration of evolutionary parameters
-\item
-  Limited computational resources
-\end{enumerate}
-
-The combination of both issues might have resulted in a significant
-reduction in overall robustness of the results represented in sections 5
-and 6. The performance of all evolutionary algorithms, including GP, can
-be highly dependent on a large number of parameters and implementation
-details. Although the series of experiments conducted in this project
-are comparably simple in nature, the results achieved might still be
-highly influenced by the GP configuration detailed in subsection 4.2.
-
-To gain further insight into the differences in generalization behavior
-and to address the research question with a higher level of certainty,
-the experiment I conducted should be repeated for different GP
-configurations. Some exemplary parameters that could drastically
-influence the results of this experiment include the population size,
-the total number of generations, the inclusion of elitism or the
-computation of \(\epsilon\) in lexicase selection. Other factors whose
-influence on generalization behavior could be important include
-different methods to compute an individuals fitness (e.g.~mean absolute
-error or root mean squared error), testing different variations of the
-genetic operators for mutation and crossover and experimenting with
-different variations of the training:testing data set split size.
-
-Another obvious weakness of this project is that the experiment is only
-based on a single symbolic regression application, the prediction of the
-heating load of buildings based on a relatively small data set. To gain
-further trust in the obtained results, the experiment should be repeated
-for other symbolic regression tasks and data sets. It remains unknown if
-the generalization behavior of both selection operators might be highly
-problem specific and dependent on the data set used for training and
-testing.
-
-All limitations and shortcomings described above are also connected to
-the second limitation of this research project, the limited amount of
-computational resources. The total computation time consumed to run the
-current experiments (50 runs, 2 algorithms, 100 generations, 500
-individuals) has already been close to 24 hours on a modern, arm-based
-Apple-M1 system. Even though the overall computation time could probably
-be reduced by further optimization of the source code, e.g.~using tools
-for parallel computation, the resources necessary to repeat the whole
-experiment for different parameter configurations would certainly be
-beyond the scope of this research project.
-
-\newpage
-
-\hypertarget{I}{%
-\section*{I - References}\label{I}}
-\addcontentsline{toc}{section}{I - References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-DAgostino1971AnOT}{}}%
-D'Agostino, R.B. (1971) {``An omnibus test of normality for moderate and
-large size samples,''} \emph{Biometrika}, 58, pp. 341--348.
-
-\leavevmode\vadjust pre{\hypertarget{ref-DAgostino1973TestsFD}{}}%
-D'Agostino, R.B. and Pearson, E.S. (1973) {``Tests for departure from
-normality,''} in.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Dua:2019}{}}%
-Dua, D. and Graff, C. (2017) {``{UCI} machine learning repository.''}
-University of California, Irvine, School of Information; Computer
-Sciences. Available at: \url{http://archive.ics.uci.edu/ml}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-10.1007ux2f978-3-642-16493-4_19}{}}%
-Fang, Y. and Li, J. (2010) {``A review of tournament selection in
-genetic programming,''} in Cai, Z. et al. (eds.) \emph{Advances in
-computation and intelligence}. Berlin, Heidelberg: Springer Berlin
-Heidelberg, pp. 181--192.
-
-\leavevmode\vadjust pre{\hypertarget{ref-DEAP_JMLR2012}{}}%
-Fortin, F.-A. \emph{et al.} (2012) {``{DEAP}: Evolutionary algorithms
-made easy,''} \emph{Journal of Machine Learning Research}, 13, pp.
-2171--2175.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Gonalves2016AnEO}{}}%
-Gonçalves, I. (2016) {``An exploration of generalization and overfitting
-in genetic programming: Standard and geometric semantic approaches,''}
-in.
-
-\leavevmode\vadjust pre{\hypertarget{ref-6920034}{}}%
-Helmuth, T., Spector, L. and Matheson, J. (2015) {``Solving
-uncompromising problems with lexicase selection,''} \emph{IEEE
-Transactions on Evolutionary Computation}, 19(5), pp. 630--643.
-doi:\href{https://doi.org/10.1109/TEVC.2014.2362729}{10.1109/TEVC.2014.2362729}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-soft_sensor_gp}{}}%
-Jordaan, E. \emph{et al.} (2004) {``Robust inferential sensors based on
-ensemble of predictors generated by genetic programming,''} in Yao, X.
-et al. (eds.) \emph{Parallel problem solving from nature - PPSN VIII}.
-Berlin, Heidelberg: Springer Berlin Heidelberg, pp. 522--531.
-
-\leavevmode\vadjust pre{\hypertarget{ref-koza_main}{}}%
-Koza, J.R. (1992) \emph{Genetic programming: On the programming of
-computers by means of natural selection}. Cambridge, MA, USA: MIT Press.
-Available at: \url{http://mitpress.mit.edu/books/genetic-programming}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-generalisation_in_gp}{}}%
-Kushchu, I. (2002) {``An evaluation of EvolutionaryGeneralisation in
-genetic programming,''} \emph{Artificial Intelligence Review - AIR}, 18,
-pp. 3--14.
-doi:\href{https://doi.org/10.1023/A:1016379201230}{10.1023/A:1016379201230}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-https:ux2fux2fdoi.orgux2f10.48550ux2farxiv.1709.05394}{}}%
-La Cava, W. \emph{et al.} (2017) {``A probabilistic and multi-objective
-analysis of lexicase selection and epsilon-lexicase selection.''} arXiv.
-doi:\href{https://doi.org/10.48550/ARXIV.1709.05394}{10.48550/ARXIV.1709.05394}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-epsilon_lexicase_main}{}}%
-La Cava, W., Spector, L. and Danai, K. (2016) {``Epsilon-lexicase
-selection for regression,''} in \emph{Proceedings of the genetic and
-evolutionary computation conference 2016}. New York, NY, USA:
-Association for Computing Machinery (GECCO '16), pp. 741--748.
-doi:\href{https://doi.org/10.1145/2908812.2908898}{10.1145/2908812.2908898}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-open_issues_gp}{}}%
-O'Neill, M. \emph{et al.} (2010) {``Open issues in genetic
-programming,''} \emph{Genetic Programming and Evolvable Machines}, 11,
-pp. 339--363.
-doi:\href{https://doi.org/10.1007/s10710-010-9113-2}{10.1007/s10710-010-9113-2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Orzechowski_2018}{}}%
-Orzechowski, P., Cava, W.L. and Moore, J.H. (2018) {``Where are we
-now?''} in \emph{Proceedings of the genetic and evolutionary computation
-conference}. {ACM}.
-doi:\href{https://doi.org/10.1145/3205455.3205539}{10.1145/3205455.3205539}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-10.1007ux2f978-3-540-24621-3_22}{}}%
-Paris, G., Robilliard, D. and Fonlupt, C. (2004) {``Exploring
-overfitting in genetic programming,''} in Liardet, P. et al. (eds.)
-\emph{Artificial evolution}. Berlin, Heidelberg: Springer Berlin
-Heidelberg, pp. 267--277.
-
-\leavevmode\vadjust pre{\hypertarget{ref-poli08:fieldguide}{}}%
-Poli, R., Langdon, W.B. and McPhee, N.F. (2008) \emph{A field guide to
-genetic programming}. Published via \texttt{http://lulu.com}; freely
-available at \texttt{http://www.gp-field-guide.org.uk}. Available at:
-\url{https://digitalcommons.morris.umn.edu/cgi/viewcontent.cgi?article=1001\&context=cs_facpubs}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bloat_overfitting_gp}{}}%
-Silva, S. and Vanneschi, L. (2009) {``Operator equalisation, bloat and
-overfitting: A study on human oral bioavailability prediction,''} in
-\emph{Proceedings of the 11th Annual Genetic and Evolutionary
-Computation Conference, GECCO-2009}, pp. 1115--1122.
-doi:\href{https://doi.org/10.1145/1569901.1570051}{10.1145/1569901.1570051}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lexicase_first_desciption}{}}%
-Spector, L. (2012) {``Assessment of problem modality by differential
-performance of lexicase selection in genetic programming: A preliminary
-report.''}
-doi:\href{https://doi.org/10.1145/2330784.2330846}{10.1145/2330784.2330846}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-fe8fa39e88a040bbacba5a465c48043f}{}}%
-Tsanas, A. and Xifara, A. (2012) {``Accurate quantitative estimation of
-energy performance of residential buildings using statistical machine
-learning tools,''} \emph{Energy and buildings}, 49, pp. 560--567.
-doi:\href{https://doi.org/10.1016/j.enbuild.2012.03.003}{10.1016/j.enbuild.2012.03.003}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wang_wagner_rondinelli_2019}{}}%
-Wang, Y., Wagner, N. and Rondinelli, J.M. (2019) {``Symbolic regression
-in materials science,''} \emph{MRS Communications}, 9(3), pp. 793--805.
-doi:\href{https://doi.org/10.1557/mrc.2019.85}{10.1557/mrc.2019.85}.
-
-\end{CSLReferences}
-
-\hypertarget{II}{%
-\section*{II - Statutory Declaration}\label{II}}
-\addcontentsline{toc}{section}{II - Statutory Declaration}
-
-\includegraphics{./private/erklaerung.pdf}\\
+\hypertarget{references}{%
+\section{References}\label{references}}
 

### 2022-06-25 docs/rmd/presentation/slides.Rmd
lil progress

@@
 subtitle: "A comparison of epsilon-lexicase Selection and Tournament Selection"
-author: "Roman Hoehn"
-affiliation: "Johannes Gutenberg University Mainz"
 date: "2022-06-29"
@@
   beamer_presentation:
-    #theme: "Darmstadt"
     colortheme: "beaver"
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+library(kableExtra)
+library(magrittr)
+knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(fig.pos = "!H")
 ```
@@
 
-...
+## Research Question
+
+* Does the usage of $\epsilon$-lexicase parent selection influence the generalization behaviour of genetic programming in symbolic regression if compared to tournament selection?
 
@@
 
-...
+* A metaheuristic that searches for computer programs that solve a given problem
+* Evolutionary algorithm that simulates the process of Darwinian evolution:
+  1. Fitnessfunction: Evaluates the quality of a solution
+  1. Selection: Solutions are selected based on their individual fitness
+  2. Variation: Mutation and recombination of solutions
+* Unique Features:
+  * Evolve solutions of variable length and structure
+  * Solutions are typically represented by recursive tree structures
 
@@
 ## Generalization
+
+## Motivation
+
+- Generalization
+
+# Theoretical Foundations
+
+## Genetic Programming
+
+...
 
@@
 
+## Descriptive Statistics
+
+![Distribution of Errors](../plots/mean_error_boxplot_all.png)
+
 ...

### 2022-06-25 docs/rmd/presentation/slides.tex
lil progress

@@
 }
-\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
   \usepackage[T1]{fontenc}
@@
 \usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  pdftitle={Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
 \newif\ifbibliography
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 \setcounter{secnumdepth}{5}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdftitle={Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression},
-  pdfauthor={Roman Hoehn},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
 
@@
 Selection}
-\author{Roman Hoehn}
-\date{2022-06-29}
+\author{}
+\date{\vspace{-2.5em}2022-06-29}
 
@@
 
-\begin{frame}{Introduction}
-\ldots{}
+\begin{frame}{Research Question}
+\protect\hypertarget{research-question}{}
+\begin{itemize}
+\tightlist
+\item
+  Does the usage of \(\epsilon\)-lexicase parent selection influence the
+  generalization behaviour of genetic programming in symbolic regression
+  if compared to tournament selection?
+\end{itemize}
 \end{frame}
@@
 \protect\hypertarget{genetic-programming}{}
-\ldots{}
+\begin{itemize}
+\tightlist
+\item
+  A metaheuristic that searches for computer programs that solve a given
+  problem
+\item
+  Evolutionary algorithm that simulates the process of Darwinian
+  evolution:
+
+  \begin{enumerate}
+  \tightlist
+  \item
+    Fitnessfunction: Evaluates the quality of a solution
+  \item
+    Selection: Solutions are selected based on their individual fitness
+  \item
+    Variation: Mutation and recombination of solutions
+  \end{enumerate}
+\item
+  Unique Features:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Evolve solutions of variable length and structure
+  \item
+    Solutions are typically represented by recursive tree structures
+  \end{itemize}
+\end{itemize}
 \end{frame}
@@
 \protect\hypertarget{generalization}{}
+\end{frame}
+
+\begin{frame}{Motivation}
+\protect\hypertarget{motivation}{}
+\begin{itemize}
+\tightlist
+\item
+  Generalization
+\end{itemize}
+\end{frame}
+
+\hypertarget{theoretical-foundations}{%
+\section{Theoretical Foundations}\label{theoretical-foundations}}
+
+\begin{frame}{Genetic Programming}
+\protect\hypertarget{genetic-programming-1}{}
+\ldots{}
+
 \ldots{}
@@
 
-\begin{frame}{Results}
+\begin{frame}{Descriptive Statistics}
+\protect\hypertarget{descriptive-statistics}{}
+\begin{figure}
+\centering
+\includegraphics{../plots/mean_error_boxplot_all.png}
+\caption{Distribution of Errors}
+\end{figure}
+
 \ldots{}

### 2022-06-25 docs/rmd/presentation/slides.tex
autogit from ubuntu2204

@@
 \setcounter{secnumdepth}{5}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
 
@@
 
+\hypertarget{research-question}{%
+\section{Research Question}\label{research-question}}
+
 \begin{frame}{Research Question}
-\protect\hypertarget{research-question}{}
 \begin{itemize}
@@
   if compared to tournament selection?
+\item
 \end{itemize}
@@
 \item
+  Inventor: John R. Koza\footnote<.->{Koza (1992)}
+\item
   Evolutionary algorithm that simulates the process of Darwinian
@@
   \item
-    Fitnessfunction: Evaluates the quality of a solution
+    Population based
+  \item
+    The quality of solutions is evaluated by a fitness function
   \item
@@
 \protect\hypertarget{parent-selection}{}
-\ldots{}
+\begin{itemize}
+\tightlist
+\item
+  Operator that selects individual solutions from the population for
+  reproduction and mutation
+\item
+  Most commonly used selection operator in GP: Tournament
+  selection\footnote<.->{Fang and Li (2010), p.181}
+\item
+  Intuition: High chance for ``generalist'' solutions to be selected
+  since it is based on aggregated fitness scores
+\end{itemize}
+\end{frame}
+
+\begin{frame}{epsilon-Lexicase Selection}
+\protect\hypertarget{epsilon-lexicase-selection}{}
+\begin{itemize}
+\tightlist
+\item
+  Recent alternative: Lexicase Selection and it's variation
+  \(\epsilon\)-lexicase selection
+\item
+  Idea: Selection method for uncompromising, continous-valued symbolic
+  regression problems\footnote<.->{Helmuth, Spector and Matheson (2015),
+    p.12}
+\item
+  Increases genetic diversity inside the population\footnote<.->{Helmuth,
+    Spector and Matheson (2015), p.1}
+\item
+  Higher chance for ``specialist'' solutions to be selected since it is
+  decided on a per case basis
+\item
+  Performance increases have been demosntrated in many benchmarking
+  problems\footnote<.->{La Cava, Spector and Danai (2016), p.744-745}
+\end{itemize}
 \end{frame}
@@
 \protect\hypertarget{symbolic-regression}{}
+\begin{itemize}
+\tightlist
+\item
+  Task: Find a mathematical model that fits a given set of datapoints
+\item
+  One of the first applications of Genetic Programming introduced
+\end{itemize}
+
 \ldots{}
@@
 \end{frame}
-
-\hypertarget{theoretical-foundations}{%
-\section{Theoretical Foundations}\label{theoretical-foundations}}
 
@@
 \begin{frame}{Limitations and open Questions}
-\ldots{}
+\ldots{} \newpage
+
+\hypertarget{refs}{}
+\begin{cslreferences}
+\leavevmode\hypertarget{ref-10.1007ux2f978-3-642-16493-4_19}{}%
+Fang, Y. and Li, J. (2010) `A review of tournament selection in genetic
+programming', in Cai, Z. et al. (eds) \emph{Advances in computation and
+intelligence}. Berlin, Heidelberg: Springer Berlin Heidelberg, pp.
+181--192.
+
+\leavevmode\hypertarget{ref-6920034}{}%
+Helmuth, T., Spector, L. and Matheson, J. (2015) `Solving uncompromising
+problems with lexicase selection', \emph{IEEE Transactions on
+Evolutionary Computation}, 19(5), pp. 630--643.
+doi:\href{https://doi.org/10.1109/TEVC.2014.2362729}{10.1109/TEVC.2014.2362729}.
+
+\leavevmode\hypertarget{ref-koza_main}{}%
+Koza, J.R. (1992) \emph{Genetic programming: On the programming of
+computers by means of natural selection}. Cambridge, MA, USA: MIT Press.
+Available at: \url{http://mitpress.mit.edu/books/genetic-programming}.
+
+\leavevmode\hypertarget{ref-epsilon_lexicase_main}{}%
+La Cava, W., Spector, L. and Danai, K. (2016) `Epsilon-lexicase
+selection for regression', in \emph{Proceedings of the genetic and
+evolutionary computation conference 2016}. New York, NY, USA:
+Association for Computing Machinery (GECCO '16), pp. 741--748.
+doi:\href{https://doi.org/10.1145/2908812.2908898}{10.1145/2908812.2908898}.
+\end{cslreferences}
 \end{frame}
 
-\hypertarget{references}{%
-\section{References}\label{references}}
-
 \end{document}

### 2022-06-26 docs/rmd/presentation/slides.Rmd
autogit

--- 2022-06-25 docs/rmd/presentation/slides.Rmd
@@
   beamer_presentation:
-    #theme: "Darmstadt"
-    colortheme: "beaver"
+    theme: "Frankfurt"
+    #colortheme: "beaver"
     fonttheme: "structurebold"

### 2022-06-26 docs/rmd/presentation/slides.Rmd

--- 2022-06-25 docs/rmd/presentation/slides.Rmd
@@
 toc: yes
-toc_depth: 1
+toc_depth: 2
 bibliography: ../bib/bib/ref.bib
@@
 
-## Research Question
+# Research Question
 
@@
 * A metaheuristic that searches for computer programs that solve a given problem
+* Inventor: John R. Koza ^[@koza_main]
 * Evolutionary algorithm that simulates the process of Darwinian evolution:
-  1. Fitnessfunction: Evaluates the quality of a solution
-  1. Selection: Solutions are selected based on their individual fitness
-  2. Variation: Mutation and recombination of solutions
+  1. Population based
+  2. The quality of solutions is evaluated by a fitness function
+  3. Selection: Solutions are selected based on their individual fitness
+  4. Variation: Mutation and recombination of solutions
 * Unique Features:
@@
 
-...
+* Operator that selects individual solutions from the population for reproduction and mutation
+* Most commonly used selection operator in GP: Tournament selection^[@10.1007/978-3-642-16493-4_19, p.181]
+* Intuition: High chance for "generalist" solutions to be selected since it is based on aggregated fitness scores
+
+## epsilon-Lexicase Selection
+
+* Recent alternative: Lexicase Selection and it's variation $\epsilon$-lexicase selection
+* Idea: Selection method for uncompromising, continous-valued symbolic regression problems ^[@6920034, p.12]
+* Increases genetic diversity inside the population^[@6920034, p.1]
+* Higher chance for "specialist" solutions to be selected since it is decided on a per case basis
+* Performance increases have been demosntrated in many benchmarking problems ^[@epsilon_lexicase_main, p.744-745]
 
 ## Symbolic Regression
+
+* Task: Find a mathematical model that fits a given set of datapoints
+* One of the first applications of Genetic Programming introduced by @koza_main
+* GP, even though more time consuming, can
 
@@
 - Generalization
-
-# Theoretical Foundations
 
@@
 ...
-
-# References
+\newpage

### 2022-06-26 docs/rmd/presentation/slides.Rmd
autogit

@@
 subtitle: "A comparison of epsilon-lexicase Selection and Tournament Selection"
+author: "Roman Hoehn, B.Sc. Wirtschaftspaedagogik"
 date: "2022-06-29"
@@
 toc: yes
-toc_depth: 2
+# toc_depth: 3
 bibliography: ../bib/bib/ref.bib
@@
 
-# Research Question
+## Research Question
 
@@
 * Operator that selects individual solutions from the population for reproduction and mutation
-* Most commonly used selection operator in GP: Tournament selection^[@10.1007/978-3-642-16493-4_19, p.181]
+* Most commonly used selection operator in Genetic Programming (GP): Tournament selection^[@10.1007/978-3-642-16493-4_19, p.181]
 * Intuition: High chance for "generalist" solutions to be selected since it is based on aggregated fitness scores
@@
 * Task: Find a mathematical model that fits a given set of datapoints
-* One of the first applications of Genetic Programming introduced by @koza_main
-* GP, even though more time consuming, can
-
-...
+* One of the first applications of GP described by @koza_main
+* High relevance: GP can outperform state-of-the-art machine learning algorithms like gradient boosting ^[@Orzechowski_2018]
 
@@
 
+* The ability of a model to perform well on previously unseen fitness cases
+* Main objective in most supervised machine learning problems
+* Challenge: Avoid overfitting to training data
+
 ## Motivation
 
-- Generalization
-
-## Genetic Programming
-
-...
-
-...
+* Little attention has been paid to generalization in GP ^[@open_issues_gp, @generalisation_in_gp]
+* High practical relevance of symbolic regression in many fields, e.g. financial forecasting
 
@@
 
-...
+## Benchmark problem
 
-## Research Design
+UC Irvine Machine Learning Repository: Prediction of energy efficiency in buildings ^[@Dua:2019]
 
-...
+```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+fig.pos="H"
+
+knitr::kable(
+  read.csv("../tables/csv/variables_energy_heating.csv"),
+  digits=3,
+  caption="Overview - Energy Heating data set"
+)    %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
+```
+
+## Experiment
+
+### Single run
+
+* Total dataset ($N=768$) is randomly split into a training and testing dataset (50:50)
+* Train two models using GP with the training dataset only, one using tournament selection and the other $\epsilon$-lexicase selection
+* For each generation: Select elite model and compute its fitness for the testing dataset
+
+### Full experiment
+
+* Stochastic algorithm: Repeat the basic experiment for 50 total runs
+* Collect and aggregate results for training error, testing error and program length
 

### 2022-06-26 docs/rmd/presentation/slides.tex
autogit

@@
 }
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\usepackage{iftex}
+\ifPDFTeX
   \usepackage[T1]{fontenc}
@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \newif\ifbibliography
@@
 \setlength{\cslhangindent}{1.5em}
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression},
+  pdfauthor={Roman Hoehn, B.Sc. Wirtschaftspaedagogik},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
@@
 Selection}
-\author{}
-\date{\vspace{-2.5em}2022-06-29}
+\author{Roman Hoehn, B.Sc. Wirtschaftspaedagogik}
+\date{2022-06-29}
 
@@
 
-\hypertarget{research-question}{%
-\section{Research Question}\label{research-question}}
-
 \begin{frame}{Research Question}
+\protect\hypertarget{research-question}{}
 \begin{itemize}
@@
   if compared to tournament selection?
-\item
 \end{itemize}
@@
 \item
-  Inventor: John R. Koza\footnote<.->{Koza (1992)}
+  Inventor: John R. Koza \footnote<.->{Koza (1992)}
 \item
@@
 \item
-  Most commonly used selection operator in GP: Tournament
-  selection\footnote<.->{Fang and Li (2010), p.181}
+  Most commonly used selection operator in Genetic Programming (GP):
+  Tournament selection\footnote<.->{Fang and Li (2010), p.181}
 \item
@@
   Idea: Selection method for uncompromising, continous-valued symbolic
-  regression problems\footnote<.->{Helmuth, Spector and Matheson (2015),
-    p.12}
+  regression problems \footnote<.->{Helmuth, Spector and Matheson
+    (2015), p.12}
 \item
@@
   Performance increases have been demosntrated in many benchmarking
-  problems\footnote<.->{La Cava, Spector and Danai (2016), p.744-745}
+  problems \footnote<.->{La Cava, Spector and Danai (2016), p.744-745}
 \end{itemize}
@@
 \item
-  One of the first applications of Genetic Programming introduced
+  One of the first applications of GP described by Koza (1992)
+\item
+  High relevance: GP can outperform state-of-the-art machine learning
+  algorithms like gradient boosting \footnote<.->{Orzechowski, Cava and
+    Moore (2018)}
 \end{itemize}
-
-\ldots{}
 \end{frame}
@@
 \protect\hypertarget{generalization}{}
+\begin{itemize}
+\tightlist
+\item
+  The ability of a model to perform well on previously unseen fitness
+  cases
+\item
+  Main objective in most supervised machine learning problems
+\item
+  Challenge: Avoid overfitting to training data
+\end{itemize}
 \end{frame}
@@
 \item
-  Generalization
+  Little attention has been paid to generalization in GP \footnote<.->{O'Neill
+    \emph{et al.} (2010), Kushchu (2002)}
+\item
+  High practical relevance of symbolic regression in many fields,
+  e.g.~financial forecasting
 \end{itemize}
-\end{frame}
-
-\begin{frame}{Genetic Programming}
-\protect\hypertarget{genetic-programming-1}{}
-\ldots{}
-
-\ldots{}
 \end{frame}
@@
 
-\begin{frame}{Experimental Study}
-\ldots{}
+\begin{frame}{Benchmark problem}
+\protect\hypertarget{benchmark-problem}{}
+UC Irvine Machine Learning Repository: Prediction of energy efficiency
+in buildings \footnote<.->{Dua and Graff (2017)}
+
+\begin{table}[!h]
+
+\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating data set}
+\centering
+\begin{tabular}[t]{l|l}
+\hline
+\textbf{Variable} & \textbf{Description}\\
+\hline
+X1 & Relative Compactness\\
+\hline
+X2 & Surface Area\\
+\hline
+X3 & Wall Area\\
+\hline
+X4 & Roof Area\\
+\hline
+X5 & Overall Height\\
+\hline
+X6 & Orientation\\
+\hline
+X7 & Glazing Area\\
+\hline
+X8 & Glazing Area Distribution\\
+\hline
+y1 & Heating Load\\
+\hline
+y2 & Cooling Load\\
+\hline
+\end{tabular}
+\end{table}
 \end{frame}
 
-\begin{frame}{Research Design}
-\protect\hypertarget{research-design}{}
-\ldots{}
+\begin{frame}{Experiment}
+\protect\hypertarget{experiment}{}
+\begin{block}{Single run}
+\protect\hypertarget{single-run}{}
+\begin{itemize}
+\tightlist
+\item
+  Total dataset (\(N=768\)) is randomly split into a training and
+  testing dataset (50:50)
+\item
+  Train two models using GP with the training dataset only, one using
+  tournament selection and the other \(\epsilon\)-lexicase selection
+\item
+  For each generation: Select elite model and compute its fitness for
+  the testing dataset
+\end{itemize}
+\end{block}
+
+\begin{block}{Full experiment}
+\protect\hypertarget{full-experiment}{}
+\begin{itemize}
+\tightlist
+\item
+  Stochastic algorithm: Repeat the basic experiment for 50 total runs
+\item
+  Collect stats for: min\_training\_error, testing\_error,
+\end{itemize}
+\end{block}
 \end{frame}
@@
 \hypertarget{refs}{}
-\begin{cslreferences}
-\leavevmode\hypertarget{ref-10.1007ux2f978-3-642-16493-4_19}{}%
-Fang, Y. and Li, J. (2010) `A review of tournament selection in genetic
-programming', in Cai, Z. et al. (eds) \emph{Advances in computation and
-intelligence}. Berlin, Heidelberg: Springer Berlin Heidelberg, pp.
-181--192.
+\begin{CSLReferences}{0}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-Dua:2019}{}}%
+Dua, D. and Graff, C. (2017) {``{UCI} machine learning repository.''}
+University of California, Irvine, School of Information; Computer
+Sciences. Available at: \url{http://archive.ics.uci.edu/ml}.
 
-\leavevmode\hypertarget{ref-6920034}{}%
-Helmuth, T., Spector, L. and Matheson, J. (2015) `Solving uncompromising
-problems with lexicase selection', \emph{IEEE Transactions on
-Evolutionary Computation}, 19(5), pp. 630--643.
+\leavevmode\vadjust pre{\hypertarget{ref-10.1007ux2f978-3-642-16493-4_19}{}}%
+Fang, Y. and Li, J. (2010) {``A review of tournament selection in
+genetic programming,''} in Cai, Z. et al. (eds.) \emph{Advances in
+computation and intelligence}. Berlin, Heidelberg: Springer Berlin
+Heidelberg, pp. 181--192.
+
+\leavevmode\vadjust pre{\hypertarget{ref-6920034}{}}%
+Helmuth, T., Spector, L. and Matheson, J. (2015) {``Solving
+uncompromising problems with lexicase selection,''} \emph{IEEE
+Transactions on Evolutionary Computation}, 19(5), pp. 630--643.
 doi:\href{https://doi.org/10.1109/TEVC.2014.2362729}{10.1109/TEVC.2014.2362729}.
 
-\leavevmode\hypertarget{ref-koza_main}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-koza_main}{}}%
 Koza, J.R. (1992) \emph{Genetic programming: On the programming of
@@
 
-\leavevmode\hypertarget{ref-epsilon_lexicase_main}{}%
-La Cava, W., Spector, L. and Danai, K. (2016) `Epsilon-lexicase
-selection for regression', in \emph{Proceedings of the genetic and
+\leavevmode\vadjust pre{\hypertarget{ref-generalisation_in_gp}{}}%
+Kushchu, I. (2002) {``An evaluation of EvolutionaryGeneralisation in
+genetic programming,''} \emph{Artificial Intelligence Review - AIR}, 18,
+pp. 3--14.
+doi:\href{https://doi.org/10.1023/A:1016379201230}{10.1023/A:1016379201230}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-epsilon_lexicase_main}{}}%
+La Cava, W., Spector, L. and Danai, K. (2016) {``Epsilon-lexicase
+selection for regression,''} in \emph{Proceedings of the genetic and
 evolutionary computation conference 2016}. New York, NY, USA:
@@
 doi:\href{https://doi.org/10.1145/2908812.2908898}{10.1145/2908812.2908898}.
-\end{cslreferences}
+
+\leavevmode\vadjust pre{\hypertarget{ref-open_issues_gp}{}}%
+O'Neill, M. \emph{et al.} (2010) {``Open issues in genetic
+programming,''} \emph{Genetic Programming and Evolvable Machines}, 11,
+pp. 339--363.
+doi:\href{https://doi.org/10.1007/s10710-010-9113-2}{10.1007/s10710-010-9113-2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Orzechowski_2018}{}}%
+Orzechowski, P., Cava, W.L. and Moore, J.H. (2018) {``Where are we
+now?''} in \emph{Proceedings of the genetic and evolutionary computation
+conference}. {ACM}.
+doi:\href{https://doi.org/10.1145/3205455.3205539}{10.1145/3205455.3205539}.
+
+\end{CSLReferences}
 \end{frame}

### 2022-06-26 docs/rmd/paper.Rmd
autogit
...

@@
 ```
+\setcounter{page}{1}
 
@@
 
-> $H0_{1}$: The distribution underlying the samples of training/testing errors produced by torunament selection is the same as the distribution underlying samples of training/testing errors produced by $\epsilon$-lexicase selection
+> $H0_{1}$: The distribution underlying the samples of training/testing errors produced by tournament selection is the same as the distribution underlying samples of training/testing errors produced by $\epsilon$-lexicase selection
 
@@
 
-> $H0_{2}$: The distribution underlying the samples of training errors produced by $\epsilon$-lexicase/torunament selection is the same as the distribution underlying samples of testing errors produced by $\epsilon$-lexicase/tournament selection
+> $H0_{2}$: The distribution underlying the samples of training errors produced by $\epsilon$-lexicase/tournament selection is the same as the distribution underlying samples of testing errors produced by $\epsilon$-lexicase/tournament selection
 

### 2022-06-26 docs/rmd/presentation/slides.Rmd
autogit
...

@@
 * Higher chance for "specialist" solutions to be selected since it is decided on a per case basis
-* Performance increases have been demosntrated in many benchmarking problems ^[@epsilon_lexicase_main, p.744-745]
+* Performance increases have been demonstrated in many benchmarking problems ^[@epsilon_lexicase_main, p.744-745]
 
@@
 
+## Research Question
+
+* Does the usage of $\epsilon$-lexicase parent selection influence the generalization behaviour of genetic programming in symbolic regression if compared to tournament selection?
+
+### Hypothesis testing
+
+#### TODO: reformulate hypothesis
+1. The usage of $\epsilon$-lexicase selection will result in models that perform significantly different than models that are evolved using tournament selection:
+2. Statistical significant differences between the mean errors of training and testing data exist for both selection operators
+3. No differences in program size exist between the distribution underlying the samples produced by $\epsilon$-lexicase and the distribution underlying samples of tournament selection
+
 # Results

### 2022-06-26 docs/rmd/presentation/slides.tex
...

--- 2022-06-25 docs/rmd/presentation/slides.tex
@@
   pdftitle={Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression},
+  pdfauthor={Roman Hoehn, B.Sc. Wirtschaftspaedagogik},
   hidelinks,
@@
 \item
-  One of the first applications of Genetic Programming introduced
+  One of the first applications of GP described by Koza (1992)
+\item
+  High relevance: GP can outperform state-of-the-art machine learning
+  algorithms like gradient boosting\footnote<.->{Orzechowski, Cava and
+    Moore (2018)}
 \end{itemize}
-
-\ldots{}
 \end{frame}
@@
 \item
-  Generalization
+  Little attention has been paid to generalization in GP\footnote<.->{O'Neill
+    \emph{et al.} (2010), Kushchu (2002)}
+\item
+  High practical relevance of symbolic regression in many fields,
+  e.g.~financial forecasting
 \end{itemize}
-\end{frame}
-
-\begin{frame}{Genetic Programming}
-\protect\hypertarget{genetic-programming-1}{}
-\ldots{}
-
-\ldots{}
 \end{frame}
@@
 
-\begin{frame}{Experimental Study}
-\ldots{}
+\begin{frame}{Benchmark problem}
+\protect\hypertarget{benchmark-problem}{}
+UC Irvine Machine Learning Repository: Prediction of energy efficiency
+in buildings\footnote<.->{Dua and Graff (2017)}
+
+\begin{table}[!h]
+
+\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating data set}
+\centering
+\begin{tabular}[t]{l|l}
+\hline
+\textbf{Variable} & \textbf{Description}\\
+\hline
+X1 & Relative Compactness\\
+\hline
+X2 & Surface Area\\
+\hline
+X3 & Wall Area\\
+\hline
+X4 & Roof Area\\
+\hline
+X5 & Overall Height\\
+\hline
+X6 & Orientation\\
+\hline
+X7 & Glazing Area\\
+\hline
+X8 & Glazing Area Distribution\\
+\hline
+y1 & Heating Load\\
+\hline
+y2 & Cooling Load\\
+\hline
+\end{tabular}
+\end{table}
 \end{frame}
 
-\begin{frame}{Research Design}
-\protect\hypertarget{research-design}{}
-\ldots{}
+\begin{frame}{Experiment}
+\protect\hypertarget{experiment}{}
+\begin{block}{Single run}
+\protect\hypertarget{single-run}{}
+\begin{itemize}
+\tightlist
+\item
+  Total dataset (\(N=768\)) is randomly split into a training and
+  testing dataset (50:50)
+\item
+  Train two models using GP with the training dataset only, one using
+  tournament selection and the other \(\epsilon\)-lexicase selection
+\item
+  For each generation: Select elite model and compute its fitness for
+  the testing dataset
+\end{itemize}
+\end{block}
+
+\begin{block}{Full experiment}
+\protect\hypertarget{full-experiment}{}
+\begin{itemize}
+\tightlist
+\item
+  Stochastic algorithm: Repeat the basic experiment for 50 total runs
+\item
+  Collect and aggregate results for training error, testing error and
+  program length
+\end{itemize}
+\end{block}
 \end{frame}
@@
 \ldots{}
+\end{frame}
+
+\begin{frame}{Research Question}
+\protect\hypertarget{research-question-1}{}
+\begin{itemize}
+\tightlist
+\item
+  Does the usage of \(\epsilon\)-lexicase parent selection influence the
+  generalization behaviour of genetic programming in symbolic regression
+  if compared to tournament selection?
+\end{itemize}
+
+\begin{block}{Hypothesis testing}
+\protect\hypertarget{hypothesis-testing}{}
+\begin{block}{TODO: reformulate hypothesis}
+\protect\hypertarget{todo-reformulate-hypothesis}{}
+\begin{enumerate}
+\tightlist
+\item
+  The usage of \(\epsilon\)-lexicase selection will result in models
+  that perform significantly different than models that are evolved
+  using tournament selection:
+\item
+  Statistical significant differences between the mean errors of
+  training and testing data exist for both selection operators
+\item
+  No differences in program size exist between the distribution
+  underlying the samples produced by \(\epsilon\)-lexicase and the
+  distribution underlying samples of tournament selection
+\end{enumerate}
+\end{block}
+\end{block}
 \end{frame}
@@
 \begin{cslreferences}
+\leavevmode\hypertarget{ref-Dua:2019}{}%
+Dua, D. and Graff, C. (2017) `UCI machine learning repository'.
+University of California, Irvine, School of Information; Computer
+Sciences. Available at: \url{http://archive.ics.uci.edu/ml}.
+
 \leavevmode\hypertarget{ref-10.1007ux2f978-3-642-16493-4_19}{}%
@@
 
+\leavevmode\hypertarget{ref-generalisation_in_gp}{}%
+Kushchu, I. (2002) `An evaluation of evolutionarygeneralisation in
+genetic programming', \emph{Artificial Intelligence Review - AIR}, 18,
+pp. 3--14.
+doi:\href{https://doi.org/10.1023/A:1016379201230}{10.1023/A:1016379201230}.
+
 \leavevmode\hypertarget{ref-epsilon_lexicase_main}{}%
@@
 doi:\href{https://doi.org/10.1145/2908812.2908898}{10.1145/2908812.2908898}.
+
+\leavevmode\hypertarget{ref-open_issues_gp}{}%
+O'Neill, M. \emph{et al.} (2010) `Open issues in genetic programming',
+\emph{Genetic Programming and Evolvable Machines}, 11, pp. 339--363.
+doi:\href{https://doi.org/10.1007/s10710-010-9113-2}{10.1007/s10710-010-9113-2}.
+
+\leavevmode\hypertarget{ref-Orzechowski_2018}{}%
+Orzechowski, P., Cava, W.L. and Moore, J.H. (2018) `Where are we now?',
+in \emph{Proceedings of the genetic and evolutionary computation
+conference}. ACM.
+doi:\href{https://doi.org/10.1145/3205455.3205539}{10.1145/3205455.3205539}.
 \end{cslreferences}

### 2022-06-26 src/__old__/single_run.py

@@ -1,7 +1,6 @@
 from copy import deepcopy
 from sys import argv
-
-from seminar.src.old_version.gp_symreg import *
+from gp_symreg import *
 
 
 def run(target_var: str, err_metric: str):

### 2022-06-26 src/__old__/single_run.py
added plot

@@ -35,6 +35,8 @@ def run(target_var: str, err_metric: str):
         toolbox=deepcopy(TOOLBOX),
         ea_config=EA_CONFIG,
         training_dataset=TRAINING_DATASET,
+        pset=PRIMITIVE_SET,
+        testing_dataset=TESTING_DATASET
     )
 
     # test model on unknown dataset

### 2022-06-26 docs/rmd/presentation/slides.Rmd
addded some tables

@@
     fonttheme: "structurebold"
+    #extra_dependencies: ["float"]
     keep_tex: true
@@
     slide_level: 2
+    #includes:
+      #in_header: ./settings.tex
 toc: yes
@@
 * Total dataset ($N=768$) is randomly split into a training and testing dataset (50:50)
+* Fitness metric: Mean squared Error (MSE)
 * Train two models using GP with the training dataset only, one using tournament selection and the other $\epsilon$-lexicase selection
@@
 
-## Genetic Programming Configuration
+## Evolutionary Parameters
 
-...
+```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+fig.pos="H"
+
+knitr::kable(
+  read.csv("../tables/csv/gp_parameters.csv"),
+  digits=3,
+  caption="Evolutionary Parameters"
+) %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
+```
+
+## Function Set
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+fig.pos="b"
+
+knitr::kable(
+  read.csv("../tables/csv/primitives.csv"),
+  digits=3,
+) 
+```
+
+## Terminal Set
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+fig.pos="t"
+
+knitr::kable(
+  read.csv("../tables/csv/terminals.csv"),
+  digits=3,
+) 
+```
+
+## Examples
+
+Model evolved by tournament selection after 100 generations:
+
+![](../plots/example_model.png)
 

### 2022-06-26 docs/rmd/presentation/slides.tex
addded some tables

@@
 \item
-  Performance increases have been demosntrated in many benchmarking
+  Performance increases have been demonstrated in many benchmarking
   problems\footnote<.->{La Cava, Spector and Danai (2016), p.744-745}
@@
 \item
+  Fitness metric: Mean squared Error (MSE)
+\item
   Train two models using GP with the training dataset only, one using
@@
 
-\begin{frame}{Genetic Programming Configuration}
-\protect\hypertarget{genetic-programming-configuration}{}
-\ldots{}
+\begin{frame}{Evolutionary Parameters}
+\protect\hypertarget{evolutionary-parameters}{}
+\begin{table}[!h]
+
+\caption{\label{tab:unnamed-chunk-2}Evolutionary Parameters}
+\centering
+\begin{tabular}[t]{l|l}
+\hline
+\textbf{Parameter} & \textbf{Value}\\
+\hline
+Population Size & 500\\
+\hline
+Number of Generations & 100\\
+\hline
+Mutation Rate & 20\%\\
+\hline
+Crossover Rate & 80\%\\
+\hline
+Tournament Size & 3\\
+\hline
+Epsilon selection & automatic\\
+\hline
+Elite Size & 0\\
+\hline
+\end{tabular}
+\end{table}
+\end{frame}
+
+\begin{frame}{Terminal Set}
+\protect\hypertarget{terminal-set}{}
+\begin{tabular}{l|l}
+\hline
+Terminal & Description\\
+\hline
+X1 & Relative Compactness\\
+\hline
+X2 & Surface Area\\
+\hline
+X3 & Wall Area\\
+\hline
+X4 & Roof Area\\
+\hline
+X5 & Overall Height\\
+\hline
+X6 & Orientation\\
+\hline
+X7 & Glazing Area\\
+\hline
+X8 & Glazing Area Distribution\\
+\hline
+random\_int & Ephemeral Constant (integer)\\
+\hline
+random\_float & Ephemeral Constant(float)\\
+\hline
+\end{tabular}
+\end{frame}
+
+\begin{frame}{Function Set}
+\protect\hypertarget{function-set}{}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-4}Functions}
+\centering
+\begin{tabular}[t]{l|r}
+\hline
+Function & Arity\\
+\hline
+Addition & 2\\
+\hline
+Subtraction & 2\\
+\hline
+Multiplication & 2\\
+\hline
+Negation & 1\\
+\hline
+Sine & 1\\
+\hline
+Cosine & 1\\
+\hline
+Protected Division & 2\\
+\hline
+Protected Natural Logarithm & 1\\
+\hline
+Protected Square Root & 1\\
+\hline
+\end{tabular}
+\end{table}
+\end{frame}
+
+\begin{frame}{Examples}
+\protect\hypertarget{examples}{}
+Model evolved by tournament selection after 100 generations:
+
+\includegraphics{../plots/example_model.png}
 \end{frame}

### 2022-06-26 docs/rmd/presentation/slides.Rmd
bugged latex fragment

@@
     fonttheme: "structurebold"
-    #extra_dependencies: ["float"]
     keep_tex: true
@@
     slide_level: 2
-    #includes:
-      #in_header: ./settings.tex
 toc: yes
-# toc_depth: 3
 bibliography: ../bib/bib/ref.bib
@@
 knitr::opts_chunk$set(echo = TRUE)
-knitr::opts_chunk$set(fig.pos = "!H")
 ```
@@
   caption="Overview - Energy Heating data set"
-)    %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
+) 
 ```
@@
   caption="Evolutionary Parameters"
-) %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
+)
 ```
@@
 
-## Examples
+## Example
 
@@
 
-* Does the usage of $\epsilon$-lexicase parent selection influence the generalization behaviour of genetic programming in symbolic regression if compared to tournament selection?
+* Does the usage of $\epsilon$-lexicase parent selection influence the generalization behavior of genetic programming in symbolic regression if compared to tournament selection?
 
-### Hypothesis testing
+### 3 Questions
 
-#### TODO: reformulate hypothesis
-1. The usage of $\epsilon$-lexicase selection will result in models that perform significantly different than models that are evolved using tournament selection:
-2. Statistical significant differences between the mean errors of training and testing data exist for both selection operators
-3. No differences in program size exist between the distribution underlying the samples produced by $\epsilon$-lexicase and the distribution underlying samples of tournament selection
+1. Differences in average fitness for both algorithms?
+2. Differences in fitness for training and testing data?
+3. Differences in program size?
 
@@
 
-## Descriptive Statistics
+## Distribution of Fitness
 
-![Distribution of Errors](../plots/mean_error_boxplot_all.png)
+![](../plots/mean_error_boxplot_all.png)
+
+## Finding 1
+
+* The differences in average fitness of the final solutions between tournament selection and $\epsilon$-lexicase selection are highly statistical significant ($\alpha=0.01$)
+* Tournament selection-based GP achieves a higher fitness for both training and testing data
+* Unexpected results based on the reviewed literature [@epsilon_lexicase_main], [@https://doi.org/10.48550/arxiv.1709.05394]
+
+## Evolution of Fitness
+
+![](../plots/mean_error_combined.png)
+
+## Finding 2
+
+* The gap between training and testing error is not statistically significant for both selection algorithms
+* Both algorithms achieve a slightly better performance for the training data
+* No proof for overfitting
+
+## Statistical Test
+
+```{r, echo=FALSE, warning=FALSE}
+
+df <- read.csv("../tables/csv/mwu_matrix_error.csv")
+
+knitr::kable(
+  df,
+  format = "latex",
+  booktabs = TRUE,
+  digits=3,
+  caption="Mean Error - P-Values (MWU)"
+)
+
+```
+
+## Evolution of Size
+
+\begin{figure}
+\centering
+\includegraphics{../plots/size_subplotted.png}
+\end{figure}
+
+## Finding 3
 

### 2022-06-26 docs/rmd/presentation/slides.tex
bugged latex fragment

@@
 
-\begin{table}[!h]
+\begin{table}
 
@@
 \hline
-\textbf{Variable} & \textbf{Description}\\
+Variable & Description\\
 \hline
@@
 \protect\hypertarget{evolutionary-parameters}{}
-\begin{table}[!h]
+\begin{table}
 
@@
 \hline
-\textbf{Parameter} & \textbf{Value}\\
+Parameter & Value\\
 \hline
@@
 \end{table}
+\end{frame}
+
+\begin{frame}{Function Set}
+\protect\hypertarget{function-set}{}
+\begin{tabular}{l|r}
+\hline
+Function & Arity\\
+\hline
+Addition & 2\\
+\hline
+Subtraction & 2\\
+\hline
+Multiplication & 2\\
+\hline
+Negation & 1\\
+\hline
+Sine & 1\\
+\hline
+Cosine & 1\\
+\hline
+Protected Division & 2\\
+\hline
+Protected Natural Logarithm & 1\\
+\hline
+Protected Square Root & 1\\
+\hline
+\end{tabular}
 \end{frame}
@@
 
-\begin{frame}{Function Set}
-\protect\hypertarget{function-set}{}
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-4}Functions}
-\centering
-\begin{tabular}[t]{l|r}
-\hline
-Function & Arity\\
-\hline
-Addition & 2\\
-\hline
-Subtraction & 2\\
-\hline
-Multiplication & 2\\
-\hline
-Negation & 1\\
-\hline
-Sine & 1\\
-\hline
-Cosine & 1\\
-\hline
-Protected Division & 2\\
-\hline
-Protected Natural Logarithm & 1\\
-\hline
-Protected Square Root & 1\\
-\hline
-\end{tabular}
-\end{table}
-\end{frame}
-
-\begin{frame}{Examples}
-\protect\hypertarget{examples}{}
+\begin{frame}{Example}
+\protect\hypertarget{example}{}
 Model evolved by tournament selection after 100 generations:
@@
   Does the usage of \(\epsilon\)-lexicase parent selection influence the
-  generalization behaviour of genetic programming in symbolic regression
+  generalization behavior of genetic programming in symbolic regression
   if compared to tournament selection?
@@
 
-\begin{block}{Hypothesis testing}
-\protect\hypertarget{hypothesis-testing}{}
-\begin{block}{TODO: reformulate hypothesis}
-\protect\hypertarget{todo-reformulate-hypothesis}{}
+\begin{block}{3 Questions}
+\protect\hypertarget{questions}{}
 \begin{enumerate}
@@
 \item
-  The usage of \(\epsilon\)-lexicase selection will result in models
-  that perform significantly different than models that are evolved
-  using tournament selection:
+  Differences in average fitness for both algorithms?
 \item
-  Statistical significant differences between the mean errors of
-  training and testing data exist for both selection operators
+  Differences in fitness for training and testing data?
 \item
-  No differences in program size exist between the distribution
-  underlying the samples produced by \(\epsilon\)-lexicase and the
-  distribution underlying samples of tournament selection
+  Differences in program size?
 \end{enumerate}
-\end{block}
 \end{block}
@@
 
-\begin{frame}{Descriptive Statistics}
-\protect\hypertarget{descriptive-statistics}{}
+\begin{frame}{Finding 1}
+\protect\hypertarget{finding-1}{}
+\begin{itemize}
+\tightlist
+\item
+  The differences in average fitness of the final solutions between
+  tournament selection and \(\epsilon\)-lexicase selection are highly
+  statistical significant (\(\alpha=0.01\))
+\item
+  Tournament selection-based GP achieves a higher fitness for both
+  training and testing data
+\item
+  Unexpected results based on the reviewed literature (La Cava, Spector
+  and Danai, 2016), (La Cava \emph{et al.}, 2017)
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Statistical Test}
+\protect\hypertarget{statistical-test}{}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-5}Mean Error - P-Values (MWU)}
+\centering
+\begin{tabular}[t]{lrrrr}
+\toprule
+  & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
+\midrule
+tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
+tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
+elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
+elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
+\bottomrule
+\end{tabular}
+\end{table}
+\end{frame}
+
+\begin{frame}{Distribution of Fitness}
+\protect\hypertarget{distribution-of-fitness}{}
+\includegraphics{../plots/mean_error_boxplot_all.png}
+\end{frame}
+
+\begin{frame}{Finding 2}
+\protect\hypertarget{finding-2}{}
+\ldots{}
+\end{frame}
+
+\begin{frame}{Evolution of Fitness}
+\protect\hypertarget{evolution-of-fitness}{}
+\includegraphics{../plots/mean_error_combined.png}
+\end{frame}
+
+\begin{frame}{Evolution of Size}
+\protect\hypertarget{evolution-of-size}{}
 \begin{figure}
 \centering
-\includegraphics{../plots/mean_error_boxplot_all.png}
-\caption{Distribution of Errors}
+\includegraphics{../plots/size_subplotted.png}
 \end{figure}
@@
 
+\leavevmode\hypertarget{ref-https:ux2fux2fdoi.orgux2f10.48550ux2farxiv.1709.05394}{}%
+La Cava, W. \emph{et al.} (2017) `A probabilistic and multi-objective
+analysis of lexicase selection and epsilon-lexicase selection'. arXiv.
+doi:\href{https://doi.org/10.48550/ARXIV.1709.05394}{10.48550/ARXIV.1709.05394}.
+
 \leavevmode\hypertarget{ref-epsilon_lexicase_main}{}%

### 2022-06-26 docs/rmd/presentation/slides.tex
autogit

@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression},
-  pdfauthor={Roman Hoehn, B.Sc. Wirtschaftspaedagogik},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \newif\ifbibliography
@@
   Performance increases have been demonstrated in many benchmarking
-  problems\footnote<.->{La Cava, Spector and Danai (2016), p.744-745}
+  problems \footnote<.->{La Cava, Spector and Danai (2016), p.744-745}
 \end{itemize}
@@
   High relevance: GP can outperform state-of-the-art machine learning
-  algorithms like gradient boosting\footnote<.->{Orzechowski, Cava and
+  algorithms like gradient boosting \footnote<.->{Orzechowski, Cava and
     Moore (2018)}
@@
 \item
-  Little attention has been paid to generalization in GP\footnote<.->{O'Neill
+  Little attention has been paid to generalization in GP \footnote<.->{O'Neill
     \emph{et al.} (2010), Kushchu (2002)}
@@
 UC Irvine Machine Learning Repository: Prediction of energy efficiency
-in buildings\footnote<.->{Dua and Graff (2017)}
+in buildings \footnote<.->{Dua and Graff (2017)}
 
@@
 
+\begin{frame}{Distribution of Fitness}
+\protect\hypertarget{distribution-of-fitness}{}
+\includegraphics{../plots/mean_error_boxplot_all.png}
+\end{frame}
+
 \begin{frame}{Finding 1}
@@
   and Danai, 2016), (La Cava \emph{et al.}, 2017)
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Evolution of Fitness}
+\protect\hypertarget{evolution-of-fitness}{}
+\includegraphics{../plots/mean_error_combined.png}
+\end{frame}
+
+\begin{frame}{Finding 2}
+\protect\hypertarget{finding-2}{}
+\begin{itemize}
+\tightlist
+\item
+  The gap between training and testing error is not statistically
+  significant for both selection algorithms
+\item
+  Both algorithms achieve a slightly better performance for the training
+  data
+\item
+  No proof for overfitting
 \end{itemize}
@@
 
-\begin{frame}{Distribution of Fitness}
-\protect\hypertarget{distribution-of-fitness}{}
-\includegraphics{../plots/mean_error_boxplot_all.png}
-\end{frame}
-
-\begin{frame}{Finding 2}
-\protect\hypertarget{finding-2}{}
-\ldots{}
-\end{frame}
-
-\begin{frame}{Evolution of Fitness}
-\protect\hypertarget{evolution-of-fitness}{}
-\includegraphics{../plots/mean_error_combined.png}
-\end{frame}
-
 \begin{frame}{Evolution of Size}
@@
 \end{figure}
+\end{frame}
 
+\begin{frame}{Finding 3}
+\protect\hypertarget{finding-3}{}
 \ldots{}
@@
 \hypertarget{refs}{}
-\begin{cslreferences}
-\leavevmode\hypertarget{ref-Dua:2019}{}%
-Dua, D. and Graff, C. (2017) `UCI machine learning repository'.
+\begin{CSLReferences}{0}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-Dua:2019}{}}%
+Dua, D. and Graff, C. (2017) {``{UCI} machine learning repository.''}
 University of California, Irvine, School of Information; Computer
@@
 
-\leavevmode\hypertarget{ref-10.1007ux2f978-3-642-16493-4_19}{}%
-Fang, Y. and Li, J. (2010) `A review of tournament selection in genetic
-programming', in Cai, Z. et al. (eds) \emph{Advances in computation and
-intelligence}. Berlin, Heidelberg: Springer Berlin Heidelberg, pp.
-181--192.
+\leavevmode\vadjust pre{\hypertarget{ref-10.1007ux2f978-3-642-16493-4_19}{}}%
+Fang, Y. and Li, J. (2010) {``A review of tournament selection in
+genetic programming,''} in Cai, Z. et al. (eds.) \emph{Advances in
+computation and intelligence}. Berlin, Heidelberg: Springer Berlin
+Heidelberg, pp. 181--192.
 
-\leavevmode\hypertarget{ref-6920034}{}%
-Helmuth, T., Spector, L. and Matheson, J. (2015) `Solving uncompromising
-problems with lexicase selection', \emph{IEEE Transactions on
-Evolutionary Computation}, 19(5), pp. 630--643.
+\leavevmode\vadjust pre{\hypertarget{ref-6920034}{}}%
+Helmuth, T., Spector, L. and Matheson, J. (2015) {``Solving
+uncompromising problems with lexicase selection,''} \emph{IEEE
+Transactions on Evolutionary Computation}, 19(5), pp. 630--643.
 doi:\href{https://doi.org/10.1109/TEVC.2014.2362729}{10.1109/TEVC.2014.2362729}.
 
-\leavevmode\hypertarget{ref-koza_main}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-koza_main}{}}%
 Koza, J.R. (1992) \emph{Genetic programming: On the programming of
@@
 
-\leavevmode\hypertarget{ref-generalisation_in_gp}{}%
-Kushchu, I. (2002) `An evaluation of evolutionarygeneralisation in
-genetic programming', \emph{Artificial Intelligence Review - AIR}, 18,
+\leavevmode\vadjust pre{\hypertarget{ref-generalisation_in_gp}{}}%
+Kushchu, I. (2002) {``An evaluation of EvolutionaryGeneralisation in
+genetic programming,''} \emph{Artificial Intelligence Review - AIR}, 18,
 pp. 3--14.
@@
 
-\leavevmode\hypertarget{ref-https:ux2fux2fdoi.orgux2f10.48550ux2farxiv.1709.05394}{}%
-La Cava, W. \emph{et al.} (2017) `A probabilistic and multi-objective
-analysis of lexicase selection and epsilon-lexicase selection'. arXiv.
+\leavevmode\vadjust pre{\hypertarget{ref-https:ux2fux2fdoi.orgux2f10.48550ux2farxiv.1709.05394}{}}%
+La Cava, W. \emph{et al.} (2017) {``A probabilistic and multi-objective
+analysis of lexicase selection and epsilon-lexicase selection.''} arXiv.
 doi:\href{https://doi.org/10.48550/ARXIV.1709.05394}{10.48550/ARXIV.1709.05394}.
 
-\leavevmode\hypertarget{ref-epsilon_lexicase_main}{}%
-La Cava, W., Spector, L. and Danai, K. (2016) `Epsilon-lexicase
-selection for regression', in \emph{Proceedings of the genetic and
+\leavevmode\vadjust pre{\hypertarget{ref-epsilon_lexicase_main}{}}%
+La Cava, W., Spector, L. and Danai, K. (2016) {``Epsilon-lexicase
+selection for regression,''} in \emph{Proceedings of the genetic and
 evolutionary computation conference 2016}. New York, NY, USA:
@@
 
-\leavevmode\hypertarget{ref-open_issues_gp}{}%
-O'Neill, M. \emph{et al.} (2010) `Open issues in genetic programming',
-\emph{Genetic Programming and Evolvable Machines}, 11, pp. 339--363.
+\leavevmode\vadjust pre{\hypertarget{ref-open_issues_gp}{}}%
+O'Neill, M. \emph{et al.} (2010) {``Open issues in genetic
+programming,''} \emph{Genetic Programming and Evolvable Machines}, 11,
+pp. 339--363.
 doi:\href{https://doi.org/10.1007/s10710-010-9113-2}{10.1007/s10710-010-9113-2}.
 
-\leavevmode\hypertarget{ref-Orzechowski_2018}{}%
-Orzechowski, P., Cava, W.L. and Moore, J.H. (2018) `Where are we now?',
-in \emph{Proceedings of the genetic and evolutionary computation
-conference}. ACM.
+\leavevmode\vadjust pre{\hypertarget{ref-Orzechowski_2018}{}}%
+Orzechowski, P., Cava, W.L. and Moore, J.H. (2018) {``Where are we
+now?''} in \emph{Proceedings of the genetic and evolutionary computation
+conference}. {ACM}.
 doi:\href{https://doi.org/10.1145/3205455.3205539}{10.1145/3205455.3205539}.
-\end{cslreferences}
+
+\end{CSLReferences}
 \end{frame}

### 2022-06-26 docs/rmd/presentation/slides.tex
autogit

@@
 
-\begin{frame}{Statistical Test}
+\begin{frame}[fragile]{Statistical Test}
 \protect\hypertarget{statistical-test}{}
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-5}Mean Error - P-Values (MWU)}
-\centering
-\begin{tabular}[t]{lrrrr}
-\toprule
-  & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
-\midrule
-tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
-tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
-elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
-elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
-\bottomrule
-\end{tabular}
-\end{table}
+\begin{verbatim}
+##                            tournament_training_errors tournament_testing_errors
+## tournament_training_errors               1.000000e+00              0.3092302963
+## tournament_testing_errors                3.092303e-01              1.0000000000
+## elexicase_training_errors                1.180222e-04              0.0024468832
+## elexicase_testing_errors                 4.192079e-06              0.0001213887
+##                            elexicase_training_errors elexicase_testing_errors
+## tournament_training_errors              0.0001180222             4.192079e-06
+## tournament_testing_errors               0.0024468832             1.213887e-04
+## elexicase_training_errors               1.0000000000             2.567806e-01
+## elexicase_testing_errors                0.2567805508             1.000000e+00
+\end{verbatim}
 \end{frame}

### 2022-06-26 docs/rmd/presentation/slides.Rmd
autogit
kinda done

@@
 csl: ../bib/csl/harvard1.csl
+header-includes:
+   - \usepackage{float}
 ---
@@
 ```{r setup, include=FALSE}
-library(kableExtra)
 library(magrittr)
@@
   read.csv("../tables/csv/gp_parameters.csv"),
+  format="latex",
   digits=3,
@@
 
-## Function Set
+## Primitive Set {.allowframebreaks}
 
@@
   read.csv("../tables/csv/primitives.csv"),
+  format="latex",
   digits=3,
+  caption="Function Set"
 ) 
 ```
-
-## Terminal Set
+\framebreak
 
@@
   read.csv("../tables/csv/terminals.csv"),
+  format="latex",
   digits=3,
+  caption="Terminal Set"
 ) 
@@
 
-### 3 Questions
+### Statistical Testing Strategy
 
-1. Differences in average fitness for both algorithms?
+1. Differences in average fitness for both algorithms on both datasets?
 2. Differences in fitness for training and testing data?
-3. Differences in program size?
 
 # Results
-
-## Distribution of Fitness
-
-![](../plots/mean_error_boxplot_all.png)
 
@@
 
-## Evolution of Fitness
+## Distribution of Fitness
 
-![](../plots/mean_error_combined.png)
+![](../plots/mean_error_boxplot_all.png)
 
@@
 * Both algorithms achieve a slightly better performance for the training data
-* No proof for overfitting
+* Good generalization: No evidence of overfitting
+
+## Evolution of Fitness
+
+![](../plots/mean_error_combined.png)
 
@@
 ```{r, echo=FALSE, warning=FALSE}
-
 df <- read.csv("../tables/csv/mwu_matrix_error.csv")
@@
   df,
-  format = "latex",
-  booktabs = TRUE,
+  format="latex",
   digits=3,
   caption="Mean Error - P-Values (MWU)"
-)
+) %>% kableExtra::kable_styling(font_size = 4)
 
 ```
+
+## Program Growth
+
+* So far: No proof of differences in generalization
+* New approach: Program growth as a possible indicator for overfitting?
+* Theory: Minimum description length principle (MDLP) ^[@wang_wagner_rondinelli_2019, p. 268]
+* Downside: Growth/Bloat is no clear indicator of overfitting ^[@bloat_overfitting_gp, p. 8]
 
@@
 
-...
+* GP typical growth behaviour for both operators
+* Solutions grow at a similiar rate in each generation
+* No statistically significant differences in overall program size based on selection
 
@@
 
-...
+* Experiment did not yield evidence for differences in generalization behavior between tournament and $\epsilon$-lexicase selection
+* The performance of tournament selection is significantly higher than that of $\epsilon$-lexicase selection for the selected symbolic regression problem
+* No evidence for differences in growth behavior between both algorithms
 
@@
 
-...
-\newpage
+1. Configuration of evolutionary parameters
+2. Results are based on a single symbolic regression
+2. Limited by computational resources
+
+# References {.allowframebreaks}

### 2022-06-26 docs/rmd/presentation/slides.tex
kinda done

--- 2022-06-26 docs/rmd/presentation/slides.tex
@@
 \newif\ifbibliography
+\usepackage{longtable,booktabs}
+\usepackage{caption}
+% Make caption package work with longtable
+\makeatletter
+\def\fnum@table{\tablename~\thetable}
+\makeatother
 \usepackage{graphicx}
@@
 \setcounter{secnumdepth}{5}
+\usepackage{float}
 \newlength{\cslhangindent}
@@
 
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating data set}
-\centering
-\begin{tabular}[t]{l|l}
-\hline
-Variable & Description\\
-\hline
-X1 & Relative Compactness\\
-\hline
-X2 & Surface Area\\
-\hline
-X3 & Wall Area\\
-\hline
-X4 & Roof Area\\
-\hline
-X5 & Overall Height\\
-\hline
-X6 & Orientation\\
-\hline
-X7 & Glazing Area\\
-\hline
-X8 & Glazing Area Distribution\\
-\hline
-y1 & Heating Load\\
-\hline
-y2 & Cooling Load\\
-\hline
-\end{tabular}
-\end{table}
+\begin{longtable}[]{@{}ll@{}}
+\caption{Overview - Energy Heating data set}\tabularnewline
+\toprule
+Variable & Description\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+Variable & Description\tabularnewline
+\midrule
+\endhead
+X1 & Relative Compactness\tabularnewline
+X2 & Surface Area\tabularnewline
+X3 & Wall Area\tabularnewline
+X4 & Roof Area\tabularnewline
+X5 & Overall Height\tabularnewline
+X6 & Orientation\tabularnewline
+X7 & Glazing Area\tabularnewline
+X8 & Glazing Area Distribution\tabularnewline
+y1 & Heating Load\tabularnewline
+y2 & Cooling Load\tabularnewline
+\bottomrule
+\end{longtable}
 \end{frame}
@@
 
-\begin{frame}{Function Set}
-\protect\hypertarget{function-set}{}
-\begin{tabular}{l|r}
+\begin{frame}[allowframebreaks]{Primitive Set}
+\protect\hypertarget{primitive-set}{}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-3}Functions}
+\centering
+\begin{tabular}[t]{l|r}
 \hline
@@
 \end{tabular}
-\end{frame}
+\end{table}
+\framebreak
 
-\begin{frame}{Terminal Set}
-\protect\hypertarget{terminal-set}{}
-\begin{tabular}{l|l}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-4}Terminals}
+\centering
+\begin{tabular}[t]{l|l}
 \hline
@@
 \end{tabular}
+\end{table}
 \end{frame}
@@
 
-\begin{block}{3 Questions}
-\protect\hypertarget{questions}{}
+\begin{block}{Statistical Testing Strategy}
+\protect\hypertarget{statistical-testing-strategy}{}
 \begin{enumerate}
@@
 \item
-  Differences in average fitness for both algorithms?
+  Differences in average fitness for both algorithms on both datasets?
 \item
   Differences in fitness for training and testing data?
-\item
-  Differences in program size?
 \end{enumerate}
@@
 
-\begin{frame}{Statistical Test}
-\protect\hypertarget{statistical-test}{}
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-5}Mean Error - P-Values (MWU)}
-\centering
-\begin{tabular}[t]{lrrrr}
-\toprule
-  & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
-\midrule
-tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
-tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
-elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
-elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
-\bottomrule
-\end{tabular}
-\end{table}
-\end{frame}
-
 \begin{frame}{Distribution of Fitness}
@@
 \protect\hypertarget{finding-2}{}
-\ldots{}
+\begin{itemize}
+\tightlist
+\item
+  The gap between training and testing error is not statistically
+  significant for both selection algorithms
+\item
+  Both algorithms achieve a slightly better performance for the training
+  data
+\item
+  Good generalization: No evidence of overfitting
+\end{itemize}
 \end{frame}
@@
 \includegraphics{../plots/mean_error_combined.png}
+\end{frame}
+
+\begin{frame}{Statistical Test}
+\protect\hypertarget{statistical-test}{}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-5}Mean Error - P-Values (MWU)}
+\centering
+\fontsize{4}{6}\selectfont
+\begin{tabular}[t]{l|r|r|r|r}
+\hline
+X & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
+\hline
+tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
+\hline
+tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
+\hline
+elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
+\hline
+elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
+\hline
+\end{tabular}
+\end{table}
+\end{frame}
+
+\begin{frame}{Program Growth}
+\protect\hypertarget{program-growth}{}
+\begin{itemize}
+\tightlist
+\item
+  So far: No proof of differences in generalization
+\item
+  New approach: Program growth as a possible indicator for overfitting?
+\item
+  Theory: Minimum description length principle (MDLP)\footnote<.->{Wang,
+    Wagner and Rondinelli (2019), p.~268}
+\item
+  Downside: Growth/Bloat is no clear indicator of
+  overfitting\footnote<.->{Silva and Vanneschi (2009), p.~8}
+\end{itemize}
 \end{frame}
@@
 \end{figure}
+\end{frame}
 
-\ldots{}
+\begin{frame}{Finding 3}
+\protect\hypertarget{finding-3}{}
+\begin{itemize}
+\tightlist
+\item
+  GP typical growth behaviour for both operators
+\item
+  Solutions grow at a similiar rate in each generation
+\item
+  No statistically significant differences in overall program size based
+  on selection
+\end{itemize}
 \end{frame}
@@
 \begin{frame}{Conclusions}
-\ldots{}
+\begin{itemize}
+\tightlist
+\item
+  Experiment did not yield evidence for differences in generalization
+  behavior between tournament and \(\epsilon\)-lexicase selection
+\item
+  The performance of tournament selection is significantly higher than
+  that of \(\epsilon\)-lexicase selection for the selected symbolic
+  regression problem
+\item
+  No evidence for differences in growth behavior between both algorithms
+\end{itemize}
 \end{frame}
@@
 \begin{frame}{Limitations and open Questions}
-\ldots{} \newpage
+\begin{enumerate}
+\tightlist
+\item
+  Configuration of evolutionary parameters
+\item
+  Results are based on a single symbolic regression
+\item
+  Limited by computational resources
+\end{enumerate}
+\end{frame}
 
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\begin{frame}[allowframebreaks]{References}
 \hypertarget{refs}{}
@@
 doi:\href{https://doi.org/10.1145/3205455.3205539}{10.1145/3205455.3205539}.
+
+\leavevmode\hypertarget{ref-bloat_overfitting_gp}{}%
+Silva, S. and Vanneschi, L. (2009) `Operator equalisation, bloat and
+overfitting: A study on human oral bioavailability prediction', in
+\emph{Proceedings of the 11th Annual Genetic and Evolutionary
+Computation Conference, GECCO-2009}, pp. 1115--1122.
+doi:\href{https://doi.org/10.1145/1569901.1570051}{10.1145/1569901.1570051}.
+
+\leavevmode\hypertarget{ref-wang_wagner_rondinelli_2019}{}%
+Wang, Y., Wagner, N. and Rondinelli, J.M. (2019) `Symbolic regression in
+materials science', \emph{MRS Communications}, 9(3), pp. 793--805.
+doi:\href{https://doi.org/10.1557/mrc.2019.85}{10.1557/mrc.2019.85}.
 \end{cslreferences}

### 2022-06-27 docs/rmd/presentation/slides.Rmd
autogit

@@
   beamer_presentation:
-    colortheme: "beaver"
     fonttheme: "structurebold"
@@
 
-* A metaheuristic that searches for computer programs that solve a given problem
-* Inventor: John R. Koza ^[@koza_main]
-* Evolutionary algorithm that simulates the process of Darwinian evolution:
-  1. Population based
-  2. The quality of solutions is evaluated by a fitness function
-  3. Selection: Solutions are selected based on their individual fitness
-  4. Variation: Mutation and recombination of solutions
+* A metaheuristic that searches for computer programs that solve a given problem^[@koza_main]
+* Evolutionary Algorithm: Simulates the process of Darwinian evolution
 * Unique Features:
@@
 
-* Operator that selects individual solutions from the population for reproduction and mutation
-* Most commonly used selection operator in Genetic Programming (GP): Tournament selection^[@10.1007/978-3-642-16493-4_19, p.181]
+### Torunament Selection
+
+* Most commonly used selection operator in Genetic Programming (GP)^[@10.1007/978-3-642-16493-4_19, p.181]
 * Intuition: High chance for "generalist" solutions to be selected since it is based on aggregated fitness scores
 
-## epsilon-Lexicase Selection
+### $\epsilon$-Lexicase Selection
 
-* Recent alternative: Lexicase Selection and it's variation $\epsilon$-lexicase selection
-* Idea: Selection method for uncompromising, continous-valued symbolic regression problems ^[@6920034, p.12]
-* Increases genetic diversity inside the population^[@6920034, p.1]
-* Higher chance for "specialist" solutions to be selected since it is decided on a per case basis
+* Objective: Create Selection method for uncompromising, continous-valued symbolic regression problems ^[@6920034, p.12]
 * Performance increases have been demonstrated in many benchmarking problems ^[@epsilon_lexicase_main, p.744-745]
+* Intuition: Higher chance for "specialist" solutions to be selected since it is decided on a per case basis
 
-## Symbolic Regression
+## Related Concepts
+
+### Symbolic Regression
 
@@
 
-## Generalization
+### Generalization
 
-* The ability of a model to perform well on previously unseen fitness cases
-* Main objective in most supervised machine learning problems
+* Main objective in most supervised machine learning problems: Achieve good performance for unseen data
 * Challenge: Avoid overfitting to training data
-
-## Motivation
-
 * Little attention has been paid to generalization in GP ^[@open_issues_gp, @generalisation_in_gp]
-* High practical relevance of symbolic regression in many fields, e.g. financial forecasting
 
@@
 * Collect and aggregate results for training error, testing error and program length
-
-## Evolutionary Parameters
-
-```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
-fig.pos="H"
-
-knitr::kable(
-  read.csv("../tables/csv/gp_parameters.csv"),
-  format="latex",
-  digits=3,
-  caption="Evolutionary Parameters"
-)
-```
-
-## Primitive Set {.allowframebreaks}
-
-```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
-fig.pos="b"
-
-knitr::kable(
-  read.csv("../tables/csv/primitives.csv"),
-  format="latex",
-  digits=3,
-  caption="Function Set"
-) 
-```
-\framebreak
-
-```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
-fig.pos="t"
-
-knitr::kable(
-  read.csv("../tables/csv/terminals.csv"),
-  format="latex",
-  digits=3,
-  caption="Terminal Set"
-) 
-```
-
-## Example
-
-Model evolved by tournament selection after 100 generations:
-
-![](../plots/example_model.png)
 
@@
 
+# Appendix
+
+## Evolutionary Parameters
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+fig.pos="H"
+
+knitr::kable(
+  read.csv("../tables/csv/gp_parameters.csv"),
+  format="latex",
+  digits=3,
+  caption="Evolutionary Parameters"
+)
+```
+
+## Primitive Set {.allowframebreaks}
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+fig.pos="b"
+
+knitr::kable(
+  read.csv("../tables/csv/primitives.csv"),
+  format="latex",
+  digits=3,
+  caption="Function Set"
+) 
+```
+\framebreak
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+fig.pos="t"
+
+knitr::kable(
+  read.csv("../tables/csv/terminals.csv"),
+  format="latex",
+  digits=3,
+  caption="Terminal Set"
+) 
+```
+
+## Example
+
+Model evolved by tournament selection after 100 generations:
+
+![](../plots/example_model.png)
+
 # References {.allowframebreaks}

### 2022-06-27 docs/rmd/presentation/slides.tex
....
autogit

--- 2022-06-26 docs/rmd/presentation/slides.tex
@@
 \fi
-\usecolortheme{beaver}
 \usefonttheme{structurebold}
@@
 \newif\ifbibliography
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+\usepackage{caption}
+% Make caption package work with longtable
+\makeatletter
+\def\fnum@table{\tablename~\thetable}
+\makeatother
 \usepackage{graphicx}
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\usepackage{float}
 \ifLuaTeX
@@
   A metaheuristic that searches for computer programs that solve a given
-  problem
+  problem\footnote<.->{Koza (1992)}
 \item
-  Inventor: John R. Koza \footnote<.->{Koza (1992)}
-\item
-  Evolutionary algorithm that simulates the process of Darwinian
-  evolution:
-
-  \begin{enumerate}
-  \tightlist
-  \item
-    Population based
-  \item
-    The quality of solutions is evaluated by a fitness function
-  \item
-    Selection: Solutions are selected based on their individual fitness
-  \item
-    Variation: Mutation and recombination of solutions
-  \end{enumerate}
+  Evolutionary Algorithm: Simulates the process of Darwinian evolution
 \item
@@
 \protect\hypertarget{parent-selection}{}
+\begin{block}{Torunament Selection}
+\protect\hypertarget{torunament-selection}{}
 \begin{itemize}
@@
 \item
-  Operator that selects individual solutions from the population for
-  reproduction and mutation
-\item
-  Most commonly used selection operator in Genetic Programming (GP):
-  Tournament selection\footnote<.->{Fang and Li (2010), p.181}
+  Most commonly used selection operator in Genetic Programming
+  (GP)\footnote<.->{Fang and Li (2010), p.181}
 \item
@@
 \end{itemize}
-\end{frame}
+\end{block}
 
-\begin{frame}{epsilon-Lexicase Selection}
+\begin{block}{\(\epsilon\)-Lexicase Selection}
 \protect\hypertarget{epsilon-lexicase-selection}{}
@@
 \item
-  Recent alternative: Lexicase Selection and it's variation
-  \(\epsilon\)-lexicase selection
-\item
-  Idea: Selection method for uncompromising, continous-valued symbolic
-  regression problems \footnote<.->{Helmuth, Spector and Matheson
-    (2015), p.12}
-\item
-  Increases genetic diversity inside the population\footnote<.->{Helmuth,
-    Spector and Matheson (2015), p.1}
-\item
-  Higher chance for ``specialist'' solutions to be selected since it is
-  decided on a per case basis
+  Objective: Create Selection method for uncompromising,
+  continous-valued symbolic regression problems \footnote<.->{Helmuth,
+    Spector and Matheson (2015), p.12}
 \item
@@
   problems \footnote<.->{La Cava, Spector and Danai (2016), p.744-745}
+\item
+  Intuition: Higher chance for ``specialist'' solutions to be selected
+  since it is decided on a per case basis
 \end{itemize}
+\end{block}
 \end{frame}
 
-\begin{frame}{Symbolic Regression}
+\begin{frame}{Related Concepts}
+\protect\hypertarget{related-concepts}{}
+\begin{block}{Symbolic Regression}
 \protect\hypertarget{symbolic-regression}{}
@@
 \end{itemize}
-\end{frame}
+\end{block}
 
-\begin{frame}{Generalization}
+\begin{block}{Generalization}
 \protect\hypertarget{generalization}{}
@@
 \item
-  The ability of a model to perform well on previously unseen fitness
-  cases
-\item
-  Main objective in most supervised machine learning problems
+  Main objective in most supervised machine learning problems: Achieve
+  good performance for unseen data
 \item
   Challenge: Avoid overfitting to training data
-\end{itemize}
-\end{frame}
-
-\begin{frame}{Motivation}
-\protect\hypertarget{motivation}{}
-\begin{itemize}
-\tightlist
 \item
@@
     \emph{et al.} (2010), Kushchu (2002)}
-\item
-  High practical relevance of symbolic regression in many fields,
-  e.g.~financial forecasting
 \end{itemize}
+\end{block}
 \end{frame}
@@
 
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating data set}
-\centering
-\begin{tabular}[t]{l|l}
-\hline
-Variable & Description\\
-\hline
-X1 & Relative Compactness\\
-\hline
-X2 & Surface Area\\
-\hline
-X3 & Wall Area\\
-\hline
-X4 & Roof Area\\
-\hline
-X5 & Overall Height\\
-\hline
-X6 & Orientation\\
-\hline
-X7 & Glazing Area\\
-\hline
-X8 & Glazing Area Distribution\\
-\hline
-y1 & Heating Load\\
-\hline
-y2 & Cooling Load\\
-\hline
-\end{tabular}
-\end{table}
+\begin{longtable}[]{@{}ll@{}}
+\caption{Overview - Energy Heating data set}\tabularnewline
+\toprule()
+Variable & Description \\
+\midrule()
+\endfirsthead
+\toprule()
+Variable & Description \\
+\midrule()
+\endhead
+X1 & Relative Compactness \\
+X2 & Surface Area \\
+X3 & Wall Area \\
+X4 & Roof Area \\
+X5 & Overall Height \\
+X6 & Orientation \\
+X7 & Glazing Area \\
+X8 & Glazing Area Distribution \\
+y1 & Heating Load \\
+y2 & Cooling Load \\
+\bottomrule()
+\end{longtable}
 \end{frame}
@@
 
+\begin{frame}{Research Question}
+\protect\hypertarget{research-question-1}{}
+\begin{itemize}
+\tightlist
+\item
+  Does the usage of \(\epsilon\)-lexicase parent selection influence the
+  generalization behavior of genetic programming in symbolic regression
+  if compared to tournament selection?
+\end{itemize}
+
+\begin{block}{Statistical Testing Strategy}
+\protect\hypertarget{statistical-testing-strategy}{}
+\begin{enumerate}
+\tightlist
+\item
+  Differences in average fitness for both algorithms on both datasets?
+\item
+  Differences in fitness for training and testing data?
+\end{enumerate}
+\end{block}
+\end{frame}
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\begin{frame}{Finding 1}
+\protect\hypertarget{finding-1}{}
+\begin{itemize}
+\tightlist
+\item
+  The differences in average fitness of the final solutions between
+  tournament selection and \(\epsilon\)-lexicase selection are highly
+  statistical significant (\(\alpha=0.01\))
+\item
+  Tournament selection-based GP achieves a higher fitness for both
+  training and testing data
+\item
+  Unexpected results based on the reviewed literature (La Cava, Spector
+  and Danai, 2016), (La Cava \emph{et al.}, 2017)
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Distribution of Fitness}
+\protect\hypertarget{distribution-of-fitness}{}
+\includegraphics{../plots/mean_error_boxplot_all.png}
+\end{frame}
+
+\begin{frame}{Finding 2}
+\protect\hypertarget{finding-2}{}
+\begin{itemize}
+\tightlist
+\item
+  The gap between training and testing error is not statistically
+  significant for both selection algorithms
+\item
+  Both algorithms achieve a slightly better performance for the training
+  data
+\item
+  Good generalization: No evidence of overfitting
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Evolution of Fitness}
+\protect\hypertarget{evolution-of-fitness}{}
+\includegraphics{../plots/mean_error_combined.png}
+\end{frame}
+
+\begin{frame}{Statistical Test}
+\protect\hypertarget{statistical-test}{}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-2}Mean Error - P-Values (MWU)}
+\centering
+\fontsize{4}{6}\selectfont
+\begin{tabular}[t]{l|r|r|r|r}
+\hline
+X & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
+\hline
+tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
+\hline
+tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
+\hline
+elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
+\hline
+elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
+\hline
+\end{tabular}
+\end{table}
+\end{frame}
+
+\begin{frame}{Program Growth}
+\protect\hypertarget{program-growth}{}
+\begin{itemize}
+\tightlist
+\item
+  So far: No proof of differences in generalization
+\item
+  New approach: Program growth as a possible indicator for overfitting?
+\item
+  Theory: Minimum description length principle (MDLP) \footnote<.->{Wang,
+    Wagner and Rondinelli (2019), p.~268}
+\item
+  Downside: Growth/Bloat is no clear indicator of overfitting
+  \footnote<.->{Silva and Vanneschi (2009), p.~8}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Evolution of Size}
+\protect\hypertarget{evolution-of-size}{}
+\begin{figure}
+\centering
+\includegraphics{../plots/size_subplotted.png}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Finding 3}
+\protect\hypertarget{finding-3}{}
+\begin{itemize}
+\tightlist
+\item
+  GP typical growth behaviour for both operators
+\item
+  Solutions grow at a similiar rate in each generation
+\item
+  No statistically significant differences in overall program size based
+  on selection
+\end{itemize}
+\end{frame}
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
+
+\begin{frame}{Conclusions}
+\begin{itemize}
+\tightlist
+\item
+  Experiment did not yield evidence for differences in generalization
+  behavior between tournament and \(\epsilon\)-lexicase selection
+\item
+  The performance of tournament selection is significantly higher than
+  that of \(\epsilon\)-lexicase selection for the selected symbolic
+  regression problem
+\item
+  No evidence for differences in growth behavior between both algorithms
+\end{itemize}
+\end{frame}
+
+\hypertarget{limitations-and-open-questions}{%
+\section{Limitations and open
+Questions}\label{limitations-and-open-questions}}
+
+\begin{frame}{Limitations and open Questions}
+\begin{enumerate}
+\tightlist
+\item
+  Configuration of evolutionary parameters
+\item
+  Results are based on a single symbolic regression
+\item
+  Limited by computational resources
+\end{enumerate}
+\end{frame}
+
+\hypertarget{appendix}{%
+\section{Appendix}\label{appendix}}
+
 \begin{frame}{Evolutionary Parameters}
@@
 
-\caption{\label{tab:unnamed-chunk-2}Evolutionary Parameters}
+\caption{\label{tab:unnamed-chunk-3}Evolutionary Parameters}
 \centering
@@
 
-\begin{frame}{Function Set}
-\protect\hypertarget{function-set}{}
-\begin{tabular}{l|r}
+\begin{frame}[allowframebreaks]{Primitive Set}
+\protect\hypertarget{primitive-set}{}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-4}Function Set}
+\centering
+\begin{tabular}[t]{l|r}
 \hline
@@
 \end{tabular}
-\end{frame}
+\end{table}
+\framebreak
 
-\begin{frame}{Terminal Set}
-\protect\hypertarget{terminal-set}{}
-\begin{tabular}{l|l}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-5}Terminal Set}
+\centering
+\begin{tabular}[t]{l|l}
 \hline
@@
 
-\begin{frame}{Research Question}
-\protect\hypertarget{research-question-1}{}
-\begin{itemize}
-\tightlist
-\item
-  Does the usage of \(\epsilon\)-lexicase parent selection influence the
-  generalization behavior of genetic programming in symbolic regression
-  if compared to tournament selection?
-\end{itemize}
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
 
-\begin{block}{3 Questions}
-\protect\hypertarget{questions}{}
-\begin{enumerate}
-\tightlist
-\item
-  Differences in average fitness for both algorithms?
-\item
-  Differences in fitness for training and testing data?
-\item
-  Differences in program size?
-\end{enumerate}
-\end{block}
-\end{frame}
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\begin{frame}{Distribution of Fitness}
-\protect\hypertarget{distribution-of-fitness}{}
-\includegraphics{../plots/mean_error_boxplot_all.png}
-\end{frame}
-
-\begin{frame}{Finding 1}
-\protect\hypertarget{finding-1}{}
-\begin{itemize}
-\tightlist
-\item
-  The differences in average fitness of the final solutions between
-  tournament selection and \(\epsilon\)-lexicase selection are highly
-  statistical significant (\(\alpha=0.01\))
-\item
-  Tournament selection-based GP achieves a higher fitness for both
-  training and testing data
-\item
-  Unexpected results based on the reviewed literature (La Cava, Spector
-  and Danai, 2016), (La Cava \emph{et al.}, 2017)
-\end{itemize}
-\end{frame}
-
-\begin{frame}{Evolution of Fitness}
-\protect\hypertarget{evolution-of-fitness}{}
-\includegraphics{../plots/mean_error_combined.png}
-\end{frame}
-
-\begin{frame}{Finding 2}
-\protect\hypertarget{finding-2}{}
-\begin{itemize}
-\tightlist
-\item
-  The gap between training and testing error is not statistically
-  significant for both selection algorithms
-\item
-  Both algorithms achieve a slightly better performance for the training
-  data
-\item
-  No proof for overfitting
-\end{itemize}
-\end{frame}
-
-\begin{frame}{Statistical Test}
-\protect\hypertarget{statistical-test}{}
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-5}Mean Error - P-Values (MWU)}
-\centering
-\begin{tabular}[t]{lrrrr}
-\toprule
-  & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
-\midrule
-tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
-tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
-elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
-elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
-\bottomrule
-\end{tabular}
-\end{table}
-\end{frame}
-
-\begin{frame}{Evolution of Size}
-\protect\hypertarget{evolution-of-size}{}
-\begin{figure}
-\centering
-\includegraphics{../plots/size_subplotted.png}
-\end{figure}
-\end{frame}
-
-\begin{frame}{Finding 3}
-\protect\hypertarget{finding-3}{}
-\ldots{}
-\end{frame}
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-\begin{frame}{Conclusions}
-\ldots{}
-\end{frame}
-
-\hypertarget{limitations-and-open-questions}{%
-\section{Limitations and open
-Questions}\label{limitations-and-open-questions}}
-
-\begin{frame}{Limitations and open Questions}
-\ldots{} \newpage
-
+\begin{frame}[allowframebreaks]{References}
 \hypertarget{refs}{}
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-bloat_overfitting_gp}{}}%
+Silva, S. and Vanneschi, L. (2009) {``Operator equalisation, bloat and
+overfitting: A study on human oral bioavailability prediction,''} in
+\emph{Proceedings of the 11th Annual Genetic and Evolutionary
+Computation Conference, GECCO-2009}, pp. 1115--1122.
+doi:\href{https://doi.org/10.1145/1569901.1570051}{10.1145/1569901.1570051}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wang_wagner_rondinelli_2019}{}}%
+Wang, Y., Wagner, N. and Rondinelli, J.M. (2019) {``Symbolic regression
+in materials science,''} \emph{MRS Communications}, 9(3), pp. 793--805.
+doi:\href{https://doi.org/10.1557/mrc.2019.85}{10.1557/mrc.2019.85}.
+
 \end{CSLReferences}

### 2022-06-27 docs/rmd/presentation/slides.Rmd
autogit

@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+```{r vars_dataset echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
 fig.pos="H"
@@
 
-## Research Question
+## Hypothesis Testing
 
-* Does the usage of $\epsilon$-lexicase parent selection influence the generalization behavior of genetic programming in symbolic regression if compared to tournament selection?
+**Does the usage of $\epsilon$-lexicase parent selection influence the generalization behavior of genetic programming in symbolic regression if compared to tournament selection?**
 
-### Statistical Testing Strategy
-
-1. Differences in average fitness for both algorithms on both datasets?
-2. Differences in fitness for training and testing data?
+1. Test for differences in average fitness between both algorithms
+2. Test for differences in average fitness between training and testing data
 
@@
 
+## Example
+
+Model evolved by tournament selection after 100 generations:
+
+![](../plots/example_model.png)
+
 ## Finding 2
@@
 
-```{r, echo=FALSE, warning=FALSE}
+```{r pvals_fitness, echo=FALSE, warning=FALSE}
 df <- read.csv("../tables/csv/mwu_matrix_error.csv")
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+```{r ea_config, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
 fig.pos="H"
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+```{r primitive_table, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
 fig.pos="b"
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+```{r function_table, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
 fig.pos="t"
@@
 
-## Example
-
-Model evolved by tournament selection after 100 generations:
-
-![](../plots/example_model.png)
-
 # References {.allowframebreaks}

### 2022-06-27 docs/rmd/presentation/slides.tex
autogit

@@
 
-\begin{frame}{Research Question}
-\protect\hypertarget{research-question-1}{}
-\begin{itemize}
-\tightlist
-\item
-  Does the usage of \(\epsilon\)-lexicase parent selection influence the
-  generalization behavior of genetic programming in symbolic regression
-  if compared to tournament selection?
-\end{itemize}
+\begin{frame}{Hypothesis Testing}
+\protect\hypertarget{hypothesis-testing}{}
+\textbf{Does the usage of \(\epsilon\)-lexicase parent selection
+influence the generalization behavior of genetic programming in symbolic
+regression if compared to tournament selection?}
 
-\begin{block}{Statistical Testing Strategy}
-\protect\hypertarget{statistical-testing-strategy}{}
 \begin{enumerate}
@@
 \item
-  Differences in average fitness for both algorithms on both datasets?
+  Test for differences in average fitness between both algorithms
 \item
-  Differences in fitness for training and testing data?
+  Test for differences in average fitness between training and testing
+  data
 \end{enumerate}
-\end{block}
 \end{frame}
@@
 
+\begin{frame}{Example}
+\protect\hypertarget{example}{}
+Model evolved by tournament selection after 100 generations:
+
+\includegraphics{../plots/example_model.png}
+\end{frame}
+
 \begin{frame}{Finding 2}
@@
 
-\caption{\label{tab:unnamed-chunk-2}Mean Error - P-Values (MWU)}
+\caption{\label{tab:pvals_fitness}Mean Error - P-Values (MWU)}
 \centering
@@
 
-\caption{\label{tab:unnamed-chunk-3}Evolutionary Parameters}
+\caption{\label{tab:ea_config}Evolutionary Parameters}
 \centering
@@
 
-\caption{\label{tab:unnamed-chunk-4}Function Set}
+\caption{\label{tab:primitive_table}Function Set}
 \centering
@@
 
-\caption{\label{tab:unnamed-chunk-5}Terminal Set}
+\caption{\label{tab:function_table}Terminal Set}
 \centering
@@
 \end{table}
-\end{frame}
-
-\begin{frame}{Example}
-\protect\hypertarget{example}{}
-Model evolved by tournament selection after 100 generations:
-
-\includegraphics{../plots/example_model.png}
 \end{frame}

### 2022-06-27 docs/rmd/presentation/slides.Rmd
autogit

@@
   beamer_presentation:
+    theme: "Warsaw"
     fonttheme: "structurebold"
@@
 csl: ../bib/csl/harvard1.csl
-header-includes:
-   - \usepackage{float}
+includes:
+      in_header: header.tex
 ---
@@
 
+\setbeamertemplate{footline}[frame number]
+
 # Introduction
@@
 
-* Does the usage of $\epsilon$-lexicase parent selection influence the generalization behaviour of genetic programming in symbolic regression if compared to tournament selection?
+* Does the usage of $\epsilon$-lexicase parent selection influence the generalization behavior of genetic programming in symbolic regression if compared to tournament selection?
 
@@
 
-* A metaheuristic that searches for computer programs that solve a given problem^[@koza_main]
+* A meta-heuristic that searches for computer programs that solve a given problem^[@koza_main]
 * Evolutionary Algorithm: Simulates the process of Darwinian evolution
@@
 
-### Torunament Selection
+### Tournament Selection
 
@@
 
-* Objective: Create Selection method for uncompromising, continous-valued symbolic regression problems ^[@6920034, p.12]
+* Objective: Create Selection method for uncompromising, continuous-valued symbolic regression problems ^[@6920034, p.12]
 * Performance increases have been demonstrated in many benchmarking problems ^[@epsilon_lexicase_main, p.744-745]
@@
 * One of the first applications of GP described by @koza_main
-* High relevance: GP can outperform state-of-the-art machine learning algorithms like gradient boosting ^[@Orzechowski_2018]
+* High relevance: GP can outperform state-of-the-art machine learning algorithms, e.g. gradient boosting ^[@Orzechowski_2018]
 
@@
 * Main objective in most supervised machine learning problems: Achieve good performance for unseen data
-* Challenge: Avoid overfitting to training data
+* Challenge: Avoid overfitting to training dataset
 * Little attention has been paid to generalization in GP ^[@open_issues_gp, @generalisation_in_gp]
@@
 
-```{r vars_dataset echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+```{r vars_dataset, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
 fig.pos="H"
@@
 
-* Total dataset ($N=768$) is randomly split into a training and testing dataset (50:50)
+* Total dataset ($N=768$) is randomly split into a training/testing dataset (50/50)
 * Fitness metric: Mean squared Error (MSE)
@@
 
-## Example
-
-Model evolved by tournament selection after 100 generations:
-
-![](../plots/example_model.png)
-
 ## Finding 2
@@
 
-```{r pvals_fitness, echo=FALSE, warning=FALSE}
+```{r pvals_fitness_table, echo=FALSE, warning=FALSE}
 df <- read.csv("../tables/csv/mwu_matrix_error.csv")
@@
 
-* GP typical growth behaviour for both operators
-* Solutions grow at a similiar rate in each generation
+* GP typical growth behavior for both operators
+* Solutions grow at a similar rate in each generation
 * No statistically significant differences in overall program size based on selection
@@
 
-```{r ea_config, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
+```{r ea_config_table, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
 fig.pos="H"

### 2022-06-27 docs/rmd/presentation/slides.tex
autogit

@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\usepackage{float}
 \ifLuaTeX
@@
 \end{frame}
+\begin{frame}
+\setbeamertemplate{footline}{[}frame number{]}
+\end{frame}
+
 \hypertarget{introduction}{%
@@
 \item
-  A metaheuristic that searches for computer programs that solve a given
-  problem\footnote<.->{Koza (1992)}
+  A meta-heuristic that searches for computer programs that solve a
+  given problem\footnote<.->{Koza (1992)}
 \item
@@
 \protect\hypertarget{parent-selection}{}
-\begin{block}{Torunament Selection}
-\protect\hypertarget{torunament-selection}{}
+\begin{block}{Tournament Selection}
+\protect\hypertarget{tournament-selection}{}
 \begin{itemize}
@@
   Objective: Create Selection method for uncompromising,
-  continous-valued symbolic regression problems \footnote<.->{Helmuth,
+  continuous-valued symbolic regression problems \footnote<.->{Helmuth,
     Spector and Matheson (2015), p.12}
@@
 
-\begin{frame}{Example}
-\protect\hypertarget{example}{}
-Model evolved by tournament selection after 100 generations:
-
-\includegraphics{../plots/example_model.png}
-\end{frame}
-
 \begin{frame}{Finding 2}
@@
 
-\caption{\label{tab:pvals_fitness}Mean Error - P-Values (MWU)}
+\caption{\label{tab:pvals_fitness_table}Mean Error - P-Values (MWU)}
 \centering
@@
 \hline
-X & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
+  & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
 \hline
@@
 \item
-  GP typical growth behaviour for both operators
+  GP typical growth behavior for both operators
 \item
-  Solutions grow at a similiar rate in each generation
+  Solutions grow at a similar rate in each generation
 \item
@@
 
-\caption{\label{tab:ea_config}Evolutionary Parameters}
+\caption{\label{tab:ea_config_table}Evolutionary Parameters}
 \centering

### 2022-06-27 docs/rmd/presentation/slides.Rmd
kinda done

@@
 subtitle: "A comparison of epsilon-lexicase Selection and Tournament Selection"
-author: "Roman Hoehn, B.Sc. Wirtschaftspaedagogik"
+author: "Roman Hoehn"
+institute: "JGU Mainz "
 date: "2022-06-29"
@@
   beamer_presentation:
-    theme: "Warsaw"
+    theme: "Madrid"
     fonttheme: "structurebold"
+    fig_width: 12
+    fig_height: 10
     keep_tex: true
@@
 ```
-
-\setbeamertemplate{footline}[frame number]
 

### 2022-06-27 docs/rmd/presentation/slides.tex
kinda done

@@
 \fi
+\usetheme[]{Madrid}
 \usefonttheme{structurebold}
@@
   pdftitle={Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression},
-  pdfauthor={Roman Hoehn, B.Sc. Wirtschaftspaedagogik},
+  pdfauthor={Roman Hoehn},
   hidelinks,
@@
 Selection}
-\author{Roman Hoehn, B.Sc. Wirtschaftspaedagogik}
+\author{Roman Hoehn}
 \date{2022-06-29}
+\institute{JGU Mainz}
 
@@
 \end{frame}
-\begin{frame}
-\setbeamertemplate{footline}{[}frame number{]}
-\end{frame}
-
 \hypertarget{introduction}{%
@@
   High relevance: GP can outperform state-of-the-art machine learning
-  algorithms like gradient boosting \footnote<.->{Orzechowski, Cava and
+  algorithms, e.g.~gradient boosting \footnote<.->{Orzechowski, Cava and
     Moore (2018)}
@@
 \item
-  Challenge: Avoid overfitting to training data
+  Challenge: Avoid overfitting to training dataset
 \item
@@
 \item
-  Total dataset (\(N=768\)) is randomly split into a training and
-  testing dataset (50:50)
+  Total dataset (\(N=768\)) is randomly split into a training/testing
+  dataset (50/50)
 \item

### 2022-06-27 docs/rmd/presentation/slides.tex
done

@@
 
+\begin{frame}{Github Repository}
+\protect\hypertarget{github-repository}{}
+Full paper, source code, results, plots and tables available at:
+
+\textbf{\url{https://github.com/roman91DE/intelligent_information_systems_research_project/}}
+\end{frame}
+
 \hypertarget{references}{%
@@
 Fang, Y. and Li, J. (2010) {``A review of tournament selection in
-genetic programming,''} in Cai, Z. et al. (eds.) \emph{Advances in
+genetic programming,''} in Z. Cai et al. (eds.) \emph{Advances in
 computation and intelligence}. Berlin, Heidelberg: Springer Berlin
@@
 Transactions on Evolutionary Computation}, 19(5), pp. 630--643.
-doi:\href{https://doi.org/10.1109/TEVC.2014.2362729}{10.1109/TEVC.2014.2362729}.
+\url{https://doi.org/10.1109/TEVC.2014.2362729}.
 
@@
 genetic programming,''} \emph{Artificial Intelligence Review - AIR}, 18,
-pp. 3--14.
-doi:\href{https://doi.org/10.1023/A:1016379201230}{10.1023/A:1016379201230}.
+pp. 3--14. \url{https://doi.org/10.1023/A:1016379201230}.
 
@@
 analysis of lexicase selection and epsilon-lexicase selection.''} arXiv.
-doi:\href{https://doi.org/10.48550/ARXIV.1709.05394}{10.48550/ARXIV.1709.05394}.
+\url{https://doi.org/10.48550/ARXIV.1709.05394}.
 
@@
 Association for Computing Machinery (GECCO '16), pp. 741--748.
-doi:\href{https://doi.org/10.1145/2908812.2908898}{10.1145/2908812.2908898}.
+\url{https://doi.org/10.1145/2908812.2908898}.
 
@@
 programming,''} \emph{Genetic Programming and Evolvable Machines}, 11,
-pp. 339--363.
-doi:\href{https://doi.org/10.1007/s10710-010-9113-2}{10.1007/s10710-010-9113-2}.
+pp. 339--363. \url{https://doi.org/10.1007/s10710-010-9113-2}.
 
@@
 now?''} in \emph{Proceedings of the genetic and evolutionary computation
-conference}. {ACM}.
-doi:\href{https://doi.org/10.1145/3205455.3205539}{10.1145/3205455.3205539}.
+conference}. {ACM}. \url{https://doi.org/10.1145/3205455.3205539}.
 
@@
 Computation Conference, GECCO-2009}, pp. 1115--1122.
-doi:\href{https://doi.org/10.1145/1569901.1570051}{10.1145/1569901.1570051}.
+\url{https://doi.org/10.1145/1569901.1570051}.
 
@@
 in materials science,''} \emph{MRS Communications}, 9(3), pp. 793--805.
-doi:\href{https://doi.org/10.1557/mrc.2019.85}{10.1557/mrc.2019.85}.
+\url{https://doi.org/10.1557/mrc.2019.85}.
 
